### 2012-07-10 letter.tex
first commit

\documentclass[11pt]{newlfm}
\usepackage{setspace}
\onehalfspacing
\newlfmP{leftmarginsize=.75in,leftmarginskipright=15pt,topmarginskip=2em,%
stdletter,
noheadline,nofootline,dateright,addrfromright,%
orderfromdateto,sigleft}

\nameto{Prof.~William Jorgensen}
\addrto{%\parbox{2.5in}{
Department of Chemistry\\
Yale University\\
PO Box 208107\\
New Haven, CT 06520-8107}

\namefrom{Rajarshi Guha}
\addrfrom{%\parbox{2.5in}{
NIH Center for Advancing Translational Science \\
9800 Medical Center Drive \\
Rockville, MD 20852
}%}
\phonefrom{814 404 5449}

\begin{document}
    \closeline{Yours Sincerely,}
   \greetto{Dear Prof.~Jorgensen,}
    \begin{newlfm}
      I would like to submit an article titled ``\emph{Exploring
        Uncharted Territories - Predicting Activty Cliffs in
        Structure-Activity Landscapes}'' for publication in the
      Journal of Chemical Information and Modeling.

      The article builds on previous work that defined a numerical
      quantification of structure activity landscapes, termed the
      Structure Activity Landscape Index. This (and other methods such
      as SARI) have been used to characterize SAR data using the
      landscape paradigm, allowing users to identify activity cliffs
      in their datasets. However, these approaches have primarily been
      retrospective. In this paper I highlight how SALI can be used in
      a prospective manner, such that one can predict SALI values for
      pairs of molecules, one of which is a new, untested molecule. In
      other words, the method discussed in this paper allows one to
      prospectively prioritize new molecules as likely to exhibit
      activity cliffs with one or more members of the original
      dataset. The predictive approach is based on random forest
      models and considers the data set in a pairwise fashion - thus
      making it amenable for small SAR datasets. The models exhibit
      good predictive performance, suggesting the utility of this
      approach.
%  \vskip 0.5em
    \end{newlfm}
\end{document}
%%% Local Variables: 
%%% mode: latex
%%% TeX-master: t
%%% End:

### 2012-07-10 paper.tex
first commit

\documentclass[letterpaper, 12pt]{article}
%%\documentclass[letterpaper, 11pt, twocolumn]{article}
\usepackage{url}
\usepackage{ctable}
\usepackage{overcite}
\usepackage{amsmath,amssymb}      % for \AA etc
\usepackage[onehalfspacing]{setspace}
\usepackage{graphicx}
\usepackage{fullpage, times}
\usepackage{algorithmic}
\usepackage{algorithm}
\usepackage{longtable}

\DeclareGraphicsExtensions{.pdf,.png}
\onehalfspacing

\bibliographystyle{jcics}

\usepackage{anysize}
%\marginsize{1in}{2in}{1in}{1in}
%\marginparwidth 1.75in
%\newcommand{\mnote}[1]{\marginpar{\raggedright{\footnotesize\textbf{#1}}}}

\def\imagetop#1{\vtop{\null\hbox{#1}}}

%%%
%%% Don't change the size of the figures as they are defined for
%%% a single column environment, which is what we will be submitting
%%% as
%%%

\begin{document}
\title{Exploring Uncharted Territories - Predicting Activty Cliffs in
  Structure-Activity Landscapes}
\author{Rajarshi Guha${}^{\ddagger}$\\
NIH Center for Advancing Translational Sciences \\ 9800 Medical Center Drive  Rockville, MD 20850}
\date{}

\maketitle
\begin{abstract}
  The notion of activity cliffs is an intuitive approach to characterizing structural features that
  play a key role in modulating biological activity of a molecule. A variety of methods have been
  described to quantitatively characterize activity cliffs, such as SALI and SARI. However, these
  methods are primarily retrospective in nature; highlighting cliffs that are already present in the
  dataset. The current study focuses on employing a pairwise characterization of a dataset to train
  a model to predict whether a new molecule will exhibit an activity cliff with one or more members
  of the dataset. The approach is based on predicting a value for pairs of objects rather than the
  individual objects themselves (and thus allows for robust models even for small SAR datasets). We
  extracted structure-activity data for several ChEMBL assays and developed random forest models to
  predict SALI values, from pairwise combinations of molecular descriptors. The models exhibited
  reasonable RMSE's though, surprisingly, performance on the more significant cliffs tended to be
  better than on the lesser ones. Our results indicate that the predictive models are able to
  prioritize molecules in terms of their ability to activity cliffs and can be used as a way to
  extend an observed activity landscape to identify molecules that could lead to significant
  improvements in activity. Conversely, the method can be used to deprioritize molecules that are
  predicted to show an activity cliff with a potent molecule already in the dataset.
\end{abstract}

\section{Introduction}

The landscape paradigm for SAR data was first proposed 20 years ago \cite{Johnson:1990ys} and has
recently seen a resurgence with a number of studies describing new ways to quantify and visualize
activity landscapes. When SAR data is viewed as a landscape, with the X-Y plane representing
structural characteristics (which will usually be a 2-dimensional representation of a
multi-dimensional descriptor space) and the Z-axis representing the observed activities, one can
identify two broad types of regions on the landscape - smooth rolling regions, corresponding to set
of molecules exhibiting continuous SAR (i.e., similar structures and similar activities) and rough,
gorge-like regions (i.e., very similar structures, but large differences in activity) corresponding
to molecules that exhibit SAR discontinuity. The latter have also been term activity cliffs
\cite{Maggiora:2006aa}. From a medicinal chemistry point of view, the latter regions of a landscape
can be the most interesting as they can provide insight into structural features that are key to
improving (or conversely reducing) potency. There is a rich history of methods that have correlated
structural differences with corresponding differences in activity -- matched molecular pairs
\cite{Leach:2006aa}, SAS maps \cite{Maggiora:2011kx} and more recently SALI \cite{Guha:2008aa} and
SARI \cite{Peltason:2007aa}. Both SALI and SARI focus on numerically characterizing a structure
activity landscape. The former is defined for a pair of molecules as
\begin{equation}
  \label{eq:1}
  S_{i,j} = \frac{|A_i - A_j|}{1 - sim(i,j)}
\end{equation}
where $A_i$ and $A_j$ represent the observed activities of molecules $i$ and $j$, and $sim(i,j)$
represents the structural similarity between the two molecules (usually based on some form of
fingerprint similarity). Using the SALI metric, one can take a collection of $n$ molecules and
represent them as an $n \times n$ matrix of SALI values - larger values representing more
significant activity cliffs. The SARI approach is based on a score defined as
\begin{equation}
  \label{eq:2}
  \mathrm{SARI} = \frac{1}{2} \left(\mathrm{score}_{cont} + (1 -
    \mathrm{score}_{disc}) \right)
\end{equation}
where the individual score terms are derived on the basis of potency and pairwise similarities. The
reader is referred to Ref. \citen{Peltason:2007aa} for a detailed discussion of this approach.

In either case, one can take the numerical values and visualize them in a variety of ways ranging
from heatmaps of SALI matrices to network representations \cite{Wawer:2008aa,Guha:2008aa}. These
visualizations then allow the user to explore the landscape, quickly identifying a range of activity
cliffs, which an then be examined in detail. Apart from identifying individual activity cliffs, a
variety of other SAR constructs, such as ``activity ridges'' \cite{Vogt:2011bs} and multitarget
landscapes\cite{Dimova:2011fk} can also be identified and characterized.

A feature common to all, recently published work on activity landscapes is that they are primarily
retrospective. That is, the methodologies developed are used to analyze SAR datasets for which
activities have already been experimentally obtained. For example, using the SALI, one can
characterize SAR patterns in a dataset but does not provide insight into whether a new molecule may
be part of an activity cliff, with respect to the original dataset. A number of applications have
attempted to extract SAR rules based on the landscape (e.g., similarity potency trees
\cite{Wawer:2010fv} and multi-target landscape analyses \cite{Hu:2010kl}) or directly identify
structural modifications that lead to activity cliffs \cite{Wassermann:2010oq, Peltason:2009vn}.

\subsection{Motivation}
\label{sec:motivation}
The preceding discussion highlights the utility of retrospective analyses of SAR data using the
activity landscape paradigm. But equally, if not more, interesting is determining whether a new,
untested molecule might be an activity cliff in the context of the original dataset. For the
remainder of this work we focus on the use of SALI to quantify activity landscapes. More
specifically, the ability to \emph{predict} SALI values would be useful as it would allow us to both
fill in empty regions of an activity landscape as well as extend a structure-activity
landscape. Note that this approach to expanding the extent of a SAR dataset does not lend itself to
scaffold hopping since the premise of scaffold hopping is that one generates new cores, which differ
substantially from the starting structure.

In traditional QSAR modeling approaches, one simply predicts the activity of a new molecule and
would then evaluate the SALI (or SARI or some other measure) to determine whether the molecule leads
to an activity cliff. However, the fact that an activity cliff represents a SAR
discontinuity\cite{Maggiora:2006aa} implies that most statistical and machine learning methods will
be unlikely to predict very different activities for two structurally similar molecules. In other
words, a new molecule, similar to a subset of the training set, will tend to have a predicted value
that is similar to those molecules, rather than a drastically different value.

An alternative approach, that is the focus of this paper, is to directly predict SALI values for
pairs of molecules. Thus rather than predict individual activities, we predict SALI values for pairs
of molecules. This approach is somewhat similar to the SPREAD method \cite{Scheiber:2009fk} which
identified substructures that were predictive of activity differences. However, our solution
considers both activity differences and structural similarities. As a result, instead of ranking
compounds in terms of their predicted activity, we instead rank a compound in terms of its predicted
SALI; i.e., its predicted ability to exhibit an activity cliff when paired with other molecules in
the dataset. This approach could be useful when deciding how far to extend an analog series as well
as prioritizing scaffolds for further study.

This does not completely alleviate the problem of discontinuities, since SALI values are infinite
when the $T_c$ is 1.0.  However, predicting SALI values allows us to work with smaller datasets
(since the objects to predict are pairs of molecules), that would ordinarily lead to unreliable
models if we were working with activities. Of course, this means that the approach is not practical
for very large datasets.

The paper is organized as follows. Section \ref{sec:datasets} describes the datasets used in this
study. Section \ref{sec:methodology} presents the methodology we employ to predict activity cliffs
and Section \ref{sec:applications} discusses the results of the predictive models. Finally, Sections
\ref{sec:discussion} and \ref{sec:conclusions} discusses some of the issues underlying this approach
and possible extensions of this work.

\section{Datasets}            
\label{sec:datasets}
For this study we considered a number of datasets, which are summarized in Table \ref{tab:data}. The
Cavalli dataset was employed in our previous studies consisted of 30 molecules studied by Cavalli et
al\cite{Cavalli:2002aa} as possible hErg inhibitors using a pharmacophore modeling approach. The
remaining datasets were obtained from ChEMBL. All three assays involved direct binding to a human
target and we considered the subset of molecules in each assay that had non-censored experimental
values. The Costanzo dataset\cite{Costanzo:2005uq} consisted of 60 $\alpha$-ketoheterocyclic
inhibitors of $\alpha$-thrombin. The reported $\mathrm{IC_{50}}$ values ranged from 3 nM to 82
$\mu$M. The Kalla dataset\cite{Kalla:2006kx} consisted of 38 8-(C-4-pyrazolyl) xanthines, identified
as antagonists of the A2B adenosine receptor. The reported $K_i$ values ranged from 0.9 nM to 42
$\mu$M. Finally, the Dai dataset\cite{Dai:2007kl} consisted of 44 3-aminoindazole derivatives
studied for their ability to inhibit the VEGF and PDGF receptor families, with $\mathrm{IC_{50}}$
values ranging from 3 nM to 12~$\mu$M.

% To obtain the remaining datasets we mined the
% ChEMBL database to identify assays with the following constraints:
% cntaining between 30 and 500 molecules, direct binding to a human
% protein target, a data type of $K_i$ or $\mathrm{IC_{50}}$ and a
% reported confidence score of 8 or more. This identified 780
% assays. For each assay we considered those molecules with non-censored
% measured values and identified those assays for which at least one
% pair of molecules exhibited 1000-fold (or better) difference in
% activity. This led to a final set of 347 assays. From this set we
% considered three assays for three different targets

\section{Methodology}
\label{sec:methodology}
As noted above, the problem of identifying activity cliffs involving new molecules can be reduced to
predicting the SALI value for the new molecule and a pre-existing molecule. Note that this does not
result in an activity prediction for the new molecule; rather, it allows us to rank a set of new
molecules in terms of their predicted ability to exhibit a significant activity difference from one
or more of the molecules in the training set.

Given a training set of $N$ molecules, we generate a new training set
of $\frac{N(N-1)}{2}$ objects, where each object is a pair of
molecules from the original training set. The dependent variable for
each pair $i,j$, is the SALI value, $S_{i,j}$. SALI values were
evaluated using the 1051-bit BCI keyed fingerprints or the CDK
1024-bit path fingerprints. Based on the
definition of SALI, it is possible that a pair of molecules have a
$T_c = 1.0$ resulting in infinite values. For such cases, we replaced
the infinite value with the highest non-infinite SALI value for that
dataset.

The next step is to generate a set of independent variables. Since the
new dataset consists of pairs of molecules, we consider the
descriptors for the resultant objects as a function of the descriptor
values of the individual molecules. For an object, representing the
$i$'th and $j$'th molecules, its descriptor vector can be taken as the
arithmetic mean of the descriptor vectors of the individual
molecules. We denote this aggregation function as
$f_{\textrm{mean}}$. Alternative functions that were investigated included
the absolute difference of the individual descriptors, denoted by
$f_{\textrm{diff}}$ and the geometric mean of the individual
descriptor vectors, denoted by $f_{\textrm{geom}}$.

Given the independent and dependent variables for the pairwise
dataset, we can now proceed to model development. For this study we
focused on the use of random forest models\cite{Breiman:1984aa}. This
was motivated by the fact that such models can be proven to not
overfit and the fact that the algorithm implicitly performs feature
selection. As a result, this allows us to forgo an explicit feature
selection step and work directly with the descriptor pool (after
removal of correlated and constant descriptors). Furthermore, there is
no reason, \emph{a priori} to assume that the underlying SARs are
linear. A random forest model, being an algorithmic
approach\cite{Breiman:2001nx} (as opposed to a distributional one such
as linear regression) makes no such assumptions. We employed the  
implementation of random forest from the \texttt{randomForest}
package, in R 2.11.0\cite{r}, using the defult settings.

In this work, we employed the CDK to evaluate 109 2D and
constitutional descriptors for each individual molecule. For each
dataset, we performed objective feature selection by removing
descriptors with constant or near-constant values followed by removal
of descriptors that are highly correlated with others (using an $R^2$
cutoff of 0.8). The size of the final descriptor pools is summarized
in Table \ref{tab:data}.

One feature of this approach is that one can build relatively robust
models even for datasets of small size - say, 20 molecules. Of course,
a large training set is just one aspect of a reliable model and other
considerations such as diversity, descriptor selection and so on still
play an important role.

\subsection{Is structural information duplicated?}
\label{sec:are-we-duplicating}

One question that arises from the preceding discussion is the
possibility that we are duplicating information between the dependent
and independent variables. More specifically, both the dependent
variable (i.e., the SALI values) and the aggregated descriptors
characterize the molecules' structure. One might therefore ask whether
models built using such data perform over-optimistically. Given the
nature of the descriptors used in the indepdent and dependent
variables (the former being based on multiple atom and bond features
and the latter derived from purel topological paths), we believe that such a
phenomenon is unlikely.  Fig.~\ref{fig:desccor1} displays a histogram
of the pairwise Pearson correlations between the SALI values and each
of the descriptor values for the Cavalli\cite{Cavalli:2002aa} dataset,
using the different aggregation functions described above. It is
evident, that the highest $R^2$, between any of the descriptors and
the SALI values is less than 0.15. Similar behavior was observed for
all the datasets used in this study. These observations
suggest that the problem of including correlated structural
information simultaneously in the dependent and independent variables
is minimal.

\section{Results}
\label{sec:applications}
We first consider the application of the activity cliff prediction
methodology to the Cavalli dataset. We developed three random forest
models, using the three aggregation descriptor aggregation functions
described in Section
\ref{sec:methodology}. Fig.~\ref{fig:rfmodel-cavalli} presents the
predicted versus observed SALI values from the three models and Table
\ref{tab:rfmodel-results} summarizes the performance metrics for the
three models. Overall the different aggregation functions do not
differ dramatically in terms of the final model performance, though
$f_{\textrm{mean}}$ appears to lead to the best performance.

While the models are certainly not the most accurate models, the
RMSE's and $R^2$ values are reasonable. Note that the predictions on
the low end of the SALI values are not as important as those at the
high end - simply because low SALI values correspond to small activity
cliffs, which are likely not very interesting. Given that observation,
it is encouraging to observe that all three models perform relatively
well at the higher end of the SALI spectrum. For this example, the
three most significant activity cliffs are in fact not very
significant activity cliffs in an absolute sense - the $T_c$ for the
the three pairs are just 0.2, 0.30 and 0.29, though the activity
differences were 4.93, 5.0 and 5.11 log units respectively.

We observed that the use of the geometric mean did not lead to models
that were any different from those generated using the arithmetic mean
as the aggregation function. This was also true of the other datasets
and so the following discussion ommits the results obtained when using
$f_{\textrm{geom}}$. Fig.~\ref{fig:rfmodel-chembl} summarizes the
models built on the ChEMBL datasets using $f_{\textrm{diff}}$ and
$f_{\textrm{mean}}$ as the aggregation functions. Table
\ref{tab:rfmodel-results} reports the model statistics.  While the
differences in model performance do not vary significantly with the
aggregation function used, we see that for the Costanzo and Kalla
datasets, $f_{\textrm{diff}}$ leads to a slightly better model than
when using $f_{\textrm{mean}}$, though for the Dai dataset, the
difference is negligable.

Given the relatively small differences between models based on
different aggregation functions, we arbitrarily chose
$f_{\textrm{mean}}$ as the aggregagtion functin for subsequent
analyses. Figs.~\ref{fig:costanzo-pred}, \ref{fig:kalla-pred} and
\ref{fig:dai-pred} summarize the performance of the models built using
the reported activity data and the pairwise SALI values for the
Costanzo, Kalla and Froimowitz datasets, respectively. As with the
Cavalli dataset, the SALI models exhibit more variance for small SALI
values (i.e., less significant cliffs) but appear to perform better at
the the higher end of the SALI values. However in all three cases, the
most significant cliffs exhibit significant variance. In comparison,
the models developed on the reported activity values are relatively
poor. This can be partially ascribed to the small sizes of the
dataset.

\subsection{Extending a landscape}
\label{sec:extending-landscape}

Clearly, modeling the pairwise SALI values leads to reasonable models
the datasets discussed here. However, the primary goal of this
approach is to extend the the structure-activity landscape for a set
of molecules. In other words, can one use the model to identify
compounds that will exhibit an activity cliff with one or more
molecules in the dataset. Given a new molecule, one can evaluate the
descriptors for this molecule and then evaluate the pairwise
descriptor values with each member of the training set. Thus, given a
new molecule, we must obtain predictions for $n$ ``pairwise objects'',
where $n$ is the number of molecules in the training set. Given these
$n$ predictions one can proceed to identify training set molecules
with which the new molecule might exhibit a cliff.

We first performed this analysis with the Kalla dataset. Since
predictions of new molecules involve $n$ predictions per molecule, we
removed three molecules from the dataset and rebuilt the random
forest model. The hold out molecules, were specifically selected since
they displayed significant activity cliffs with various members of the
training set. We then evaluted the pairwise descriptor values for
these three with the training set molecules and obtained predictions
of the log(SALI) values. Fig.~\ref{fig:tset-pset-kalla}A
summarizes the performance of the model on the training set as well as
for the prediction set. From a numerical point of view, the
performance of the model on he prediction set degrades somewhat (RMSE
of 0.51 versus 0.41 for the training set). As noted above, the
variance of the prediction is higher at the lower scale of log(SALI)
values. However, at the higher end of the scale there a number of
pairs that are well predicted. Yet, it is clear that a number of
relatively significant activity cliffs have been underestimated.

Given that large SALI values can arise due to high structural
similarity, even when the activity difference is small, simply
examining log(SALI) values is not comletely informative. Specifically,
we are interested in the predictions for the actual activity cliffs,
where both the difference in activity and the structural similarity is
very high. Fig.~\ref{fig:tset-pset-kalla}B visualizes this
information. Each point coresponds to a pair of molecules, which are
shaded by the absolute prediction residual for that pair. Thus, the
``true'' cliffs are represented by points lying towards the bottom
right corner of the plot. In this case we see that there are a number
of pairs of molecules with $1 - T_C \leq 0.1$ and 100-fold or better
difference in activity. While many of these pairs are associated with
a high prediction residual, there are a number with low to medium
residuals.

Fig.~\ref{fig:kalla-comparison} displays the structures of the three
hold out molecules (top row) and a selection of structures, which
which these hold out molecules were predicted to exhibit an activity
cliff. For each molecule, the $K_i$ value is listed along with the
absoute residual value (log(SALI) units). For molecules
\textbf{349288} and \textbf{349699}, the predicted cliffs are
relatively accurate, as evidenced by the low residuals. For
\textbf{349138} on the other hand, most of the residuals were were
relatively high. While we have highlighted a number of cliffs,
Fig.~\ref{fig:tset-pset-kalla}C makes it clear that the predictions
for this molecule were, on average, worse than for the other two
molecules.

Fig.~\ref{fig:tset-pset-dai} summarizes the results for the same set
of analyzes described above, applied to the Dai dataset. As before we
removed three molecules (top row of Fig.~\ref{fig:dai-comparison}) and
rebuilt the random forest on the paired cases with the remaining 41
molecules. We then predicted the pairwise SALI values for the three
hold-out molecules with the remaining 41 molecules (i.e., the
prediction set). A plot of the predicted versus observed log(SALI)
values for the training and prediction sets are displayed in
Fig.~\ref{fig:tset-pset-dai}A. As before the model performs poorly on
the lesser cliffs, but improves for the more significant cliffs (RMSE
for the training set and prediction sets were 0.37 and 0.43 log units
respectively). However, the 2 most significant cliffs are
under-estimated. However, in comparison to the Kalla dataset, the
distribution of absolute residuals aggregated by the hold-out molecule
(Fig.~\ref{fig:tset-pset-dai}C) are reatively similar, with a median
absolute residual of less than 0.3 log units.  For this dataset, the
number of ``real'' cliffs is relatively low as shown in
Fig.~\ref{fig:tset-pset-dai}B. Interestingly, the compound pairs that
display moderate to significant cliffs are relatively well predicted -
in fact, the maximum residuals are observed for the least significant
cliffs.

We then considered the three hold-out molecules and some of the members
of the training set with which they are predicted to show activity
cliffs. For this dataset, there are relatively few ``true'' activity
cliffs. For example, while molecules
\textbf{371259} and \textbf{371307} have a 1000-fold difference in
activity, their $T_c = 0.68$, and would not represent a truly
significant cliff. However, the pair is well poredicted with a
residual of 0.04 log units. if we consider the subset of predictions
where the $T_c \geq 0.8$ we note that the median absolute residual is
0.24 log units. However, the activity differences are not always
significant ranging from 1.1-fold to 13-fold. Two such examples are shown
in Fig.~\ref{fig:dai-comparison}, where the only difference is in the
position of the methyl substituent. Similar behavior is observed with
the other hold-out molecules.

\section{Discussion}
\label{sec:discussion}

The results described above suggest that predicting pairwise SALI
values is a useful way to identify whether a new molecule will
form an activity cliff with one or more members of the training
set. As with all QSAR models, one is limited by the nature of the
underlying data, both in terms of accuracy as well as
applicability. Thus while we have selected a few ChEMBL assays for the
purposes of highlighting this approach, we have found that for some
assays, the model built on pairwise data performs worse than the model
built on the original data. Given that most of the datasets are small
in size (less than 40 molecules), the poor performance of the initial
model is not always surprising. However, in cases where the pairwise
model does not exhibit improved performance, we believe that
predictive modeling of activity cliffs in the original dataset will
not be fruitful.

Given that many of the models show relatively poor statistics on the
pairwise data, it is important to note that much of this is due to the
large variance in the predictions for the less significant cliffs. One
possible reason for the increase in variance for these portions of the
dataset is the fact that the distribution of the log(SALI) values
tends to be skewed to the right (Fig.~\ref{fig:logsalidist}). In
general we see that the smaller SALI values constitute a relatively
small portion of the dataset.

Another factor affecting the models built on pairwise data is the
descriptors that are employed. While the choice of initial descriptors
is always open to debate, we partially avoid a biased selecion by
employing a random forest model to perform implicit feature
selection. However, given an initial set of descriptors we observe
that the models are robust to the different aggregation functions to
derive a pairwise descriptor representation from the descriptors for
the individual compounds. This is a useful observation given that
there appears to be no rule to select one aggregation function over
another \emph{a priori}. While random forests do allow us to avoid
feature selection, it is clear that other models, such as neural
networks or support vector machines, could lead to more predictive
solutions. At this stage, our aim is to highight the modeling
procedure and we believe that the random forest allows us to highight
the utility of this approach without sacrificing too much by way of
predictive performance.

It is important to note that while the modeling approach here focuses
on predicting SALI values rather than actual activities, one can
obtain an idea of the latter based on the (known) activity of the
training set molecules. Thus for a new molecule, predicted to show a
cliff with a member of the training set, we can determine whether this
molecule might be the potent member of the cliff, by noting whether
the corresponding training set molecule is less or more potent. Given
that one is usually interested in activity cliffs because of the
possibility of increasing potency, this observation implies that for a
collection of new molecules, one can prioritize them as being
predicted to be more or less potent. From this point of view, the
approach described here is more suited for prioritization, rather than
prediction of absolute values.

\section{Conclusions}
\label{sec:conclusions}

We have presented an approach to extending a structure-activity
landscape in an indirect fashion, by predicting the propensity of a
new molecule to exhibit an activity cliff with one or more molecules
in a pre-exisiting SAR dataset. The method is based on building
a model on pairwise SALI values (dependent variable) and pairwise
aggregated descriptor values (independent variables). For a new
molecule, we obtain $n$ pairwise SALI predictions (since we must
estimate the SALI value for each training set molecule and the test
molecule). The predicted SALI values can then be used to judge whether
the new molecule will exhibit a cliff and whether such a cliff is in
the desired direction (i.e., improving potency versus worsening
potency). To test this strategy we have developed random forest modls
to predict SALI values for several ChEMBL datasets. While the model
performance statistics are not stellar, we observe that this is
primarily due to high prediction variance at the lower range of SALI
values. In contrast, the more significant cliffs are relatively well
predicted, though, unsurprisingly, the most significant cliffs are not
always well predicted.

In summary, this approach extends the activity cliff concept, from
that of a retrospective analysis tool to a prospective tool that could
be used to guide synthetic campaigns in their goals of improved
potency.

\clearpage
\newpage

\bibliography{paper}

\newpage

\ctable[cap={}, caption={Datasets considered in this study.},
label={tab:data}]
{lp{0.4\textwidth}rrrr}
{}
{\FL
Name & Description & Endpoint & ChEMBL AID & Size & Reference
\ML
Cavalli & \raggedright{Putative hERG inhibitors (via pharmacophore modeling)} & Score &  & 30 & Ref.~\citen{Cavalli:2002aa} \NN
Costanzo & \raggedright{Thrombin inhibitors} & $\mathrm{IC_{50}}$ & 305283 & 60 &
Ref.~\citen{Costanzo:2005uq} \NN
Kalla & \raggedright{A2B adenosine receptor antagonists} & $K_i$ &
364155 & 38 & Ref.~\citen{Kalla:2006kx} \NN
Dai & \raggedright{RTK inhibitors} & $K_i$ & 429056 & 44 & Ref.~\citen{Dai:2007kl}
% Mewshaw & \raggedright{ER$\beta$ inhibitors} & $\mathrm{IC_{50}}$ &
% 306229 & 70 & \cite{Mewshaw:2005vf} \NN
\LL}

\ctable[cap={}, caption={Performance metrics for the random forest
  models built for the datasets used in this study. RMSE and $R^2$ are
based on out-of-bag data\cite{Breiman:1984aa}},
label={tab:rfmodel-results}]
{lrcrr}
{}
{\FL
Dataset & Y-range & Aggregation Function & RMSE & $R^2$ 
\ML
Cavalli & 7.21 & $f_{\textrm{diff}}$ & 1.11& 0.57\NN
        & & $f_{\textrm{mean}}$ & 0.98& 0.67\NN
        & & $f_{\textrm{geom}}$ & 1.01& 0.65\NN \NN
Costanzo & 288.03& $f_{\textrm{diff}}$ & 12.88& 0.24\NN
       & & $f_{\textrm{mean}}$ & 13.56& 0.17\NN \NN
Kalla & 88.25 & $f_{\textrm{diff}}$ & 7.51& 0.62\NN
      &  & $f_{\textrm{mean}}$ & 8.26& 0.53\NN \NN
Dai & 32.71 & $f_{\textrm{diff}}$ & 2.60& 0.44\NN
     &   & $f_{\textrm{mean}}$ & 2.67& 0.41
\LL
}

\clearpage
\newpage

\ctable[cap={}, caption={Distribution of Pearson correlations between
  SALI values and the descriptors for the Cavalli
  dataset. Three descriptor aggregation functions are considered.},
label={fig:desccor1},figure,botcap]
{c}
{}
{
  \includegraphics[width=\linewidth]{sali-desc-corr}
}

\ctable[cap={}, caption={Plots of predicted versus observed SALI values
  obtained using random forest models, on the Cavalli dataset. Each
  panel corresponds to the use of a different descriptor aggregation
  function.}, label={fig:rfmodel-cavalli}, botcap, figure]
{c}
{}
{
\includegraphics[width=\linewidth]{rfmodels-cavalli}
}

\ctable[cap={}, caption={Predicted versus observed SALI values,
  obtained from random forest models for the three ChEMBL datasets. The
 three plots correspond to the two different aggregation functions
 ($f_{\textrm{diff}}$ and $f_{\textrm{mean}}$ respectively).},
label={fig:rfmodel-chembl},botcap,figure]
{p{0.5in}c}
{}
{
\imagetop{\textbf{Costanzo}} & \imagetop{\includegraphics[width=0.9\linewidth]{rfmodels-costanzo}}\NN
\imagetop{\textbf{Kalla}}  & \imagetop{\includegraphics[width=0.9\linewidth]{rfmodels-kalla}} \NN
\imagetop{\textbf{Dai}} &
\imagetop{\includegraphics[width=0.9\linewidth]{rfmodels-dai}}   
}

\ctable[cap={}, caption={Results of random forest models developed
  using the Costanzo dataset. The left plot shows the predicted versus
  observed plot for the model using the original $\mathrm{IC_{50}}$ values, and
  the right hand plot shows the predicted versus observed log(SALI)
  values.}, label={fig:costanzo-pred}, figure, botcap]
{c}
{}
{\includegraphics[width=\linewidth]{pred305283}}

\ctable[cap={}, caption={Results of random forest models developed
  using the Kalla dataset. The left plot shows the predicted versus
  observed plot for the model using the original $K_i$ values, and
  the right hand plot shows the predicted versus observed log(SALI)
  values.}, label={fig:kalla-pred}, figure, botcap]
{c}
{}
{\includegraphics[width=\linewidth]{pred364155}}

\ctable[cap={}, caption={Results of random forest models developed
  using the Dai dataset. The left plot shows the predicted versus
  observed plot for the model using the original $K_i$ values, and
  the right hand plot shows the predicted versus observed log(SALI)
  values.}, label={fig:dai-pred}, figure, botcap]
{c}
{}
{\includegraphics[width=\linewidth]{pred429056}}

\ctable[cap={}, caption={Detailed analysis of SALI predictions for the
  Kalla dataset. \textbf{A} - a plot of predicted versus observed
  log(SALI) values for the training set and the hold out
  set. \textbf{B} - a summary of the training set, where we plot the
  structural difference versus the
  logarithm of the ratio of the activities for each pair of molecules
  in the prediction set. Points are shaded by their absolute
residual. \textbf{C} - a box plot summarizing the distribution of  
residuals associated with predictions from each of the hold out molecules.},
label={fig:tset-pset-kalla}, figure, botcap]
{c}
{}
{\includegraphics[width=\linewidth]{tset-pset-kalla}}

\ctable[cap={}, caption={The hold out molecules for the Kalla dataset
  and training set members with which the hold outs exhibit predicted
  activity cliffs. Bold numbers are ChEMBL MOLREGNO values and numbers
  in parentheses are the absolute prediction
  residual in log(SALI) units.}, figure, botcap,
label={fig:kalla-comparison}]    
{c}
{}
{\includegraphics[width=\linewidth]{kalla-comparison}}

\ctable[cap={}, caption={Detailed analysis of SALI predictions for the
  Dai dataset. \textbf{A} - a plot of predicted versus observed
  log(SALI) values for the training set and the hold out
  set. \textbf{B} - a summary of the training set, where we plot the
  structural difference versus the logarithm of the ratio of the
  activities for each pair of molecules
  in the prediction set. Points are shaded by their absolute
  residual. \textbf{C} - a box plot summarizing the distribution of  
  residuals associated with predictions from each of the hold out
  molecules.}, figure, botcap,
label={fig:tset-pset-dai}]    
{c}
{}
{\includegraphics[width=\linewidth]{tset-pset-dai}}

\ctable[cap={}, caption={The hold out molecules for the Dai dataset
  and training set members with which the hold outs exhibit predicted
  activity cliffs. Bold numbers are ChEMBL MOLREGNO values and numbers
  in parentheses are the absolute prediction
  residual in log(SALI) units.}, figure, botcap,
label={fig:dai-comparison}]    
{c}
{}  
{\includegraphics[width=\linewidth]{dai-comparison}}

\ctable[cap={}, caption={Distribution of log(SALI) values for the
  three ChEMBL datasets.}, figure, botcap,
label={fig:logsalidist}]    
{c}
{}  
{\includegraphics[width=\linewidth]{logsalidist}}

\end{document}

### 2012-07-10 submit/paper.tex
first commit

--- 2012-07-10 paper.tex
@@
 \begin{abstract}
-  The notion of activity cliffs is an intuitive approach to characterizing structural features that
-  play a key role in modulating biological activity of a molecule. A variety of methods have been
-  described to quantitatively characterize activity cliffs, such as SALI and SARI. However, these
-  methods are primarily retrospective in nature; highlighting cliffs that are already present in the
-  dataset. The current study focuses on employing a pairwise characterization of a dataset to train
-  a model to predict whether a new molecule will exhibit an activity cliff with one or more members
-  of the dataset. The approach is based on predicting a value for pairs of objects rather than the
-  individual objects themselves (and thus allows for robust models even for small SAR datasets). We
-  extracted structure-activity data for several ChEMBL assays and developed random forest models to
-  predict SALI values, from pairwise combinations of molecular descriptors. The models exhibited
-  reasonable RMSE's though, surprisingly, performance on the more significant cliffs tended to be
-  better than on the lesser ones. Our results indicate that the predictive models are able to
-  prioritize molecules in terms of their ability to activity cliffs and can be used as a way to
-  extend an observed activity landscape to identify molecules that could lead to significant
-  improvements in activity. Conversely, the method can be used to deprioritize molecules that are
-  predicted to show an activity cliff with a potent molecule already in the dataset.
+  The notion of activity cliffs is an intuitive approach to
+  characterizing structural features that play a key role in
+  modulating biological activity of a molecule. A variety of methods
+  have been described to quantitatively characterize activity cliffs,
+  such as SALI and SARI. However, these methods are restrictive in the
+  sense that they focus on exploratory analyses; highlighting cliffs
+  that are already present in the dataset. The current study focuses
+  on using activity cliffs in a dataset to train a model to predict
+  whether a new molecule will exhibit an activity cliff with one or
+  more members of the dataset. The approach is based on predicting
+  pairs of objects rather than the individual objects themselves (and
+  thus allows for robust models even for small SAR datasets). We
+  extracted structure-activity data for several ChEMBL assays and
+  developed random forest models to predict SALI values, from pairwise
+  combinations of molecular descriptors. The models exhibited
+  reasonable RMSE's though, surprisingly, performance on the more
+  significant cliffs tended to be better than on the lesser ones. Our
+  results indicate that the predictive models are able to prioritize
+  molecules in terms of their ability to activity cliffs and can be
+  used as a way to extend an observed activity landscape to identify
+  molecules that could lead to significant improvements in
+  activity. Conversely, the method can be used to deprioritize
+  molecules that are predicted to show an activity cliff with a potent
+  molecule already in the dataset.
 \end{abstract}
@@
 
-The landscape paradigm for SAR data was first proposed 20 years ago \cite{Johnson:1990ys} and has
-recently seen a resurgence with a number of studies describing new ways to quantify and visualize
-activity landscapes. When SAR data is viewed as a landscape, with the X-Y plane representing
-structural characteristics (which will usually be a 2-dimensional representation of a
-multi-dimensional descriptor space) and the Z-axis representing the observed activities, one can
-identify two broad types of regions on the landscape - smooth rolling regions, corresponding to set
-of molecules exhibiting continuous SAR (i.e., similar structures and similar activities) and rough,
-gorge-like regions (i.e., very similar structures, but large differences in activity) corresponding
-to molecules that exhibit SAR discontinuity. The latter have also been term activity cliffs
-\cite{Maggiora:2006aa}. From a medicinal chemistry point of view, the latter regions of a landscape
-can be the most interesting as they can provide insight into structural features that are key to
-improving (or conversely reducing) potency. There is a rich history of methods that have correlated
-structural differences with corresponding differences in activity -- matched molecular pairs
-\cite{Leach:2006aa}, SAS maps \cite{Maggiora:2011kx} and more recently SALI \cite{Guha:2008aa} and
-SARI \cite{Peltason:2007aa}. Both SALI and SARI focus on numerically characterizing a structure
-activity landscape. The former is defined for a pair of molecules as
+The landscape paradigm for SAR data was first proposed 20 years ago
+\cite{Johnson:1990ys} and recently has seen a resurgence with a number
+of studies describing new ways to quantify and visualize activity
+landscapes. When SAR data is viewed as a landscape, with the X-Y plane
+representing structural characteristics (which will usually be a
+2-dimensional representation of a multi-dimensional descriptor space)
+and the Z-axis representing the observed activities, one can identify
+two broad types of regions on the landscape - smooth rolling regions,
+corresponding to set of molecules exhibiting continuous SAR (i.e.,
+similar structures and similar activties) and rough, gorge-like
+regions (i.e., very similar structures, but large differences in
+activity) corresponding to molecules that exhibit SAR
+discontinuity. The latter have also been term activity cliffs
+\cite{Maggiora:2006aa}. From a medicinal chemistry point of view, the
+latter regions of a landscape can be the most interesting as they can
+provide insight into structural features that are key to improving (or
+conversely reducing) potency. There is a rich history of methods that
+have correlated structural differences with corresponding differences
+in activity - matched molecular pairs \cite{Leach:2006aa}, SAS maps
+\cite{Maggiora:2011kx} and more recently SALI \cite{Guha:2008aa}
+and SARI \cite{Peltason:2007aa}. Both SALI and SARI focus on
+numerically characterizing a structure activity landscape. The former
+is defined for a pair of molecules as
 \begin{equation}
@@
 \end{equation}
-where $A_i$ and $A_j$ represent the observed activities of molecules $i$ and $j$, and $sim(i,j)$
-represents the structural similarity between the two molecules (usually based on some form of
-fingerprint similarity). Using the SALI metric, one can take a collection of $n$ molecules and
-represent them as an $n \times n$ matrix of SALI values - larger values representing more
-significant activity cliffs. The SARI approach is based on a score defined as
+where $A_i$ and $A_j$ represent the observed activities of molecules
+$i$ and $j$, and $sim(i,j)$ represents the structural similarity
+between the two molecules (usually based on some form of fingerprint
+similarity). Using the SALI metric, one can take a collection of $n$
+molecules and represent them as an $n \times n$ matrix of SALI values
+- larger values representing more significant activity cliffs. The
+SARI approach is based on a score defined as 
 \begin{equation}
@@
 \end{equation}
-where the individual score terms are derived on the basis of potency and pairwise similarities. The
-reader is referred to Ref. \citen{Peltason:2007aa} for a detailed discussion of this approach.
+where the individual score terms are derived on the basis of potency
+and pairwise similarities. The reader is referred to
+Ref. \citen{Peltason:2007aa} for a detailed discussion of this
+approach.
 
-In either case, one can take the numerical values and visualize them in a variety of ways ranging
-from heatmaps of SALI matrices to network representations \cite{Wawer:2008aa,Guha:2008aa}. These
-visualizations then allow the user to explore the landscape, quickly identifying a range of activity
-cliffs, which an then be examined in detail. Apart from identifying individual activity cliffs, a
-variety of other SAR constructs, such as ``activity ridges'' \cite{Vogt:2011bs} and multitarget
-landscapes\cite{Dimova:2011fk} can also be identified and characterized.
+In either case, one can take the numerical values and visualize them
+in a variety of ways ranging from heatmaps of SALI matrices to network
+representations \cite{Wawer:2008aa,Guha:2008aa}. These visualizations
+then allow the user to explore the landscape, quickly identifying a
+range of activity cliffs, which an then be examined in detail. Apart
+from identifying individual activity cliffs, a variety of other SAR
+constructs, such as ``activity ridges'' \cite{Vogt:2011bs} and
+multitarget landscapes\cite{Dimova:2011fk} can also be identified and
+characterized.
 
-A feature common to all, recently published work on activity landscapes is that they are primarily
-retrospective. That is, the methodologies developed are used to analyze SAR datasets for which
-activities have already been experimentally obtained. For example, using the SALI, one can
-characterize SAR patterns in a dataset but does not provide insight into whether a new molecule may
-be part of an activity cliff, with respect to the original dataset. A number of applications have
-attempted to extract SAR rules based on the landscape (e.g., similarity potency trees
-\cite{Wawer:2010fv} and multi-target landscape analyses \cite{Hu:2010kl}) or directly identify
-structural modifications that lead to activity cliffs \cite{Wassermann:2010oq, Peltason:2009vn}.
+A feature common to all, recently published work on activity
+landscapes is that they are primarily retrospective. That is, the
+methodologies developed are used to analyze SAR datasets for which
+activities have already been experimentally obtained. For example,
+using the SALI, one can characterize SAR patterns in a dataset but
+does not provide insight into whether a new molecule may be part of an
+activity cliff, with respect to the original dataset. A number of
+applications have attempted to extract SAR rules based on the
+landscape (e.g., similarity potency trees \cite{Wawer:2010fv} and
+multi-target landscape analyses \cite{Hu:2010kl}) or directly identify
+structural modifications that lead to activity cliffs
+\cite{Wassermann:2010oq, Peltason:2009vn}.
 
@@
 \label{sec:motivation}
-The preceding discussion highlights the utility of retrospective analyses of SAR data using the
-activity landscape paradigm. But equally, if not more, interesting is determining whether a new,
-untested molecule might be an activity cliff in the context of the original dataset. For the
-remainder of this work we focus on the use of SALI to quantify activity landscapes. More
-specifically, the ability to \emph{predict} SALI values would be useful as it would allow us to both
-fill in empty regions of an activity landscape as well as extend a structure-activity
-landscape. Note that this approach to expanding the extent of a SAR dataset does not lend itself to
-scaffold hopping since the premise of scaffold hopping is that one generates new cores, which differ
-substantially from the starting structure.
+The preceding discussion highlights the utility of retrospective
+analyses of SAR data using the activity landscape paradigm. But
+equally, if not more, interesting is determining whether a new,
+untested molecule might be an activity cliff in the context of the
+original dataset. For the remainder of this work we focus on the use
+of SALI to quantify activity landscapes. More specifically, the
+ability to \emph{predict} SALI values would be useful as it would
+allow us to both fill in empty regions of an activity landscape as
+well as extend a structure-activity landscape. Note that this approach
+to expanding the extent of a SAR dataset does not lend itself to
+scaffold hopping since the premise of scaffold hopping is that one
+generates new cores, which differ substantially from the starting
+structure.
 
-In traditional QSAR modeling approaches, one simply predicts the activity of a new molecule and
-would then evaluate the SALI (or SARI or some other measure) to determine whether the molecule leads
-to an activity cliff. However, the fact that an activity cliff represents a SAR
-discontinuity\cite{Maggiora:2006aa} implies that most statistical and machine learning methods will
-be unlikely to predict very different activities for two structurally similar molecules. In other
-words, a new molecule, similar to a subset of the training set, will tend to have a predicted value
-that is similar to those molecules, rather than a drastically different value.
+In traditional QSAR modeling approaches, one simply predicts the
+activity of a new molecule and would then evaluate the SALI (or SARI
+or some other measure) to determine whether the molecule leads to an
+activity cliff. However, the fact that an activty cliff represents a
+SAR discontinuity\cite{Maggiora:2006aa} implies that most statistical
+and machine learning methods will be unlikely to predict very
+different activities for two structurally similar molecules. In other
+words, a new molecule, similar to a subset of the training set, will
+tend to have a predicted value that is similar to those molecules,
+rather than a drastically different value.
 
-An alternative approach, that is the focus of this paper, is to directly predict SALI values for
-pairs of molecules. Thus rather than predict individual activities, we predict SALI values for pairs
-of molecules. This approach is somewhat similar to the SPREAD method \cite{Scheiber:2009fk} which
-identified substructures that were predictive of activity differences. However, our solution
-considers both activity differences and structural similarities. As a result, instead of ranking
-compounds in terms of their predicted activity, we instead rank a compound in terms of its predicted
-SALI; i.e., its predicted ability to exhibit an activity cliff when paired with other molecules in
-the dataset. This approach could be useful when deciding how far to extend an analog series as well
-as prioritizing scaffolds for further study.
+An alternative approach, that is the focus of this paper, is to
+directly predict SALI values for pairs of molecules. Thus rather than
+predict individual actvities, we predict SALI values for pairs of
+molecules. This approach is somewhat similar to the SPREAD method
+\cite{Scheiber:2009fk} which identifed substructural features that
+were predictive of activity differences. However, our solution
+considers both activity differences and structural similarities. As a
+result, instead of ranking compounds in terms of their predicted
+activity, we instead rank a compound in terms of its predicted SALI;
+i.e., its predicted ability to exhibit an actvity cliff when paired
+with other molecules in the dataset. This approach could be useful
+when deciding how far to extend an analog series as well as
+prioritizing scaffolds for further study.
 
-This does not completely alleviate the problem of discontinuities, since SALI values are infinite
-when the $T_c$ is 1.0.  However, predicting SALI values allows us to work with smaller datasets
-(since the objects to predict are pairs of molecules), that would ordinarily lead to unreliable
-models if we were working with activities. Of course, this means that the approach is not practical
-for very large datasets.
+This does not completely alleviate the problem of discontinuities,
+since SALI values are infinite when the $T_c$ is 1.0.  However,
+predicting SALI values allows us to work with smaller datasets (since
+the objects to predict are pairs of molecules), that would ordinarily
+lead to unreliable models if we were working with activities. Of
+course, this means that the approach is not practical for very large
+datasets.
 
-The paper is organized as follows. Section \ref{sec:datasets} describes the datasets used in this
-study. Section \ref{sec:methodology} presents the methodology we employ to predict activity cliffs
-and Section \ref{sec:applications} discusses the results of the predictive models. Finally, Sections
-\ref{sec:discussion} and \ref{sec:conclusions} discusses some of the issues underlying this approach
-and possible extensions of this work.
+The paper is organized as follows. Section \ref{sec:datasets}
+describes the datasets used in this study. Section
+\ref{sec:methodology} presents the methodology we employ to predict
+activity cliffs and Section \ref{sec:applications} discusses the
+results of the predictive models. Finally, Sections
+\ref{sec:discussion} and \ref{sec:conclusions} dicusses some of the
+issues underlying this approach and possible extensions of this work.
 
@@
 \label{sec:datasets}
-For this study we considered a number of datasets, which are summarized in Table \ref{tab:data}. The
-Cavalli dataset was employed in our previous studies consisted of 30 molecules studied by Cavalli et
-al\cite{Cavalli:2002aa} as possible hErg inhibitors using a pharmacophore modeling approach. The
-remaining datasets were obtained from ChEMBL. All three assays involved direct binding to a human
-target and we considered the subset of molecules in each assay that had non-censored experimental
-values. The Costanzo dataset\cite{Costanzo:2005uq} consisted of 60 $\alpha$-ketoheterocyclic
-inhibitors of $\alpha$-thrombin. The reported $\mathrm{IC_{50}}$ values ranged from 3 nM to 82
-$\mu$M. The Kalla dataset\cite{Kalla:2006kx} consisted of 38 8-(C-4-pyrazolyl) xanthines, identified
-as antagonists of the A2B adenosine receptor. The reported $K_i$ values ranged from 0.9 nM to 42
-$\mu$M. Finally, the Dai dataset\cite{Dai:2007kl} consisted of 44 3-aminoindazole derivatives
-studied for their ability to inhibit the VEGF and PDGF receptor families, with $\mathrm{IC_{50}}$
-values ranging from 3 nM to 12~$\mu$M.
+For this study we considered a number of datasets, which are
+summarized in Table \ref{tab:data}. The Cavalli dataset was employed
+in our previous studies consisted of 30 molecules studied by Cavalli
+et al\cite{Cavalli:2002aa} as possible hErg inhibitors using a
+pharmacophore modeling approach. The remaining datasets were obtained
+from ChEMBL. All three assays involved direct binding to a human
+target and we considered the subset of molecules in each assay that
+had non-censored experimental values. The Costanzo
+dataset\cite{Costanzo:2005uq} consisted of 60
+$\alpha$-ketoheterocyclic inhibitors of $\alpha$-thrombin. The
+reported $\mathrm{IC_{50}}$ values ranged from 3 nM to 82 $\mu$M. The
+Kalla dataset\cite{Kalla:2006kx} consisted of 38 8-(C-4-pyrazolyl)
+xanthines, identified as antagonists of the A2B adenosine
+receptor. The reported $K_i$ values ranged from 0.9 nM to 42
+$\mu$M. Finally, the Dai dataset\cite{Dai:2007kl} consisted of 44
+3-aminoindazole derivatives studied for their ability to inhibit the
+VEGF and PDGF receptor families, with $\mathrm{IC_{50}}$ values
+ranging from 3 nM to 12~$\mu$M.
 
@@
 \label{sec:methodology}
-As noted above, the problem of identifying activity cliffs involving new molecules can be reduced to
-predicting the SALI value for the new molecule and a pre-existing molecule. Note that this does not
-result in an activity prediction for the new molecule; rather, it allows us to rank a set of new
-molecules in terms of their predicted ability to exhibit a significant activity difference from one
-or more of the molecules in the training set.
+As noted above, the problem of identifying activity cliffs involving
+new molecules can be reduced to predicting the SALI value for the new
+molecule and a pre-existing molecule. Note that this does not result
+in an activity prediction for the new molecule; rather, it allows us
+to rank a set of new molecules in terms of their predicted ability to
+exhibit a significant activity difference from one or more of the
+molecules in the training set.
 
@@
 
-\bibliography{paper}
+%%\bibliography{paper}
+\providecommand{\refin}[1]{\\ \textbf{Referenced in:} #1}
+\begin{thebibliography}{10}
+
+\bibitem{Johnson:1990ys}
+Johnson,~M.;\ \ Maggiora,~G. \textit{Concepts and Applications of Molecular
+  Similarity;} John Wiley \& Sons: New York, 1990.
+
+\bibitem{Maggiora:2006aa}
+Maggiora,~G.~M.  On Outliers and Activity Cliffs--Why {QSAR} Often
+  Disappoints..  \textit{J.~Chem.~Inf.~Model.} \textbf{2006,} \textsl{46,}
+  1535--1535.
+
+\bibitem{Leach:2006aa}
+Leach,~A.;\ \ Jones,~H.;\ \ Cosgrove,~D.;\ \ Kenny,~P.;\ \ Ruston,~L.;\ \
+  MacFaul,~P.;\ \ Wood,~J.;\ \ Colclough,~N.;\ \ Law,~B.  Matched Molecular
+  Pairs as a Guide in the Optimization of Pharmaceutical Properties; a Study of
+  Aqueous Solubility, Plasma Protein Binding and Oral Exposure.
+  \textit{J.~Med.~Chem.} \textbf{2006,} \textsl{49,} 6672--6682.
+
+\bibitem{Maggiora:2011kx}
+Shanmugasundaram,~V.;\ \ Maggiora,~G.  Characterizing Property and Activity
+  Landscapes Using an Information-Theoretic Approach.   In  \textit{CINF-032.
+  222nd ACS National Meeting, Chicago, IL, United States}; American Chemical
+  Society: Washington, D.C., 2001.
+
+\bibitem{Guha:2008aa}
+Guha,~R.;\ \ Van~Drie,~J.  The Structure-Activity Landscape Index:
+  {I}dentifying and Quantifying Activity-Cliffs.  \textit{J.~Chem.~Inf.~Model.}
+  \textbf{2008,} \textsl{48,} 646--658.
+
+\bibitem{Peltason:2007aa}
+Peltason,~L.;\ \ Bajorath,~J.  SAR Index: Quantifying the Nature of
+  Structure-Activity Relationships.  \textit{J.~Med.~Chem.} \textbf{2007,}
+  \textsl{50,} 5571--5578.
+
+\bibitem{Wawer:2008aa}
+Wawer,~M.;\ \ Peltason,~L.;\ \ Weskamp,~N.;\ \ Teckentrup,~A.;\ \ Bajorath,~J.
+  Structure-Activity Relationship Anatomy by Network-like Similarity Graphs and
+  Local Structure-Activity Relationship Indices.  \textit{J.~Med.~Chem.}
+  \textbf{2008,} \textsl{51,} 6075--6084.
+
+\bibitem{Vogt:2011bs}
+Vogt,~M.;\ \ Huang,~Y.;\ \ Bajorath,~J.  From Activity Cliffs to Activity
+  Ridges: {I}nformative Data Structures for SAR Analysis.
+  \textit{J.~Chem.~Inf.~Model.} \textbf{2011,} \textsl{51,} 1848--1856.
+
+\bibitem{Dimova:2011fk}
+Dimova,~D.;\ \ Wawer,~M.;\ \ Wassermann,~A.~M.;\ \ Bajorath,~J.  Design of
+  Multitarget Activity Landscapes That Capture Hierarchical Activity Cliff
+  Distributions.  \textit{J.~Chem.~Inf.~Model.} \textbf{2011,} \textsl{51,}
+  258--266.
+
+\bibitem{Wawer:2010fv}
+Wawer,~M.;\ \ Bajorath,~J.  Similarity-potency trees: {A} method to search for
+  {SAR} information in compound data sets and derive {SAR} rules.
+  \textit{J.~Chem.~Inf.~Model.} \textbf{2010,} \textsl{50,} 1395--1409.
+
+\bibitem{Hu:2010kl}
+Hu,~Y.;\ \ Bajorath,~J.  Molecular Scaffolds with High Propensity to Form
+  Multi-Target Activity Cliffs.  \textit{J.~Chem.~Inf.~Model.} \textbf{2010,}
+  \textsl{50,} 500--510.
+
+\bibitem{Wassermann:2010oq}
+Wassermann,~A.~M.;\ \ Bajorath,~J.  Chemical Substitutions that Introduce
+  Activity Cliffs Across Different Compound Classes and Biological Targets.
+  \textit{J.~Chem.~Inf.~Model.} \textbf{2010,} \textsl{50,} 1248-1256.
+
+\bibitem{Peltason:2009vn}
+Peltason,~L.;\ \ Weskamp,~N.;\ \ Teckentrup,~A.;\ \ Bajorath,~J.  Exploration
+  of Structure-Activity Relationship Determinants in Analogue Series.
+  \textit{J.~Med.~Chem.} \textbf{2009,} \textsl{52,} 3212--3224.
+
+\bibitem{Scheiber:2009fk}
+Scheiber,~J.;\ \ Jenkins,~J.~L.;\ \ Bender,~A.;\ \ Milik,~M.;\ \ Mikhalov,~D.;\
+  \ Sukuru,~S.;\ \ Cornett,~B.;\ \ Whitebread,~S.;\ \ Laszlo,~U.;\ \
+  Davies,~J.;\ \ Glick,~M.  {SPREAD} - exploiting chemical features that cause
+  differential activity behavior.  \textit{Stat.~Anal.~Data Mining}
+  \textbf{2009,} \textsl{2,} 115-122.
+
+\bibitem{Cavalli:2002aa}
+Cavalli,~A.;\ \ Poluzzi,~E.;\ \ De~Ponti,~F.;\ \ Recanatini,~M.  Toward a
+  Pharmacophore for Drugs Inducing the Long {QT} Syndrome: Insights from a
+  {CoMFA} Study of {HERG K+} Channel Blockers.  \textit{J.~Med.~Chem.}
+  \textbf{2002,} \textsl{45,} 3844--3853.
+
+\bibitem{Costanzo:2005uq}
+Costanzo,~M.~J.;\ \ Almond,~Jr,~H.~R.;\ \ Hecker,~L.~R.;\ \ Schott,~M.~R.;\ \
+  Yabut,~S.~C.;\ \ Zhang,~H.-C.;\ \ Andrade-Gordon,~P.;\ \ Corcoran,~T.~W.;\ \
+  Giardino,~E.~C.;\ \ Kauffman,~J.~A.;\ \ Lewis,~J.~M.;\ \ de~Garavilla,~L.;\ \
+  Haertlein,~B.~J.;\ \ Maryanoff,~B.~E.  In-depth study of tripeptide-based
+  alpha-ketoheterocycles as inhibitors of thrombin. {E}ffective utilization of
+  the {S1'} subsite and its implications to structure-based drug design.
+  \textit{J.~Med.~Chem.} \textbf{2005,} \textsl{48,} 1984--2008.
+
+\bibitem{Kalla:2006kx}
+Kalla,~R.~V.;\ \ Elzein,~E.;\ \ Perry,~T.;\ \ Li,~X.;\ \ Palle,~V.;\ \
+  Varkhedkar,~V.;\ \ Gimbel,~A.;\ \ Maa,~T.;\ \ Zeng,~D.;\ \ Zablocki,~J.
+  Novel 1,3-disubstituted 8-(1-benzyl-1H-pyrazol-4-yl) xanthines: high affinity
+  and selective A2B adenosine receptor antagonists.  \textit{J.~Med.~Chem.}
+  \textbf{2006,} \textsl{49,} 3682--3692.
+
+\bibitem{Dai:2007kl}
+Dai,~Y. \textit{et al.}\   Discovery of
+  N-(4-(3-amino-1H-indazol-4-yl)phenyl)-N'-(2-fluoro-5-methylphenyl)urea
+  (ABT-869), a 3-aminoindazole-based orally active multitargeted receptor
+  tyrosine kinase inhibitor.  \textit{J.~Med.~Chem.} \textbf{2007,}
+  \textsl{50,} 1584--1597.
+
+\bibitem{Breiman:1984aa}
+Breiman,~L.;\ \ Friedman,~J.;\ \ Olshen,~R.;\ \ Stone,~C.
+  \textit{Classification and Regression Trees;} Chapman \& Hall/CRC: Boca
+  Raton, FL, 1984.
+
+\bibitem{Breiman:2001nx}
+Breiman,~L.  Statistical Modeling: Two Cultures.  \textit{Stat.~Sci.}
+  \textbf{2001,} \textsl{16,} 199--231.
+
+\bibitem{r}
+{R Development Core Team}, ``{R}: {A} Language and Environment for Statistical
+  Computing'',  R Foundation for Statistical Computing, Vienna, Austria, 2008
+  {ISBN} 3-900051-07-0.
+
+\end{thebibliography}
 

### 2012-07-10 rebuttal.tex
finished proof reading the paper

\documentclass[letterpaper, 12pt]{article}
\usepackage{url}
\usepackage{ctable}
\usepackage{amsmath,amssymb}      % for \AA etc
\usepackage[onehalfspacing]{setspace}
\usepackage{graphicx}
\usepackage{times}

\DeclareGraphicsExtensions{.eps,.png}
\onehalfspacing

\setlength{\parindent}{0pt}
\setlength{\parskip}{1em}
\usepackage{anysize}

\begin{document}

I would like to thank the reviewers for their thorough comments on the original submissions. I have
prepared a revised manuscript that address the comments and my responses are detailed below.

\fbox{\textbf{Reviewer: 1}}

Recommendation: Publish after major revisions noted.

Comments:
The prediction of activity cliffs is a nice idea that would further extend activity landscape modeling. This is attempted by predicting SALI values for compound pairs on the basis of training data. Unfortunately, there are a number of technical issues to note and, more importantly, methodological concerns that would need to be carefully addressed before publication of this study might be considered.

First, a few general comments: 
(1) Please, note that there are numerous typographical errors throughout the manuscript (which is quite disturbing). Unfortunately, the presentation of the work is sloppy at different levels.       
(2) Chosen methods and descriptors should at least be referenced. For example, no references are provided for BCI and CDK.

Second, there are shortcomings at the technical level:
(3) The derivation of RF models is essentially not described (except of making reference to default parameters settings in R). It remains largely unclear how these models were developed. 
(4) The selection of training and test sets should be clearly specified in the Methods section. For example, how were the RF models trained and evaluated for which predictions of pair-wise SALI values are reported in Figure 2? It is hoped that the reported SALI values were not predicted for the training data.
(5) No support is provided for the assumption that the RF models “are relatively robust even for datasets of small size”. In fact, the author later on states several times that limited model quality might be due to the use of small compound sets. Appropriate validation will be essential here.
(6) There is only limited statistical assessment/validation of the models. Statements made in the text that R2 values are “reasonable” and the models perform “relatively well” are difficult to reconcile. Rather, Table 2 indicates that the R2 values were rather poor in several instances.
(7) It appears that there are systematic prediction errors for pairs. At low value ranges (that are not relevant for activity cliff assessment), SALI values are consistently over-predicted. By contrast, at high value ranges (that are relevant), SALI values are consistently under-predicted, in part significantly. This is not investigated, although the apparent errors at high value ranges are the perhaps most critical features of the reported RF models.    
(8) In the section “Extending a landscape”, predictions are finally reported for only three (!) molecules taken out of each data set. Carefully put, the predictions are heterogeneous. On the basis of the results reported for this little bit of evaluation, one would be hard pressed to make a case for the ability of the approach to predict activity cliffs with any certainty.

Third, there are major scientific concerns:
(9) For the prediction of activity cliffs, SALI values must be calculated for compound pairs. This is the central idea of the approach. Accordingly, fingerprint descriptors and potency values must be combined in some ways. 
(9.1.) It is nowhere stated how activity values are treated for compounds forming a pair. Are activity differences calculated at the pair level? If so, how are activity differences compared for the calculation of pair-based SALI scores - as a difference of differences? It should be noted predicted potency values in Figures 4-6 are essentially all over the place.
(9.2.) To obtain a fingerprint descriptor for a pair, the author aggregates fingerprints of individual molecules, finally, by simple averaging. For comparisons of pairs, Tanimoto similarity is then calculated for averaged fingerprints, which presents an artificial assessment of pair similarity (one can easily come up with a few hypothetical compound similarity relationships that make this assessment questionable at best). As a consequence, activity cliffs are considered for which Tanimoto similarity calculations yield values between 0.2 and 0.3! Thus, on the basis of these similarity calculations, one could never decide which level of similarity might be cliff relevant. In this respect, the descriptions of activity cliff by the author are quite telling – he speaks of “predicted” versus “true” activity cliffs, of “relatively accurate” activity cliffs, and of “most significant activity cliffs that are in fact not very significant activity cliffs in an absolute sense”, and so on. 
Furthermore, there might be many instances of activity and fingerprint similarity relationships between pairs of compounds that might yield high pair-based SALI scores dominated by either the activity or similarity component, without forming a “true” activity cliff. The conclusion of this reviewer is that it is currently not possible to capture activity cliffs with any degree of certainty on the basis of the pair calculations, as reported.
(10) It also remains largely obscure how an activity landscape should be extended following this approach by compounds for which no activity information is available. Should the activity first be predicted by an RF model and then used for pair-based SALI score calculations? One would hope not.

In summary, there will be quite a bit of work required to get this manuscript in shape for publication, both in terms of its presentation and technical content. However, as long as points (9) and (10) are not thoroughly addressed scientifically, this work should not be published.

To give the author the benefit of the doubt -and credit for the underlying idea- this reviewer calls for “Major Revisions”, although the study might as well be considered premature at this point.

Additional Questions:
Please rate the quality of the science reported in this paper (10 - High Quality/ 1 - Low Quality): 4

Please rate the overall importance of this paper to the field of chemical information or modeling (10 - High Importance / 1 - Low Importance): 6

\fbox{\textbf{Reviewer: 2}}

\textbf{1.  The manuscript is poorly written.  There are many typos, grammatical errors, missing
  references, and incomplete descriptions of methods and results.}

The text has been thoroughly proof read to remove typos and incorrect grammar. References have been
added and methods have been described more clearly.

\textbf{2.  Qualitative assessments of results are not acceptable.  The authors should give
  statistical evaluations and corresponding metrics for models and their predictions.}

\textbf{3.  The tests sets are too small to make any meaningful conclusions.}

\textbf{4.  SALI values needs to be determined for pairs of molecules - is that not the essence of
  the methodology?}

\textbf{5.  There is no discussion/direction of how to use the methodology in new applications.}
\end{document}

### 2012-07-10 paper.tex
finished proof reading the paper

@@
   of the dataset. The approach is based on predicting a value for pairs of objects rather than the
-  individual objects themselves (and thus allows for robust models even for small SAR datasets). We
-  extracted structure-activity data for several ChEMBL assays and developed random forest models to
-  predict SALI values, from pairwise combinations of molecular descriptors. The models exhibited
-  reasonable RMSE's though, surprisingly, performance on the more significant cliffs tended to be
-  better than on the lesser ones. Our results indicate that the predictive models are able to
-  prioritize molecules in terms of their ability to activity cliffs and can be used as a way to
-  extend an observed activity landscape to identify molecules that could lead to significant
-  improvements in activity. Conversely, the method can be used to deprioritize molecules that are
-  predicted to show an activity cliff with a potent molecule already in the dataset.
+  individual objects themselves (and thus allows for robust models even for small structure-activity
+  relationship datasets). We extracted structure-activity data for several ChEMBL assays and
+  developed random forest models to predict SALI values, from pairwise combinations of molecular
+  descriptors. The models exhibited reasonable RMSE's though, surprisingly, performance on the more
+  significant cliffs tended to be better than on the lesser ones. Our results indicate that the
+  predictive models are able to prioritize molecules in terms of their ability to activity cliffs
+  and can be used as a way to extend an observed activity landscape to identify molecules that could
+  lead to significant improvements in activity. Conversely, the method can be used to deprioritize
+  molecules that are predicted to show an activity cliff with a potent molecule already in the
+  dataset.
 \end{abstract}
@@
 
-The landscape paradigm for SAR data was first proposed 20 years ago \cite{Johnson:1990ys} and has
-recently seen a resurgence with a number of studies describing new ways to quantify and visualize
-activity landscapes. When SAR data is viewed as a landscape, with the X-Y plane representing
-structural characteristics (which will usually be a 2-dimensional representation of a
-multi-dimensional descriptor space) and the Z-axis representing the observed activities, one can
-identify two broad types of regions on the landscape - smooth rolling regions, corresponding to set
-of molecules exhibiting continuous SAR (i.e., similar structures and similar activities) and rough,
-gorge-like regions (i.e., very similar structures, but large differences in activity) corresponding
-to molecules that exhibit SAR discontinuity. The latter have also been term activity cliffs
-\cite{Maggiora:2006aa}. From a medicinal chemistry point of view, the latter regions of a landscape
-can be the most interesting as they can provide insight into structural features that are key to
-improving (or conversely reducing) potency. There is a rich history of methods that have correlated
-structural differences with corresponding differences in activity -- matched molecular pairs
-\cite{Leach:2006aa}, SAS maps \cite{Maggiora:2011kx} and more recently SALI \cite{Guha:2008aa} and
-SARI \cite{Peltason:2007aa}. Both SALI and SARI focus on numerically characterizing a structure
+The landscape paradigm for structure-activity relationship (SAR) data was first proposed 20 years
+ago \cite{Johnson:1990ys} and has recently seen a resurgence with a number of studies describing new
+ways to quantify and visualize activity landscapes. When SAR data is viewed as a landscape, with the
+X-Y plane representing structural characteristics (which will usually be a 2-dimensional
+representation of a multi-dimensional descriptor space) and the Z-axis representing the observed
+activities, one can identify two broad types of regions on the landscape - smooth rolling regions,
+corresponding to set of molecules exhibiting continuous SAR (i.e., similar structures and similar
+activities) and rough, gorge-like regions (i.e., very similar structures, but large differences in
+activity) corresponding to molecules that exhibit SAR discontinuity. The latter have also been term
+activity cliffs \cite{Maggiora:2006aa}. From a medicinal chemistry point of view, the latter regions
+of a landscape can be the most interesting as they can provide insight into structural features that
+are key to improving (or conversely reducing) potency. There is a rich history of methods that have
+correlated structural differences with corresponding differences in activity -- matched molecular
+pairs \cite{Leach:2006aa}, SAS maps \cite{Maggiora:2011kx} and more recently SALI \cite{Guha:2008aa}
+and SARI \cite{Peltason:2007aa}. Both SALI and SARI focus on numerically characterizing a structure
 activity landscape. The former is defined for a pair of molecules as
@@
 Cavalli dataset was employed in our previous studies consisted of 30 molecules studied by Cavalli et
-al\cite{Cavalli:2002aa} as possible hErg inhibitors using a pharmacophore modeling approach. The
-remaining datasets were obtained from ChEMBL. All three assays involved direct binding to a human
-target and we considered the subset of molecules in each assay that had non-censored experimental
-values. The Costanzo dataset\cite{Costanzo:2005uq} consisted of 60 $\alpha$-ketoheterocyclic
-inhibitors of $\alpha$-thrombin. The reported $\mathrm{IC_{50}}$ values ranged from 3 nM to 82
-$\mu$M. The Kalla dataset\cite{Kalla:2006kx} consisted of 38 8-(C-4-pyrazolyl) xanthines, identified
-as antagonists of the A2B adenosine receptor. The reported $K_i$ values ranged from 0.9 nM to 42
-$\mu$M. Finally, the Dai dataset\cite{Dai:2007kl} consisted of 44 3-aminoindazole derivatives
-studied for their ability to inhibit the VEGF and PDGF receptor families, with $\mathrm{IC_{50}}$
-values ranging from 3 nM to 12~$\mu$M.
+al\cite{Cavalli:2002aa} as possible hErg inhibitors using a CoMFA modeling approach. The reported
+endpoint for the molecules was a pIC$_{50}$.The remaining datasets were obtained from ChEMBL. All
+three assays involved direct binding to a human target and we considered the subset of molecules in
+each assay that had non-censored experimental values. The Costanzo dataset\cite{Costanzo:2005uq}
+consisted of 60 $\alpha$-ketoheterocyclic inhibitors of $\alpha$-thrombin. The reported
+$\mathrm{IC_{50}}$ values ranged from 3 nM to 82 $\mu$M. The Kalla dataset\cite{Kalla:2006kx}
+consisted of 38 8-(C-4-pyrazolyl) xanthines, identified as antagonists of the A2B adenosine
+receptor. The reported $K_i$ values ranged from 0.9 nM to 42 $\mu$M. Finally, the Dai
+dataset\cite{Dai:2007kl} consisted of 44 3-aminoindazole derivatives studied for their ability to
+inhibit the VEGF and PDGF receptor families, with $\mathrm{IC_{50}}$ values ranging from 3 nM to
+12~$\mu$M.
 
@@
 As noted above, the problem of identifying activity cliffs involving new molecules can be reduced to
-predicting the SALI value for the new molecule and a pre-existing molecule. Note that this does not
+predicting the SALI value for the new molecule and a preexisting molecule. Note that this does not
 result in an activity prediction for the new molecule; rather, it allows us to rank a set of new
-molecules in terms of their predicted ability to exhibit a significant activity difference from one
-or more of the molecules in the training set.
+molecules in terms of their predicted ability to exhibit a significant activity difference with
+respect to one or more of the molecules in the training set.
 
-Given a training set of $N$ molecules, we generate a new training set
-of $\frac{N(N-1)}{2}$ objects, where each object is a pair of
-molecules from the original training set. The dependent variable for
-each pair $i,j$, is the SALI value, $S_{i,j}$. SALI values were
-evaluated using the 1051-bit BCI keyed fingerprints or the CDK
-1024-bit path fingerprints. Based on the
-definition of SALI, it is possible that a pair of molecules have a
-$T_c = 1.0$ resulting in infinite values. For such cases, we replaced
-the infinite value with the highest non-infinite SALI value for that
-dataset.
+Given a training set of $N$ molecules, we generate a new training set of $\frac{N(N-1)}{2}$ objects,
+where each object is a pair of molecules from the original training set. The dependent variable for
+each pair $i,j$, is the SALI value, $S_{i,j}$. SALI values were evaluated using the 1051-bit BCI
+keyed fingerprints (Digital Chemistry, UK) or the CDK\cite{Steinbeck:2006aa,Steinbeck:2003bh}
+1024-bit path fingerprints. Based on the definition of SALI, it is possible that a pair of molecules
+have a $T_c = 1.0$ resulting in infinite values. For such cases, we replaced the infinite value with
+the highest non-infinite SALI value for that dataset.
 
-The next step is to generate a set of independent variables. Since the
-new dataset consists of pairs of molecules, we consider the
-descriptors for the resultant objects as a function of the descriptor
-values of the individual molecules. For an object, representing the
-$i$'th and $j$'th molecules, its descriptor vector can be taken as the
-arithmetic mean of the descriptor vectors of the individual
-molecules. We denote this aggregation function as
-$f_{\textrm{mean}}$. Alternative functions that were investigated included
-the absolute difference of the individual descriptors, denoted by
-$f_{\textrm{diff}}$ and the geometric mean of the individual
-descriptor vectors, denoted by $f_{\textrm{geom}}$.
+The next step is to generate a set of independent variables. Since the new dataset consists of pairs
+of molecules, we consider the descriptors for the resultant objects as a function of the descriptor
+values of the individual molecules. For an object, representing the $i$'th and $j$'th molecules, its
+descriptor vector can be taken as the arithmetic mean of the descriptor vectors of the individual
+molecules. We denote this aggregation function as $f_{\textrm{mean}}$. Alternative functions that
+were investigated included the absolute difference of the individual descriptors, denoted by
+$f_{\textrm{diff}}$ and the geometric mean of the individual descriptor vectors, denoted by
+$f_{\textrm{geom}}$.
 
-Given the independent and dependent variables for the pairwise
-dataset, we can now proceed to model development. For this study we
-focused on the use of random forest models\cite{Breiman:1984aa}. This
-was motivated by the fact that such models can be proven to not
-overfit and the fact that the algorithm implicitly performs feature
-selection. As a result, this allows us to forgo an explicit feature
-selection step and work directly with the descriptor pool (after
-removal of correlated and constant descriptors). Furthermore, there is
-no reason, \emph{a priori} to assume that the underlying SARs are
-linear. A random forest model, being an algorithmic
-approach\cite{Breiman:2001nx} (as opposed to a distributional one such
-as linear regression) makes no such assumptions. We employed the  
-implementation of random forest from the \texttt{randomForest}
-package, in R 2.11.0\cite{r}, using the defult settings.
+Given the independent and dependent variables for the pairwise dataset, we can now proceed to model
+development. For this study we focused on the use of random forest models\cite{Breiman:1984aa}. This
+was motivated by the fact that such models tend not to overfit the training data (given a
+sufficiently large number of trees) and the fact that the algorithm implicitly performs feature
+selection. As a result, this allows us to forgo an explicit feature selection step and work directly
+with the descriptor pool (after removal of correlated and constant descriptors). Furthermore, there
+is no reason, \emph{a priori} to assume that the underlying SAR's are linear. A random forest model,
+being an algorithmic approach\cite{Breiman:2001nx} (as opposed to a distributional one such as
+linear regression) makes no such assumptions. We employed the implementation of random forest from
+the \texttt{randomForest} package, in R 2.11.0\cite{r}, using the default settings.
 
-In this work, we employed the CDK to evaluate 109 2D and
-constitutional descriptors for each individual molecule. For each
-dataset, we performed objective feature selection by removing
-descriptors with constant or near-constant values followed by removal
-of descriptors that are highly correlated with others (using an $R^2$
-cutoff of 0.8). The size of the final descriptor pools is summarized
-in Table \ref{tab:data}.
+In this work, we employed the CDK to evaluate 109 2D and constitutional descriptors for each
+individual molecule. For each dataset, we performed objective feature selection by removing
+descriptors with constant or near-constant values followed by removal of descriptors that are highly
+correlated with others (using an $R^2$ cutoff of 0.8). The sizes of the final descriptor pool for
+each dataset are summarized in Table \ref{tab:data}.
 
-One feature of this approach is that one can build relatively robust
-models even for datasets of small size - say, 20 molecules. Of course,
-a large training set is just one aspect of a reliable model and other
-considerations such as diversity, descriptor selection and so on still
-play an important role.
+A side effect of the proposed approach is that one can build relatively robust models for datasets of
+small size -- say, 20 molecules. Of course, a large training set is just one aspect of a reliable
+model and other considerations such as diversity, descriptor selection and so on still play an
+important role.
 
@@
 
-One question that arises from the preceding discussion is the
-possibility that we are duplicating information between the dependent
-and independent variables. More specifically, both the dependent
-variable (i.e., the SALI values) and the aggregated descriptors
-characterize the molecules' structure. One might therefore ask whether
-models built using such data perform over-optimistically. Given the
-nature of the descriptors used in the indepdent and dependent
-variables (the former being based on multiple atom and bond features
-and the latter derived from purel topological paths), we believe that such a
-phenomenon is unlikely.  Fig.~\ref{fig:desccor1} displays a histogram
-of the pairwise Pearson correlations between the SALI values and each
-of the descriptor values for the Cavalli\cite{Cavalli:2002aa} dataset,
-using the different aggregation functions described above. It is
-evident, that the highest $R^2$, between any of the descriptors and
-the SALI values is less than 0.15. Similar behavior was observed for
-all the datasets used in this study. These observations
-suggest that the problem of including correlated structural
-information simultaneously in the dependent and independent variables
-is minimal.
+The preceding discussion raises the issue of explicitly duplicating structural information in the
+independent and dependent variables. More specifically, both the dependent variable (i.e., the SALI
+values) and the aggregated descriptors characterize the molecules' structure. One might therefore
+ask whether models built using such data perform over-optimistically. Given the nature of the
+descriptors used in the independent and dependent variables (the former being based on multiple atom
+and bond features and the latter derived from purely topological paths), we believe that such a
+phenomenon is unlikely.  Fig.~\ref{fig:desccor1} displays a histogram of the pairwise Pearson
+correlations between the SALI values and each of the descriptor values for the
+Cavalli\cite{Cavalli:2002aa} dataset, using the different aggregation functions described above. It
+is evident, that the highest $R^2$, between any of the descriptors and the SALI values is less than
+0.15. Similar behavior was observed for all the datasets used in this study. These observations
+suggest that problems arising from the inclusion of structural information simultaneously in the
+dependent and independent variables is minimal.
 
@@
 \label{sec:applications}
-We first consider the application of the activity cliff prediction
-methodology to the Cavalli dataset. We developed three random forest
-models, using the three aggregation descriptor aggregation functions
-described in Section
-\ref{sec:methodology}. Fig.~\ref{fig:rfmodel-cavalli} presents the
-predicted versus observed SALI values from the three models and Table
-\ref{tab:rfmodel-results} summarizes the performance metrics for the
-three models. Overall the different aggregation functions do not
-differ dramatically in terms of the final model performance, though
-$f_{\textrm{mean}}$ appears to lead to the best performance.
+We first consider the application of the SALI prediction methodology to the Cavalli dataset. We
+developed three random forest models, using the three aggregation descriptor aggregation functions
+described in Section \ref{sec:methodology}. Fig.~\ref{fig:rfmodel-cavalli} presents the predicted
+versus observed SALI values from the three models and Table \ref{tab:rfmodel-results} summarizes the
+performance metrics for the three models. Overall the different aggregation functions do not differ
+dramatically in terms of the final model performance as root mean square error (RMSE) ranges from
+0.98 to 1.11 log units. However, $f_{\textrm{mean}}$ appears to lead to the best performance (RMSE = 0.98 and
+$R^2$ = 0.67). In contrast, the RMSE and $R^2$ values for Y-scrambled\cite{Rucker:2007aa} models
+using the different aggregation functions ranged from 1.58 to 1.61 log units and 0.003 to 0.01, respectively.
 
-While the models are certainly not the most accurate models, the
-RMSE's and $R^2$ values are reasonable. Note that the predictions on
-the low end of the SALI values are not as important as those at the
-high end - simply because low SALI values correspond to small activity
-cliffs, which are likely not very interesting. Given that observation,
-it is encouraging to observe that all three models perform relatively
-well at the higher end of the SALI spectrum. For this example, the
-three most significant activity cliffs are in fact not very
-significant activity cliffs in an absolute sense - the $T_c$ for the
-the three pairs are just 0.2, 0.30 and 0.29, though the activity
-differences were 4.93, 5.0 and 5.11 log units respectively.
+Note that the predictions on the low end of the SALI values are not as important as those at the
+high end - simply because low SALI values correspond to small activity cliffs, which are likely not
+very interesting. Given that observation, it is encouraging to observe that all three models perform
+relatively well at the higher end of the SALI spectrum. For this example, the three most significant
+activity cliffs are in fact not very significant activity cliffs in an absolute sense - the $T_c$
+for the three pairs are just 0.2, 0.30 and 0.29, though the activity differences were 4.93, 5.0 and
+5.11 log units respectively.
 
-We observed that the use of the geometric mean did not lead to models
-that were any different from those generated using the arithmetic mean
-as the aggregation function. This was also true of the other datasets
-and so the following discussion ommits the results obtained when using
-$f_{\textrm{geom}}$. Fig.~\ref{fig:rfmodel-chembl} summarizes the
-models built on the ChEMBL datasets using $f_{\textrm{diff}}$ and
-$f_{\textrm{mean}}$ as the aggregation functions. Table
-\ref{tab:rfmodel-results} reports the model statistics.  While the
-differences in model performance do not vary significantly with the
-aggregation function used, we see that for the Costanzo and Kalla
-datasets, $f_{\textrm{diff}}$ leads to a slightly better model than
-when using $f_{\textrm{mean}}$, though for the Dai dataset, the
-difference is negligable.
+We observed that the performance of models based on geometric mean aggregation function did not
+differ substantially from those built with the arithmetic mean. This was also true of the other
+datasets and so the following discussion omits the results obtained when using
+$f_{\textrm{geom}}$. Fig.~\ref{fig:rfmodel-chembl} summarizes the models built on the ChEMBL
+datasets using $f_{\textrm{diff}}$ and $f_{\textrm{mean}}$ as the aggregation functions. Table
+\ref{tab:rfmodel-results} reports the model statistics.  While the differences in model performance
+do not vary significantly with the aggregation function used, we see that for the Costanzo and Kalla
+datasets, $f_{\textrm{diff}}$ leads to a slightly better model in terms of RMSE and $R^2$ than when
+using $f_{\textrm{mean}}$.
 
-Given the relatively small differences between models based on
-different aggregation functions, we arbitrarily chose
-$f_{\textrm{mean}}$ as the aggregagtion functin for subsequent
-analyses. Figs.~\ref{fig:costanzo-pred}, \ref{fig:kalla-pred} and
-\ref{fig:dai-pred} summarize the performance of the models built using
-the reported activity data and the pairwise SALI values for the
-Costanzo, Kalla and Froimowitz datasets, respectively. As with the
-Cavalli dataset, the SALI models exhibit more variance for small SALI
-values (i.e., less significant cliffs) but appear to perform better at
-the the higher end of the SALI values. However in all three cases, the
-most significant cliffs exhibit significant variance. In comparison,
-the models developed on the reported activity values are relatively
-poor. This can be partially ascribed to the small sizes of the
-dataset.
+Given the relatively small differences between models based on different aggregation functions, we
+arbitrarily chose $f_{\textrm{mean}}$ as the aggregation function for subsequent
+analyses. Figs.~\ref{fig:costanzo-pred}, \ref{fig:kalla-pred} and \ref{fig:dai-pred} summarize the
+performance of the models built using the reported activity data and the pairwise SALI values for
+the Costanzo, Kalla and Dai datasets, respectively. As with the Cavalli dataset, the SALI models
+exhibit more variance for small SALI values (i.e., less significant cliffs) but appear to perform
+better at the higher end of the SALI values. However in all three cases, the most significant cliffs
+exhibit significant variance. In comparison, the models developed on the reported activity values
+are relatively poor. This can be partially ascribed to the small sizes of the dataset. It should be
+note that this methodology is not aimed at modeling the original, usually small, SAR dataset. It is
+well known (and evident from the current analyses) that small datasets do not lead to very robust or
+reliable machine learning models. Instead, the methodology focuses on the \emph{pairwise} dataset --
+which are usually large enough for a statistically significant model.
 
@@
 
-Clearly, modeling the pairwise SALI values leads to reasonable models
-the datasets discussed here. However, the primary goal of this
-approach is to extend the the structure-activity landscape for a set
-of molecules. In other words, can one use the model to identify
-compounds that will exhibit an activity cliff with one or more
-molecules in the dataset. Given a new molecule, one can evaluate the
-descriptors for this molecule and then evaluate the pairwise
-descriptor values with each member of the training set. Thus, given a
-new molecule, we must obtain predictions for $n$ ``pairwise objects'',
-where $n$ is the number of molecules in the training set. Given these
-$n$ predictions one can proceed to identify training set molecules
-with which the new molecule might exhibit a cliff.
+While predictive modeling of the SALI values is a first step, the primary goal of this approach is
+to extend the structure-activity landscape for a set of molecules. In other words, can one use
+the model to identify compounds that will exhibit an activity cliff with one or more molecules in
+the dataset? Given a new molecule, one can evaluate the descriptors for this molecule and then
+evaluate the pairwise descriptor values with each member of the training set. Thus, given a new
+molecule, we must obtain predictions for $n$ ``pairwise objects'', where $n$ is the number of
+molecules in the training set. Given these $n$ predictions one can proceed to identify training set
+molecules with which the new molecule might exhibit a cliff.
 
-We first performed this analysis with the Kalla dataset. Since
-predictions of new molecules involve $n$ predictions per molecule, we
-removed three molecules from the dataset and rebuilt the random
-forest model. The hold out molecules, were specifically selected since
-they displayed significant activity cliffs with various members of the
-training set. We then evaluted the pairwise descriptor values for
-these three with the training set molecules and obtained predictions
-of the log(SALI) values. Fig.~\ref{fig:tset-pset-kalla}A
-summarizes the performance of the model on the training set as well as
-for the prediction set. From a numerical point of view, the
-performance of the model on he prediction set degrades somewhat (RMSE
-of 0.51 versus 0.41 for the training set). As noted above, the
-variance of the prediction is higher at the lower scale of log(SALI)
-values. However, at the higher end of the scale there a number of
-pairs that are well predicted. Yet, it is clear that a number of
-relatively significant activity cliffs have been underestimated.
+We first performed this analysis with the Kalla dataset. Since predictions of new molecules involve
+$n$ predictions per molecule, we removed three molecules from the dataset and rebuilt the random
+forest model. The hold out molecules, were specifically selected since they displayed significant
+activity cliffs with various members of the training set. We then evaluated the pairwise descriptor
+values for these three with the training set molecules and obtained predictions of the log(SALI)
+values. Fig.~\ref{fig:tset-pset-kalla}A summarizes the performance of the model on the training set
+as well as for the prediction set. From a numerical point of view, the performance of the model on
+he prediction set degrades somewhat (RMSE of 0.51 versus 0.41 for the training set). As noted above,
+the variance of the prediction is higher at the lower scale of log(SALI) values. However, at the
+higher end of the scale there a number of pairs that are well predicted. Yet, it is clear that a
+number of relatively significant activity cliffs have been underestimated.
 
-Given that large SALI values can arise due to high structural
-similarity, even when the activity difference is small, simply
-examining log(SALI) values is not comletely informative. Specifically,
-we are interested in the predictions for the actual activity cliffs,
-where both the difference in activity and the structural similarity is
-very high. Fig.~\ref{fig:tset-pset-kalla}B visualizes this
-information. Each point coresponds to a pair of molecules, which are
-shaded by the absolute prediction residual for that pair. Thus, the
-``true'' cliffs are represented by points lying towards the bottom
-right corner of the plot. In this case we see that there are a number
-of pairs of molecules with $1 - T_C \leq 0.1$ and 100-fold or better
-difference in activity. While many of these pairs are associated with
-a high prediction residual, there are a number with low to medium
+Given that large SALI values can arise due to high structural similarity, even when the activity
+difference is small, simply examining log(SALI) values is not completely informative. Specifically,
+we are interested in the predictions for the actual activity cliffs, where both the difference in
+activity and the structural similarity is very high. Fig.~\ref{fig:tset-pset-kalla}B visualizes this
+information. Each point corresponds to a pair of molecules, which are shaded by the absolute
+prediction residual for that pair. Thus, the ``true'' cliffs are represented by points lying towards
+the bottom right corner of the plot. In this case we see that there are a number of pairs of
+molecules with $1 - T_C \leq 0.1$ and 100-fold or better difference in activity. While many of these
+pairs are associated with a high prediction residual, there are a number with low to medium
 residuals.
 
-Fig.~\ref{fig:kalla-comparison} displays the structures of the three
-hold out molecules (top row) and a selection of structures, which
-which these hold out molecules were predicted to exhibit an activity
-cliff. For each molecule, the $K_i$ value is listed along with the
-absoute residual value (log(SALI) units). For molecules
-\textbf{349288} and \textbf{349699}, the predicted cliffs are
-relatively accurate, as evidenced by the low residuals. For
-\textbf{349138} on the other hand, most of the residuals were were
-relatively high. While we have highlighted a number of cliffs,
-Fig.~\ref{fig:tset-pset-kalla}C makes it clear that the predictions
-for this molecule were, on average, worse than for the other two
+Fig.~\ref{fig:kalla-comparison} displays the structures of the three hold out molecules (top row)
+and a selection of structures, which which these hold out molecules were predicted to exhibit an
+activity cliff. For each molecule, the $K_i$ value is listed along with the absolute residual value
+(log(SALI) units). For molecules \textbf{349288} and \textbf{349699}, the predicted cliffs are
+relatively accurate, as evidenced by the low residuals. For \textbf{349138} on the other hand, most
+of the residuals were were relatively high. While we have highlighted a number of cliffs,
+Fig.~\ref{fig:tset-pset-kalla}C makes it clear that the predictions for this molecule were, on
+average, worse than for the other two molecules.
+
+Fig.~\ref{fig:tset-pset-dai} summarizes the results for the same set of analyzes described above,
+applied to the Dai dataset. As before we removed three molecules (top row of
+Fig.~\ref{fig:dai-comparison}) and rebuilt the random forest on the paired cases with the remaining
+41 molecules. We then predicted the pairwise SALI values for the three hold-out molecules with the
+remaining 41 molecules (i.e., the prediction set). A plot of the predicted versus observed log(SALI)
+values for the training and prediction sets are displayed in Fig.~\ref{fig:tset-pset-dai}A. As
+before the model performs poorly on the lesser cliffs, but improves for the more significant cliffs
+(RMSE for the training set and prediction sets were 0.37 and 0.43 log units respectively). However,
+the 2 most significant cliffs are under-estimated. However, in comparison to the Kalla dataset, the
+distribution of absolute residuals aggregated by the hold-out molecule
+(Fig.~\ref{fig:tset-pset-dai}C) are similar, with a median absolute residual of less than
+0.3 log units.  For this dataset, the number of ``real'' cliffs is relatively low as shown in
+Fig.~\ref{fig:tset-pset-dai}B. Interestingly, the compound pairs that display moderate to
+significant cliffs are relatively well predicted - in fact, the maximum residuals are observed for
+the least significant cliffs.
+
+We then considered the three hold-out molecules and some of the members of the training set with
+which they are predicted to show activity cliffs. For this dataset, there are relatively few
+``true'' activity cliffs. For example, while molecules \textbf{371259} and \textbf{371307} have a
+1000-fold difference in activity, their $T_c = 0.68$, and would not represent a truly significant
+cliff. However, the pair is well predicted with a residual of 0.04 log units. if we consider the
+subset of predictions where the $T_c \geq 0.8$ we note that the median absolute residual is 0.24 log
+units. However, the activity differences are not always significant ranging from 1.1-fold to
+13-fold. Two such examples are shown in Fig.~\ref{fig:dai-comparison}, where the only difference is
+in the position of the methyl substituent. Similar behavior is observed with the other hold-out
 molecules.
-
-Fig.~\ref{fig:tset-pset-dai} summarizes the results for the same set
-of analyzes described above, applied to the Dai dataset. As before we
-removed three molecules (top row of Fig.~\ref{fig:dai-comparison}) and
-rebuilt the random forest on the paired cases with the remaining 41
-molecules. We then predicted the pairwise SALI values for the three
-hold-out molecules with the remaining 41 molecules (i.e., the
-prediction set). A plot of the predicted versus observed log(SALI)
-values for the training and prediction sets are displayed in
-Fig.~\ref{fig:tset-pset-dai}A. As before the model performs poorly on
-the lesser cliffs, but improves for the more significant cliffs (RMSE
-for the training set and prediction sets were 0.37 and 0.43 log units
-respectively). However, the 2 most significant cliffs are
-under-estimated. However, in comparison to the Kalla dataset, the
-distribution of absolute residuals aggregated by the hold-out molecule
-(Fig.~\ref{fig:tset-pset-dai}C) are reatively similar, with a median
-absolute residual of less than 0.3 log units.  For this dataset, the
-number of ``real'' cliffs is relatively low as shown in
-Fig.~\ref{fig:tset-pset-dai}B. Interestingly, the compound pairs that
-display moderate to significant cliffs are relatively well predicted -
-in fact, the maximum residuals are observed for the least significant
-cliffs.
-
-We then considered the three hold-out molecules and some of the members
-of the training set with which they are predicted to show activity
-cliffs. For this dataset, there are relatively few ``true'' activity
-cliffs. For example, while molecules
-\textbf{371259} and \textbf{371307} have a 1000-fold difference in
-activity, their $T_c = 0.68$, and would not represent a truly
-significant cliff. However, the pair is well poredicted with a
-residual of 0.04 log units. if we consider the subset of predictions
-where the $T_c \geq 0.8$ we note that the median absolute residual is
-0.24 log units. However, the activity differences are not always
-significant ranging from 1.1-fold to 13-fold. Two such examples are shown
-in Fig.~\ref{fig:dai-comparison}, where the only difference is in the
-position of the methyl substituent. Similar behavior is observed with
-the other hold-out molecules.
 
@@
 
-The results described above suggest that predicting pairwise SALI
-values is a useful way to identify whether a new molecule will
-form an activity cliff with one or more members of the training
-set. As with all QSAR models, one is limited by the nature of the
-underlying data, both in terms of accuracy as well as
-applicability. Thus while we have selected a few ChEMBL assays for the
-purposes of highlighting this approach, we have found that for some
-assays, the model built on pairwise data performs worse than the model
-built on the original data. Given that most of the datasets are small
-in size (less than 40 molecules), the poor performance of the initial
-model is not always surprising. However, in cases where the pairwise
-model does not exhibit improved performance, we believe that
-predictive modeling of activity cliffs in the original dataset will
-not be fruitful.
+The results described above suggest that predicting pairwise SALI values is a useful way to identify
+whether a new molecule will form an activity cliff with one or more members of the training set. As
+with all QSAR models, one is limited by the nature of the underlying data, both in terms of accuracy
+as well as applicability. Thus while we have selected a few ChEMBL assays for the purposes of
+highlighting this approach, we have found that for some assays, the model built on pairwise data
+performs worse than the model built on the original data. Given that most of the datasets are small
+in size (less than 40 molecules), the poor performance of the initial model is not always
+surprising. However, in cases where the pairwise model does not exhibit improved performance, we
+believe that predictive modeling of activity cliffs in the original dataset will not be fruitful.
 
-Given that many of the models show relatively poor statistics on the
-pairwise data, it is important to note that much of this is due to the
-large variance in the predictions for the less significant cliffs. One
-possible reason for the increase in variance for these portions of the
-dataset is the fact that the distribution of the log(SALI) values
-tends to be skewed to the right (Fig.~\ref{fig:logsalidist}). In
-general we see that the smaller SALI values constitute a relatively
+Given that many of the models show relatively poor statistics on the pairwise data, it is important
+to note that much of this is due to the large variance in the predictions for the less significant
+cliffs. One possible reason for the increase in variance for these portions of the dataset is the
+fact that the distribution of the log(SALI) values tends to be skewed to the right
+(Fig.~\ref{fig:logsalidist}). In general we see that the smaller SALI values constitute a relatively
 small portion of the dataset.
 
-Another factor affecting the models built on pairwise data is the
-descriptors that are employed. While the choice of initial descriptors
-is always open to debate, we partially avoid a biased selecion by
-employing a random forest model to perform implicit feature
-selection. However, given an initial set of descriptors we observe
-that the models are robust to the different aggregation functions to
-derive a pairwise descriptor representation from the descriptors for
-the individual compounds. This is a useful observation given that
-there appears to be no rule to select one aggregation function over
-another \emph{a priori}. While random forests do allow us to avoid
-feature selection, it is clear that other models, such as neural
-networks or support vector machines, could lead to more predictive
-solutions. At this stage, our aim is to highight the modeling
-procedure and we believe that the random forest allows us to highight
-the utility of this approach without sacrificing too much by way of
-predictive performance.
+Another factor affecting the models built on pairwise data is the descriptors that are
+employed. While the choice of initial descriptors is always open to debate, we partially avoid a
+biased selection by employing a random forest model to perform implicit feature selection. However,
+given an initial set of descriptors we observe that the models are robust to the different
+aggregation functions to derive a pairwise descriptor representation from the descriptors for the
+individual compounds. This is a useful observation given that there appears to be no rule to select
+one aggregation function over another \emph{a priori}. While random forests do allow us to avoid
+feature selection, it is clear that other models, such as neural networks or support vector
+machines, could lead to more predictive solutions. At this stage, our aim is to highlight the
+modeling procedure and we believe that the random forest allows us to highlight the utility of this
+approach without sacrificing too much by way of predictive performance.
 
-It is important to note that while the modeling approach here focuses
-on predicting SALI values rather than actual activities, one can
-obtain an idea of the latter based on the (known) activity of the
-training set molecules. Thus for a new molecule, predicted to show a
-cliff with a member of the training set, we can determine whether this
-molecule might be the potent member of the cliff, by noting whether
-the corresponding training set molecule is less or more potent. Given
-that one is usually interested in activity cliffs because of the
-possibility of increasing potency, this observation implies that for a
-collection of new molecules, one can prioritize them as being
-predicted to be more or less potent. From this point of view, the
-approach described here is more suited for prioritization, rather than
-prediction of absolute values.
+It is important to note that while the modeling approach here focuses on predicting SALI values
+rather than actual activities, one can obtain an idea of the latter based on the (known) activity of
+the training set molecules. Thus for a new molecule, predicted to show a cliff with a member of the
+training set, we can determine whether this molecule might be the potent member of the cliff, by
+noting whether the corresponding training set molecule is less or more potent. Given that one is
+usually interested in activity cliffs because of the possibility of increasing potency, this
+observation implies that for a collection of new molecules, one can prioritize them as being
+predicted to be more or less potent. From this point of view, the approach described here is more
+suited for prioritization, rather than prediction of absolute values.
 
@@
 
-We have presented an approach to extending a structure-activity
-landscape in an indirect fashion, by predicting the propensity of a
-new molecule to exhibit an activity cliff with one or more molecules
-in a pre-exisiting SAR dataset. The method is based on building
-a model on pairwise SALI values (dependent variable) and pairwise
-aggregated descriptor values (independent variables). For a new
-molecule, we obtain $n$ pairwise SALI predictions (since we must
-estimate the SALI value for each training set molecule and the test
-molecule). The predicted SALI values can then be used to judge whether
-the new molecule will exhibit a cliff and whether such a cliff is in
-the desired direction (i.e., improving potency versus worsening
-potency). To test this strategy we have developed random forest modls
-to predict SALI values for several ChEMBL datasets. While the model
-performance statistics are not stellar, we observe that this is
-primarily due to high prediction variance at the lower range of SALI
-values. In contrast, the more significant cliffs are relatively well
-predicted, though, unsurprisingly, the most significant cliffs are not
-always well predicted.
+We have presented an approach to extending a structure-activity landscape in an indirect fashion, by
+predicting the propensity of a new molecule to exhibit an activity cliff with one or more molecules
+in a pre-existing SAR dataset. The method is based on building a model on pairwise SALI values
+(dependent variable) and pairwise aggregated descriptor values (independent variables). For a new
+molecule, we obtain $n$ pairwise SALI predictions (since we must estimate the SALI value for each
+training set molecule and the test molecule). The predicted SALI values can then be used to judge
+whether the new molecule will exhibit a cliff and whether such a cliff is in the desired direction
+(i.e., improving potency versus worsening potency). To test this strategy we have developed random
+forest models to predict SALI values for several ChEMBL datasets. While the model performance
+statistics are not stellar, we observe that this is primarily due to high prediction variance at the
+lower range of SALI values. In contrast, the more significant cliffs are relatively well predicted,
+though, unsurprisingly, the most significant cliffs are not always well predicted.
 
-In summary, this approach extends the activity cliff concept, from
-that of a retrospective analysis tool to a prospective tool that could
-be used to guide synthetic campaigns in their goals of improved
-potency.
+In summary, this approach extends the activity cliff concept, from that of a retrospective analysis
+tool to a prospective tool that could be used to guide synthetic campaigns in their goals of
+improved potency.
 

### 2012-07-10 rebuttal.tex
Updated rebuttal a bit more

@@
 The text has been thoroughly proof read to remove typos and incorrect grammar. References have been
-added and methods have been described more clearly.
+added and methods have been described more clearly. It's not clear which descriptions were
+incomplete. However, the methodology section now reads more coherently, and some aspects are now
+clearer based on other reviewer comments.
 
@@
   statistical evaluations and corresponding metrics for models and their predictions.}
+
+This was an oversight on our part. The use of qualitative terms has been removed and results and
+comparisons are quantitatively characterized. For example,  the random forest models built to
+predict SALI values are now compared to Y-scrambled models to ensure that the the results are not
+purely due to chance.
 
@@
 
+Yes. For a given dataset, with measured activity values, one can indeed evaluate the SALI
+values. The proposed methodology takes these ``measured'' SALI values for a training set and then
+predicts the SALI values for a new molecule with each member of the training set. As with
+traditional QSAR models, a complete validation would require one to measure the activity of the new
+molecule and evaluate the actual SALI values. In absence of experimental activities for the test
+set, we resort to a training/test split.
+
 \textbf{5.  There is no discussion/direction of how to use the methodology in new applications.}
+
 \end{document}

### 2012-07-18 rebuttal.tex
Added more rebuttal

@@
 
-Recommendation: Publish after major revisions noted.
+\textbf{(1) Please, note that there are numerous typographical errors
+throughout the manuscript (which is quite disturbing). 
+\vskip 1em
+(2) Chosen methods and descriptors should at least be referenced. For
+example, no references are provided for BCI and CDK.}
 
-Comments:
-The prediction of activity cliffs is a nice idea that would further extend activity landscape modeling.
-This is attempted by predicting SALI values for compound pairs on the basis of training data.
-Unfortunately, there are a number of technical issues to note and, more importantly, methodological concerns that would need to be carefully addressed before publication of this study might be considered.
+The manuscript has been thoroughly proof read and references included
+where appropriate
 
-First, a few general comments: 
-(1) Please, note that there are numerous typographical errors throughout the manuscript (which is quite disturbing). Unfortunately, the presentation of the work is sloppy at different levels.       
-(2) Chosen methods and descriptors should at least be referenced. For example, no references are provided for BCI and CDK.
+\textbf{(3) The derivation of RF models is essentially not described
+  (except of making reference to default parameters settings in R). It
+  remains largely unclear how these models were developed. }
 
-Second, there are shortcomings at the technical level:
-(3) The derivation of RF models is essentially not described (except of making reference to default parameters settings in R). It remains largely unclear how these models were developed. 
-(4) The selection of training and test sets should be clearly specified in the Methods section.
-For example, how were the RF models trained and evaluated for which predictions of pair-wise SALI values are reported in Figure 2?
-It is hoped that the reported SALI values were not predicted for the training data.
-(5) No support is provided for the assumption that the RF models “are relatively robust even for datasets of small size”.
-In fact, the author later on states several times that limited model quality might be due to the use of small compound sets.
-Appropriate validation will be essential here.
-(6) There is only limited statistical assessment/validation of the models.
-Statements made in the text that R2 values are “reasonable” and the models perform “relatively well” are difficult to reconcile.
-Rather, Table 2 indicates that the R2 values were rather poor in several instances.
-(7) It appears that there are systematic prediction errors for pairs.
-At low value ranges (that are not relevant for activity cliff assessment), SALI values are consistently over-predicted.
-By contrast, at high value ranges (that are relevant), SALI values are consistently under-predicted, in part significantly.
-This is not investigated, although the apparent errors at high value ranges are the perhaps most critical features of the reported RF models.
-(8) In the section “Extending a landscape”, predictions are finally reported for only three (!) molecules taken out of each data set.
-Carefully put, the predictions are heterogeneous.
-On the basis of the results reported for this little bit of evaluation, one would be hard pressed to make a case for the ability of the approach to predict activity cliffs with any certainty.
+The text has been updated to be more specific on the development of
+the random forest model.
 
-Third, there are major scientific concerns:
-(9) For the prediction of activity cliffs, SALI values must be calculated for compound pairs.
-This is the central idea of the approach.
-Accordingly, fingerprint descriptors and potency values must be combined in some ways.
-(9.1.) It is nowhere stated how activity values are treated for compounds forming a pair.
-Are activity differences calculated at the pair level?
-If so, how are activity differences compared for the calculation of pair-based SALI scores - as a difference of differences?
-It should be noted predicted potency values in Figures 4-6 are essentially all over the place.
-(9.2.) To obtain a fingerprint descriptor for a pair, the author aggregates fingerprints of individual molecules, finally, by simple averaging.
-For comparisons of pairs, Tanimoto similarity is then calculated for averaged fingerprints, which presents an artificial assessment of pair similarity (one can easily come up with a few hypothetical compound similarity relationships that make this assessment questionable at best).
-As a consequence, activity cliffs are considered for which Tanimoto similarity calculations yield values between 0.2 and 0.3!
-Thus, on the basis of these similarity calculations, one could never decide which level of similarity might be cliff relevant.
-In this respect, the descriptions of activity cliff by the author are quite telling – he speaks of “predicted” versus “true” activity cliffs, of “relatively accurate” activity cliffs, and of “most significant activity cliffs that are in fact not very significant activity cliffs in an absolute sense”, and so on.
+(4) The
+selection of training and test sets should be clearly specified in the
+Methods section. For example, how were the RF models trained and
+evaluated for which predictions of pair-wise SALI values are reported
+in Figure 2? It is hoped that the reported SALI values were not
+predicted for the training data.
+
+(5) No support is provided for the
+assumption that the RF models “are relatively robust even for datasets
+of small size”. In fact, the author later on states several times that
+limited model quality might be due to the use of small compound
+sets. Appropriate validation will be essential here.
+
+(6) There is
+only limited statistical assessment/validation of the
+models. Statements made in the text that R2 values are “reasonable”
+and the models perform “relatively well” are difficult to
+reconcile. Rather, Table 2 indicates that the R2 values were rather
+poor in several instances.
+
+ (7) It appears that there are systematic
+prediction errors for pairs. At low value ranges (that are not
+relevant for activity cliff assessment), SALI values are consistently
+over-predicted. By contrast, at high value ranges (that are relevant),
+SALI values are consistently under-predicted, in part
+significantly. This is not investigated, although the apparent errors
+at high value ranges are the perhaps most critical features of the
+reported RF models.
+
+(8) In the section “Extending a landscape”,
+predictions are finally reported for only three (!) molecules taken
+out of each data set. Carefully put, the predictions are
+heterogeneous. On the basis of the results reported for this little
+bit of evaluation, one would be hard pressed to make a case for the
+ability of the approach to predict activity cliffs with any certainty.
+
+\textbf{(9) For the prediction of activity cliffs, SALI values must be
+calculated for compound pairs. This is the central idea of the
+approach. Accordingly, fingerprint descriptors and potency values must
+be combined in some ways. 
+\vskip 1em
+(9.1.) It is nowhere stated how activity values are treated for
+compounds forming a pair. Are activity differences calculated at the
+pair level? If so, how are activity differences compared for the
+calculation of pair-based SALI scores - as a difference of
+differences? It should be noted predicted potency values in Figures
+4-6 are essentially all over the place.
+}
+The entire premise of the proposed method is that one evaluates SALI
+values for a dataset that has observed activities. We state
+specifically (Equation 1) how the activities are combined when
+molecules are considered pairwise.
+
+It's not clear what the reviewer means by \emph{``\ldots how are
+  activity differences compared for the calculation of pair-based SALI
+  scores \ldots''}. The SALI value itself represents a pair of
+molecules and in this calculation, we consider the absolute activity
+difference. We do not consider pairs of pairs
+
+\textbf{(9.2.) To obtain a fingerprint descriptor for a pair, the
+  author aggregates fingerprints of individual molecules, finally, by
+  simple averaging}
+
+In fact, we do not generate pairwise fingerprint
+descriptors. Fingerprints are only used to evaluate the similarity
+term in the calculation of the SALI value.
+
+\textbf{ For comparisons of pairs, Tanimoto similarity is then
+  calculated for averaged fingerprints, which presents an artificial
+  assessment of pair similarity (one can easily come up with a few
+  hypothetical compound similarity relationships that make this
+  assessment questionable at best). As a consequence, activity cliffs
+  are considered for which Tanimoto similarity calculations yield
+  values between 0.2 and 0.3!}
+
+I believe that these comments are based on the discussion in Section
+3.1 and it appears that this section of the paper was not clear.
+
+Firstly, we have never mentioned the averaging of fingerprint
+descriptors. Instead, we generate a descriptor vector for a pair of
+molecules by averaging the (topological) descriptor vectors of the
+individual members of that pair.
+
+The goal of Section 3.1 is to indicate that even though the Y variable
+and the X variables encode structural information (the former
+indirectly and the latter directly), there is little correlation
+between the individual (aggregated, topological) descriptors and the Y
+variable (where sructural information is included via
+fingerprints). To quantify this, we evaluate the Pearson correlation
+between each (aggregated, topological) descriptor and the Y variable
+and show that the maximum value is very low ($R^2 < 0.15$).
+
+This discussion does not involve Tanimoto similarities at all
+
+ Thus, on the basis of these similarity calculations, one could never
+ decide which level of similarity might be cliff relevant.
+
+\textbf{ In this
+ respect, the descriptions of activity cliff by the author are quite
+ telling – he speaks of “predicted” versus “true” activity cliffs, of
+ “relatively accurate” activity cliffs, and of “most significant
+ activity cliffs that are in fact not very significant activity cliffs
+ in an absolute sense”, and so on. }
+
+I agree that the wording of this discussion is rather sloppy. The text
+has been updated to be more quantitative.
+
+Yet at the same time, the concept of an activity cliff is somewhat
+subjective. In many cases, it is akin to saying \emph{``I'll know it's
+  an activity cliff when I see it''}. For example, a pair of molecules
+exhibiting a $T_c = 0.9$ and 1000x difference in acivity is obviously
+an activity clif. Would a pair with $T_C = 0.85$ and 50x difference in
+activity not be an activity cliff? From this point of view, I do not
+think it unreasonable to refer somewhat qualitatively to the
+``cliffness'' of an activity cliff.
+
 Furthermore, there might be many instances of activity and fingerprint similarity relationships between pairs of compounds that might yield high pair-based SALI scores dominated by either the activity or similarity component, without forming a “true” activity cliff.
@@
 
-Yes. For a given dataset, with measured activity values, one can indeed evaluate the SALI
-values. The proposed methodology takes these ``measured'' SALI values for a training set and then
-predicts the SALI values for a new molecule with each member of the training set. As with
-traditional QSAR models, a complete validation would require one to measure the activity of the new
-molecule and evaluate the actual SALI values. In absence of experimental activities for the test
-set, we resort to a training/test split.
+Yes. For a given dataset, with measured activity values, one can
+indeed evaluate the SALI values. The proposed methodology takes these
+``measured'' SALI values for a training set and then predicts the SALI
+values for a new molecule with each member of the training set. As
+with traditional QSAR models, a complete validation would require one
+to measure the activity of the new molecule and evaluate the actual
+SALI values. In absence of experimental activities for the test set,
+we resort to a training/test split, which is an accepted approach in
+QSAR modeling.
 
@@
 
+Section 4.1 in the original manuscript specifically addresses how one
+might use the proposed methodology
+
 \end{document}

### 2012-07-18 paper.tex
Added more rebuttal

@@
 linear regression) makes no such assumptions. We employed the implementation of random forest from
-the \texttt{randomForest} package, in R 2.11.0\cite{r}, using the default settings.
+the \texttt{randomForest} package, in R 2.11.0\cite{r}. The models were built using 500 trees, and
+randomly sampled the descriptor pool for a given dataset, using $\sqrt(N)$ descriptors at a time,
+where $N$ is the number of descriptors in the pool. By default, the method builds individual trees
+using 63\% of the dataset and tested on the remainder (the so called out-of-bag data). The final
+predicted values for the entire dataset are obtained by averaging the predictions for each
+observation from from all 500 trees. We also investigated the traditional approach of a
+training/test set split
 
@@
 correlated with others (using an $R^2$ cutoff of 0.8). The sizes of the final descriptor pool for
-each dataset are summarized in Table \ref{tab:data}.
+each dataset are summarized in Table \ref{tab:data}. XXX
 
@@
 While predictive modeling of the SALI values is a first step, the primary goal of this approach is
-to extend the structure-activity landscape for a set of molecules. In other words, can one use
-the model to identify compounds that will exhibit an activity cliff with one or more molecules in
-the dataset? Given a new molecule, one can evaluate the descriptors for this molecule and then
-evaluate the pairwise descriptor values with each member of the training set. Thus, given a new
-molecule, we must obtain predictions for $n$ ``pairwise objects'', where $n$ is the number of
-molecules in the training set. Given these $n$ predictions one can proceed to identify training set
-molecules with which the new molecule might exhibit a cliff.
+to extend the structure-activity landscape for a set of molecules. In other words, can one use the
+model to identify compounds that will exhibit an activity cliff with one or more molecules in the
+dataset? Given a new molecule, one can evaluate the descriptors for this molecule and then evaluate
+the pairwise descriptor values with each member of the training set. Thus, given a new molecule, we
+must obtain predictions for $n$ ``pairwise objects'', where $n$ is the number of molecules in the
+training set. Given these $n$ predictions one can proceed to identify training set molecules with
+which the new molecule might exhibit a cliff.
 
@@
 with all QSAR models, one is limited by the nature of the underlying data, both in terms of accuracy
-as well as applicability. Thus while we have selected a few ChEMBL assays for the purposes of
+as well as applicability. We have selected a few ChEMBL assays for the purposes of
 highlighting this approach, we have found that for some assays, the model built on pairwise data
@@
 approach without sacrificing too much by way of predictive performance.
-
-It is important to note that while the modeling approach here focuses on predicting SALI values
-rather than actual activities, one can obtain an idea of the latter based on the (known) activity of
-the training set molecules. Thus for a new molecule, predicted to show a cliff with a member of the
-training set, we can determine whether this molecule might be the potent member of the cliff, by
-noting whether the corresponding training set molecule is less or more potent. Given that one is
-usually interested in activity cliffs because of the possibility of increasing potency, this
-observation implies that for a collection of new molecules, one can prioritize them as being
-predicted to be more or less potent. From this point of view, the approach described here is more
-suited for prioritization, rather than prediction of absolute values.
 

### 2012-07-19 rebuttal.tex
Added more rebuttal

@@
 
-\DeclareGraphicsExtensions{.eps,.png}
-\onehalfspacing
+\DeclareGraphicsExtensions{.eps,.png} \onehalfspacing
 
@@
 
-\textbf{(1) Please, note that there are numerous typographical errors
-throughout the manuscript (which is quite disturbing). 
-\vskip 1em
-(2) Chosen methods and descriptors should at least be referenced. For
-example, no references are provided for BCI and CDK.}
+\textbf{(1) Please, note that there are numerous typographical errors throughout the manuscript
+  (which is quite disturbing).  \vskip 1em (2) Chosen methods and descriptors should at least be
+  referenced. For example, no references are provided for BCI and CDK.}
 
-The manuscript has been thoroughly proof read and references included
-where appropriate
+The manuscript has been thoroughly proof read and references included where appropriate
 
-\textbf{(3) The derivation of RF models is essentially not described
-  (except of making reference to default parameters settings in R). It
-  remains largely unclear how these models were developed. }
+\textbf{(3) The derivation of RF models is essentially not described (except of making reference to
+  default parameters settings in R). It remains largely unclear how these models were developed. }
 
-The text has been updated to be more specific on the development of
-the random forest model.
+The text has been updated to be more specific on the development of the random forest model.
 
-(4) The
-selection of training and test sets should be clearly specified in the
-Methods section. For example, how were the RF models trained and
-evaluated for which predictions of pair-wise SALI values are reported
-in Figure 2? It is hoped that the reported SALI values were not
-predicted for the training data.
+(4) The selection of training and test sets should be clearly specified in the Methods section. For
+example, how were the RF models trained and evaluated for which predictions of pair-wise SALI values
+are reported in Figure 2? It is hoped that the reported SALI values were not predicted for the
+training data.
 
-(5) No support is provided for the
-assumption that the RF models “are relatively robust even for datasets
-of small size”. In fact, the author later on states several times that
-limited model quality might be due to the use of small compound
-sets. Appropriate validation will be essential here.
+(5) No support is provided for the assumption that the RF models “are relatively robust even for
+datasets of small size”. In fact, the author later on states several times that limited model
+quality might be due to the use of small compound sets. Appropriate validation will be essential
+here.
 
-(6) There is
-only limited statistical assessment/validation of the
-models. Statements made in the text that R2 values are “reasonable”
-and the models perform “relatively well” are difficult to
-reconcile. Rather, Table 2 indicates that the R2 values were rather
-poor in several instances.
+(6) There is only limited statistical assessment/validation of the models. Statements made in the
+text that R2 values are “reasonable” and the models perform “relatively well” are difficult to
+reconcile. Rather, Table 2 indicates that the R2 values were rather poor in several instances.
 
- (7) It appears that there are systematic
-prediction errors for pairs. At low value ranges (that are not
-relevant for activity cliff assessment), SALI values are consistently
-over-predicted. By contrast, at high value ranges (that are relevant),
-SALI values are consistently under-predicted, in part
-significantly. This is not investigated, although the apparent errors
-at high value ranges are the perhaps most critical features of the
-reported RF models.
+(7) It appears that there are systematic prediction errors for pairs. At low value ranges (that are
+not relevant for activity cliff assessment), SALI values are consistently over-predicted. By
+contrast, at high value ranges (that are relevant), SALI values are consistently under-predicted, in
+part significantly. This is not investigated, although the apparent errors at high value ranges are
+the perhaps most critical features of the reported RF models.
 
-(8) In the section “Extending a landscape”,
-predictions are finally reported for only three (!) molecules taken
-out of each data set. Carefully put, the predictions are
-heterogeneous. On the basis of the results reported for this little
-bit of evaluation, one would be hard pressed to make a case for the
-ability of the approach to predict activity cliffs with any certainty.
+(8) In the section “Extending a landscape”, predictions are finally reported for only three (!)
+molecules taken out of each data set. Carefully put, the predictions are heterogeneous. On the basis
+of the results reported for this little bit of evaluation, one would be hard pressed to make a case
+for the ability of the approach to predict activity cliffs with any certainty.
 
-\textbf{(9) For the prediction of activity cliffs, SALI values must be
-calculated for compound pairs. This is the central idea of the
-approach. Accordingly, fingerprint descriptors and potency values must
-be combined in some ways. 
-\vskip 1em
-(9.1.) It is nowhere stated how activity values are treated for
-compounds forming a pair. Are activity differences calculated at the
-pair level? If so, how are activity differences compared for the
-calculation of pair-based SALI scores - as a difference of
-differences? It should be noted predicted potency values in Figures
-4-6 are essentially all over the place.
-}
-The entire premise of the proposed method is that one evaluates SALI
-values for a dataset that has observed activities. We state
-specifically (Equation 1) how the activities are combined when
+\textbf{(9) For the prediction of activity cliffs, SALI values must be calculated for compound
+  pairs. This is the central idea of the approach. Accordingly, fingerprint descriptors and potency
+  values must be combined in some ways.  \vskip 1em (9.1.) It is nowhere stated how activity values
+  are treated for compounds forming a pair. Are activity differences calculated at the pair level?
+  If so, how are activity differences compared for the calculation of pair-based SALI scores - as a
+  difference of differences? It should be noted predicted potency values in Figures 4-6 are
+  essentially all over the place.}
+
+The entire premise of the proposed method is that one evaluates SALI values for a dataset that has
+observed activities. We state specifically (Equation 1) how the activities are combined when
 molecules are considered pairwise.
 
-It's not clear what the reviewer means by \emph{``\ldots how are
-  activity differences compared for the calculation of pair-based SALI
-  scores \ldots''}. The SALI value itself represents a pair of
-molecules and in this calculation, we consider the absolute activity
-difference. We do not consider pairs of pairs
+It's not clear what the reviewer means by \emph{``\ldots how are activity differences compared for
+  the calculation of pair-based SALI scores \ldots''}. The SALI value itself represents a pair of
+molecules and in this calculation, we consider the absolute activity difference. We do not consider
+pairs of pairs
 
-\textbf{(9.2.) To obtain a fingerprint descriptor for a pair, the
-  author aggregates fingerprints of individual molecules, finally, by
-  simple averaging}
+\textbf{(9.2.) To obtain a fingerprint descriptor for a pair, the author aggregates fingerprints of
+  individual molecules, finally, by simple averaging}
 
-In fact, we do not generate pairwise fingerprint
-descriptors. Fingerprints are only used to evaluate the similarity
-term in the calculation of the SALI value.
+In fact, we do not generate pairwise fingerprint descriptors. Fingerprints are only used to evaluate
+the similarity term in the calculation of the SALI value.
 
-\textbf{ For comparisons of pairs, Tanimoto similarity is then
-  calculated for averaged fingerprints, which presents an artificial
-  assessment of pair similarity (one can easily come up with a few
-  hypothetical compound similarity relationships that make this
-  assessment questionable at best). As a consequence, activity cliffs
-  are considered for which Tanimoto similarity calculations yield
+\textbf{For comparisons of pairs, Tanimoto similarity is then calculated for averaged fingerprints,
+  which presents an artificial assessment of pair similarity (one can easily come up with a few
+  hypothetical compound similarity relationships that make this assessment questionable at best). As
+  a consequence, activity cliffs are considered for which Tanimoto similarity calculations yield
   values between 0.2 and 0.3!}
 
-I believe that these comments are based on the discussion in Section
-3.1 and it appears that this section of the paper was not clear.
+I believe that these comments are based on the discussion in Section 3.1 and it appears that this
+section of the paper was not clear.
 
-Firstly, we have never mentioned the averaging of fingerprint
-descriptors. Instead, we generate a descriptor vector for a pair of
-molecules by averaging the (topological) descriptor vectors of the
+Firstly, we have never mentioned the averaging of fingerprint descriptors. Rather, the averaging is
+applied to the topological descritpors (i.e., the X variables). Specifically, we generate a
+descriptor vector for a pair of molecules by averaging the (topological) descriptor vectors of the
 individual members of that pair.
 
-The goal of Section 3.1 is to indicate that even though the Y variable
-and the X variables encode structural information (the former
-indirectly and the latter directly), there is little correlation
-between the individual (aggregated, topological) descriptors and the Y
-variable (where sructural information is included via
-fingerprints). To quantify this, we evaluate the Pearson correlation
-between each (aggregated, topological) descriptor and the Y variable
-and show that the maximum value is very low ($R^2 < 0.15$).
+The goal of Section 3.1 is to indicate that even though the Y variable and the X variables encode
+structural information (the former indirectly and the latter directly), there is little correlation
+between the individual (aggregated, topological) descriptors and the Y variable (where sructural
+information is included via fingerprints). To quantify this, we evaluate the Pearson correlation
+between each (aggregated, topological) descriptor and the Y variable and show that the maximum value
+is very low ($R^2 < 0.15$).
 
-This discussion does not involve Tanimoto similarities at all
+This discussion does not involve fingerprint-based Tanimoto similarities at all. In fact, if one
+does indeed consider the pairwise Tanimoto values of the molecules within a dataset, we do in fact
+observe an appreciable number of pairs wth $T_c > 0.8$.
 
- Thus, on the basis of these similarity calculations, one could never
- decide which level of similarity might be cliff relevant.
+\textbf{Thus, on the basis of these similarity calculations, one could never decide which level of
+  similarity might be cliff relevant. In this respect, the descriptions of activity cliff by the
+  author are quite telling – he speaks of “predicted” versus “true” activity cliffs, of “relatively
+  accurate” activity cliffs, and of “most significant activity cliffs that are in fact not very
+  significant activity cliffs in an absolute sense”, and so on. }
 
-\textbf{ In this
- respect, the descriptions of activity cliff by the author are quite
- telling – he speaks of “predicted” versus “true” activity cliffs, of
- “relatively accurate” activity cliffs, and of “most significant
- activity cliffs that are in fact not very significant activity cliffs
- in an absolute sense”, and so on. }
+I agree that the wording of this discussion is rather sloppy. The text has been updated to be more
+quantitative.
 
-I agree that the wording of this discussion is rather sloppy. The text
-has been updated to be more quantitative.
-
-Yet at the same time, the concept of an activity cliff is somewhat
-subjective. In many cases, it is akin to saying \emph{``I'll know it's
-  an activity cliff when I see it''}. For example, a pair of molecules
-exhibiting a $T_c = 0.9$ and 1000x difference in acivity is obviously
-an activity clif. Would a pair with $T_C = 0.85$ and 50x difference in
-activity not be an activity cliff? From this point of view, I do not
-think it unreasonable to refer somewhat qualitatively to the
+Yet at the same time, the concept of an activity cliff is somewhat subjective. In many cases, it is
+akin to saying \emph{``I'll know it's an activity cliff when I see it''}. For example, a pair of
+molecules exhibiting a $T_c = 0.9$ and 1000x difference in acivity is obviously an activity
+clif. Would a pair with $T_C = 0.85$ and 50x difference in activity not be an activity cliff? From
+this point of view, I do not think it unreasonable to refer somewhat qualitatively to the
 ``cliffness'' of an activity cliff.
 
-Furthermore, there might be many instances of activity and fingerprint similarity relationships between pairs of compounds that might yield high pair-based SALI scores dominated by either the activity or similarity component, without forming a “true” activity cliff.
-The conclusion of this reviewer is that it is currently not possible to capture activity cliffs with any degree of certainty on the basis of the pair calculations, as reported.
-(10) It also remains largely obscure how an activity landscape should be extended following this approach by compounds for which no activity information is available.
-Should the activity first be predicted by an RF model and then used for pair-based SALI score calculations?
-One would hope not.
+\textbf{Furthermore, there might be many instances of activity and fingerprint similarity
+  relationships between pairs of compounds that might yield high pair-based SALI scores dominated by
+  either the activity or similarity component, without forming a “true” activity cliff. The
+  conclusion of this reviewer is that it is currently not possible to capture activity cliffs with
+  any degree of certainty on the basis of the pair calculations, as reported.}
 
-In summary, there will be quite a bit of work required to get this manuscript in shape for publication, both in terms of its presentation and technical content.
-However, as long as points (9) and (10) are not thoroughly addressed scientifically, this work should not be published.
+\textbf{(10) It also remains largely obscure how an activity landscape should be extended following
+  this approach by compounds for which no activity information is available. Should the activity
+  first be predicted by an RF model and then used for pair-based SALI score calculations?  One would
+  hope not.}
 
-To give the author the benefit of the doubt -and credit for the underlying idea- this reviewer calls for “Major Revisions”, although the study might as well be considered premature at this point.
+The initial description was actually misleading. The prediction of SALI values permits us to first
+extend the SALI matrix. As correctly noted by the reviewer, this does not lead to an extension of
+the structure-activity landscape. In the proposed methodology, the SALI value for a new molecule is
+always predicted \emph{with respect to} another molecule whose activity is known. Thus, given a
+predicted SALI value, one can rearrange Equation 1 to obtain a ``predicted'' activity value for the
+test molecule. This value can be used to extend the original landscape. The text has been updated to
+discuss this correctly XXX
 
-Additional Questions:
-Please rate the quality of the science reported in this paper (10 - High Quality/ 1 - Low Quality): 4
-
-Please rate the overall importance of this paper to the field of chemical information or modeling (10 - High Importance / 1 - Low Importance): 6
+Of course, if one predicts a SALI value for a pair of molecules, in which both members have an
+unknown actvity, one cannot extend the landscape.
 
@@
 This was an oversight on our part. The use of qualitative terms has been removed and results and
-comparisons are quantitatively characterized. For example,  the random forest models built to
-predict SALI values are now compared to Y-scrambled models to ensure that the the results are not
-purely due to chance.
+comparisons are quantitatively characterized. For example, the random forest models built to predict
+SALI values are now compared to Y-scrambled models to ensure that the the results are not purely due
+to chance.
 
@@
 
-Yes. For a given dataset, with measured activity values, one can
-indeed evaluate the SALI values. The proposed methodology takes these
-``measured'' SALI values for a training set and then predicts the SALI
-values for a new molecule with each member of the training set. As
-with traditional QSAR models, a complete validation would require one
-to measure the activity of the new molecule and evaluate the actual
-SALI values. In absence of experimental activities for the test set,
-we resort to a training/test split, which is an accepted approach in
-QSAR modeling.
+Yes. For a given dataset, with measured activity values, one can indeed evaluate the SALI
+values. The proposed methodology takes these ``measured'' SALI values for a training set and then
+predicts the SALI values for a new molecule with each member of the training set. As with
+traditional QSAR models, a complete validation would require one to measure the activity of the new
+molecule and evaluate the actual SALI values. In absence of experimental activities for the test
+set, we resort to a training/test split, which is an accepted approach in QSAR modeling.
 
@@
 
-Section 4.1 in the original manuscript specifically addresses how one
-might use the proposed methodology
+Section 4.1 in the original manuscript specifically addresses how one might use the proposed
+methodology
 

### 2012-07-20 rebuttal.tex
removed discussion of extending landscapes. Updated rebuttal

@@
 
-The initial description was actually misleading. The prediction of SALI values permits us to first
-extend the SALI matrix. As correctly noted by the reviewer, this does not lead to an extension of
-the structure-activity landscape. In the proposed methodology, the SALI value for a new molecule is
-always predicted \emph{with respect to} another molecule whose activity is known. Thus, given a
-predicted SALI value, one can rearrange Equation 1 to obtain a ``predicted'' activity value for the
-test molecule. This value can be used to extend the original landscape. The text has been updated to
-discuss this correctly XXX
-
-Of course, if one predicts a SALI value for a pair of molecules, in which both members have an
-unknown actvity, one cannot extend the landscape.
+The initial description was in misleading. In fact, the approach does not allow one to obtain
+predicted activity values, due to the fact that the sign of the difference in activities in the SALI
+formula is ignored. While one could create another model in parallel, that predicts signed activity
+differences (and thereby determine whether the activity derived from the predicted SALI value should
+be less than or greater than that of the molecule with observed activity), this would probably not
+be very reliable. The manuscript has been updated to remove the dicussion on extending SAR
+landscapes.
 

### 2012-07-20 rebuttal.tex
more fixes

@@
 
+In refering to pairs with $T_C$ values between 0.2 and 0.3 -- it is true that such pairs are not
+really activity cliffs in the original sense of the term. As described below, defining a threshold
+SALI, above which a pair represents an activity cliff, is rather subjective.
+
 \textbf{Thus, on the basis of these similarity calculations, one could never decide which level of
@@
 
-I agree that the wording of this discussion is rather sloppy. The text has been updated to be more
+I agree that the wording of this discussion is somewhat sloppy. The text has been updated to be more
 quantitative.
 
-Yet at the same time, the concept of an activity cliff is somewhat subjective. In many cases, it is
+Yet at the same time, the concept of an activity cliff is subjective. In many cases, it is
 akin to saying \emph{``I'll know it's an activity cliff when I see it''}. For example, a pair of
@@
 ``cliffness'' of an activity cliff.
+
+Certainly, with respect to the previous comment, a pair of molecules with $T_c = 0.29$ would likely
+not be regarded as a cliff - I specifically note that such a case is not a significant cliff. To
+address this aspect of SALI values, I have presented plots of dissimilarity versus activity ratio
+(Figures 7 and 9) which allow one to more directly determine ``significant'' versus
+``insignificant'' cliffs.
 
@@
 
-The initial description was in misleading. In fact, the approach does not allow one to obtain
+The original description was misleading. In fact, the approach does not allow one to obtain
 predicted activity values, due to the fact that the sign of the difference in activities in the SALI
@@
 be less than or greater than that of the molecule with observed activity), this would probably not
-be very reliable. The manuscript has been updated to remove the dicussion on extending SAR
+be very reliable! The manuscript has been updated to remove the dicussion on extending SAR
 landscapes.

### 2012-07-20 paper.tex
Added more rebuttal
removed discussion of extending landscapes. Updated rebuttal
more fixes

@@
 independent and dependent variables. More specifically, both the dependent variable (i.e., the SALI
-values) and the aggregated descriptors characterize the molecules' structure. One might therefore
-ask whether models built using such data perform over-optimistically. Given the nature of the
-descriptors used in the independent and dependent variables (the former being based on multiple atom
-and bond features and the latter derived from purely topological paths), we believe that such a
-phenomenon is unlikely.  Fig.~\ref{fig:desccor1} displays a histogram of the pairwise Pearson
-correlations between the SALI values and each of the descriptor values for the
+values) and the aggregated (topological) descriptors characterize the molecules' structure. One
+might therefore ask whether models built using such data perform over-optimistically. Given the
+nature of the descriptors used in the independent and dependent variables (the former being based on
+multiple atom and bond features and the latter derived from purely topological paths), we believe
+that such a phenomenon is unlikely.  Fig.~\ref{fig:desccor1} displays a histogram of the pairwise
+Pearson correlations between the SALI values and each of the descriptor values for the
 Cavalli\cite{Cavalli:2002aa} dataset, using the different aggregation functions described above. It
@@
 datasets and so the following discussion omits the results obtained when using
-$f_{\textrm{geom}}$. Fig.~\ref{fig:rfmodel-chembl} summarizes the models built on the ChEMBL
+$f_{\textrm{geom}}$. In fact, a paired $t$-test between the predicted SALI values using models built
+with the different aggregation functions (i.e., $f_{\textrm{mean}}$ vs $f_{\textrm{absdiff}}$, etc.)
+indicated that the differences between the sets of predictions was statistically insignificant (all
+$p$-values $> 0.1$). Fig.~\ref{fig:rfmodel-chembl} summarizes the models built on the ChEMBL
 datasets using $f_{\textrm{diff}}$ and $f_{\textrm{mean}}$ as the aggregation functions. Table
@@
 
-Given the relatively small differences between models based on different aggregation functions, we
-arbitrarily chose $f_{\textrm{mean}}$ as the aggregation function for subsequent
-analyses. Figs.~\ref{fig:costanzo-pred}, \ref{fig:kalla-pred} and \ref{fig:dai-pred} summarize the
-performance of the models built using the reported activity data and the pairwise SALI values for
-the Costanzo, Kalla and Dai datasets, respectively. As with the Cavalli dataset, the SALI models
-exhibit more variance for small SALI values (i.e., less significant cliffs) but appear to perform
-better at the higher end of the SALI values. However in all three cases, the most significant cliffs
-exhibit significant variance. In comparison, the models developed on the reported activity values
-are relatively poor. This can be partially ascribed to the small sizes of the dataset. It should be
-note that this methodology is not aimed at modeling the original, usually small, SAR dataset. It is
-well known (and evident from the current analyses) that small datasets do not lead to very robust or
-reliable machine learning models. Instead, the methodology focuses on the \emph{pairwise} dataset --
-which are usually large enough for a statistically significant model.
+Given this observation, we arbitrarily chose $f_{\textrm{mean}}$ as the aggregation function
+for subsequent analyses. Figs.~\ref{fig:costanzo-pred}, \ref{fig:kalla-pred} and \ref{fig:dai-pred}
+summarize the performance of the models built using the reported activity data and the pairwise SALI
+values for the Costanzo, Kalla and Dai datasets, respectively. As with the Cavalli dataset, the SALI
+models exhibit more variance for small SALI values (i.e., less significant cliffs) but appear to
+perform better at the higher end of the SALI values. However in all three cases, the most
+significant cliffs exhibit significant variance. In comparison, the models developed on the reported
+activity values are relatively poor. This can be partially ascribed to the small sizes of the
+dataset. It should be note that this methodology is not aimed at modeling the original, usually
+small, SAR dataset. It is well known (and evident from the current analyses) that small datasets do
+not lead to very robust or reliable machine learning models. Instead, the methodology focuses on the
+\emph{pairwise} dataset -- which are usually large enough for a statistically significant model.
 
-\subsection{Extending a landscape}
-\label{sec:extending-landscape}
-
-While predictive modeling of the SALI values is a first step, the primary goal of this approach is
-to extend the structure-activity landscape for a set of molecules. In other words, can one use the
-model to identify compounds that will exhibit an activity cliff with one or more molecules in the
-dataset? Given a new molecule, one can evaluate the descriptors for this molecule and then evaluate
-the pairwise descriptor values with each member of the training set. Thus, given a new molecule, we
-must obtain predictions for $n$ ``pairwise objects'', where $n$ is the number of molecules in the
-training set. Given these $n$ predictions one can proceed to identify training set molecules with
-which the new molecule might exhibit a cliff.
+Since the model is built on pairs of observations, performance measures characterizing the model are
+based on the accuracy of pairwise SALI values. The model can also be tested in a slightly different
+approach by checking whether it can correctly identify cliffs between a new molecule and the member
+of the training set. To test this, we consider the original set of $n$ individual molecules and hold out
+$m$ molecules (as opposed to holding out a set of observations from the pairwise dataset). We then
+evaluate the SALI values $(n-m)(n-m-1)/2$ pairs of molecules and develop a random forest model. The
+model is then used to predict the SALI values between the $m$ holdout molecules and the $(n-m)$
+remaining molecules.
 

### 2012-07-24 rebuttal.tex
Added more to the rebuttal

@@
 
-I would like to thank the reviewers for their thorough comments on the original submissions. I have
-prepared a revised manuscript that address the comments and my responses are detailed below.
+I would like to thank the reviewers for their comments on the original submissions. I have prepared
+a revised manuscript that address the comments and my responses are detailed below.
 
@@
 
-(6) There is only limited statistical assessment/validation of the models. Statements made in the
+\textbf{(6) There is only limited statistical assessment/validation of the models. Statements made in the
 text that R2 values are “reasonable” and the models perform “relatively well” are difficult to
-reconcile. Rather, Table 2 indicates that the R2 values were rather poor in several instances.
+reconcile. Rather, Table 2 indicates that the R2 values were rather poor in several instances.}
+
+The text has been updated to read more quantitatively and avoid ambiguous statements
 
@@
 
+While it is true that the SALI definition itself allows for high SALI values to be generated for
+pairs of molecules that would be small/insignificant cliffs (highly similar structures with small
+difference in activity), the methodology does let one rank pairs of molecules and as indicated in
+Figures 7 and 9, one can compare the strucural similarity and activity difference (or ratio) in 2D,
+allowing one to separate ``true'' cliffs from apparent ones.
+
 \textbf{(10) It also remains largely obscure how an activity landscape should be extended following

### 2012-07-25 paper.tex
some more fixes

@@
 observation from from all 500 trees. We also investigated the traditional approach of a
-training/test set split
+training/test set split (80\% of the dataset was assigned to the training set) and the results were
+similar. As a result, the remainder of this work employed models built using the entire dataset.
 
@@
 descriptors with constant or near-constant values followed by removal of descriptors that are highly
-correlated with others (using an $R^2$ cutoff of 0.8). The sizes of the final descriptor pool for
-each dataset are summarized in Table \ref{tab:data}. XXX
+correlated with others (using an $R^2$ cutoff of 0.8). The sizes ($N_{desc}$) of the final
+descriptor pool for each dataset are summarized in Table \ref{tab:data}.
 
@@
 label={tab:data}]
-{lp{0.4\textwidth}rrrr}
+{lp{0.3\textwidth}rrrrr}
 {}
 {\FL
-Name & Description & Endpoint & ChEMBL AID & Size & Reference
+Name & Description & Endpoint & ChEMBL AID & Size & $N_{desc}$ & Reference
 \ML
-Cavalli & \raggedright{Putative hERG inhibitors (via pharmacophore modeling)} & Score &  & 30 & Ref.~\citen{Cavalli:2002aa} \NN
-Costanzo & \raggedright{Thrombin inhibitors} & $\mathrm{IC_{50}}$ & 305283 & 60 &
+Cavalli & \raggedright{Putative hERG inhibitors (via pharmacophore modeling)} & pIC$_{50}$ &  & 30 & 40 &  Ref.~\citen{Cavalli:2002aa} \NN
+Costanzo & \raggedright{Thrombin inhibitors} & $\mathrm{IC_{50}}$ & 305283 & 60 & 21 & 
 Ref.~\citen{Costanzo:2005uq} \NN
 Kalla & \raggedright{A2B adenosine receptor antagonists} & $K_i$ &
-364155 & 38 & Ref.~\citen{Kalla:2006kx} \NN
-Dai & \raggedright{RTK inhibitors} & $K_i$ & 429056 & 44 & Ref.~\citen{Dai:2007kl}
+364155 & 38 & 10 & Ref.~\citen{Kalla:2006kx} \NN
+Dai & \raggedright{RTK inhibitors} & $K_i$ & 429056 & 44 & 12 & Ref.~\citen{Dai:2007kl}
 % Mewshaw & \raggedright{ER$\beta$ inhibitors} & $\mathrm{IC_{50}}$ &

### 2012-07-25 rebuttal.tex
Added in tset/pset split results and updated figures for Cavalli dataset

@@
 
-(4) The selection of training and test sets should be clearly specified in the Methods section. For
+\textbf{(4) The selection of training and test sets should be clearly specified in the Methods section. For
 example, how were the RF models trained and evaluated for which predictions of pair-wise SALI values
 are reported in Figure 2? It is hoped that the reported SALI values were not predicted for the
-training data.
+training data.}
 
-(5) No support is provided for the assumption that the RF models “are relatively robust even for
+XXX
+
+\textbf{(5) No support is provided for the assumption that the RF models “are relatively robust even for
 datasets of small size”. In fact, the author later on states several times that limited model
@@
 here.
+}
+
+XXX
 

### 2012-07-25 paper.tex
Added in tset/pset split results and updated figures for Cavalli dataset

@@
 \usepackage{longtable}
+%%\usepackage{booktabs}
 
@@
 \ctable[cap={}, caption={Performance metrics for the random forest
-  models built for the datasets used in this study. RMSE and $R^2$ are
-based on out-of-bag data\cite{Breiman:1984aa}},
+  models built for the datasets used in this study. Training RMSE and $R^2$ are
+based on out-of-bag data\cite{Breiman:1984aa} and test metrics are based on a separate
+hold out set. },
 label={tab:rfmodel-results}]
-{lrcrr}
-{}
+{lrcrrrr}
+{
+\tnote{Y-range is calculated using the entire dataset for each study.}
+}
 {\FL
-Dataset & Y-range & Aggregation Function & RMSE & $R^2$ 
+        &         &                      & \multicolumn{2}{c}{Training} & \multicolumn{2}{c}{Test}
+\NN
+\cmidrule{4-5}  \cmidrule{6-7}
+Dataset & Y-range\tmark & Aggregation Function & RMSE & $R^2$ & RMSE & $R^2$   
 \ML
-Cavalli & 7.21 & $f_{\textrm{diff}}$ & 1.11& 0.57\NN
-        & & $f_{\textrm{mean}}$ & 0.98& 0.67\NN
-        & & $f_{\textrm{geom}}$ & 1.01& 0.65\NN \NN
-Costanzo & 288.03& $f_{\textrm{diff}}$ & 12.88& 0.24\NN
-       & & $f_{\textrm{mean}}$ & 13.56& 0.17\NN \NN
-Kalla & 88.25 & $f_{\textrm{diff}}$ & 7.51& 0.62\NN
-      &  & $f_{\textrm{mean}}$ & 8.26& 0.53\NN \NN
-Dai & 32.71 & $f_{\textrm{diff}}$ & 2.60& 0.44\NN
-     &   & $f_{\textrm{mean}}$ & 2.67& 0.41
+Cavalli & 7.21 & $f_{\textrm{diff}}$ & 1.16& 0.54 & 1.07& 0.56\NN
+        & & $f_{\textrm{mean}}$      & 1.03& 0.64 & 1.01& 0.58\NN
+        & & $f_{\textrm{geom}}$      & 1.07& 0.61 & 1.01& 0.59\NN \NN
+Costanzo & 288.03& $f_{\textrm{diff}}$ & 12.88& 0.24 & & \NN
+       & & $f_{\textrm{mean}}$ & 13.56& 0.17 & & \NN \NN
+Kalla & 88.25 & $f_{\textrm{diff}}$ & 7.51& 0.62 & & \NN
+      &  & $f_{\textrm{mean}}$ & 8.26& 0.53  & & \NN \NN
+Dai & 32.71 & $f_{\textrm{diff}}$ & 2.60& 0.44 & & \NN
+     &   & $f_{\textrm{mean}}$ & 2.67& 0.41  & & 
 \LL

### 2012-07-25 paper.tex
updated some figures for datasets

@@
 
-\ctable[cap={}, caption={Datasets considered in this study.},
+\ctable[cap={}, caption={Datasets considered in this study. $N_{desc}$ is the number of descriptors
+  for the dataset after removal of correlated and low-variance descriptors from the original pool of
+109 descriptors.},
 label={tab:data}]
@@
         & & $f_{\textrm{geom}}$      & 1.07& 0.61 & 1.01& 0.59\NN \NN
-Costanzo & 288.03& $f_{\textrm{diff}}$ & 12.88& 0.24 & & \NN
-       & & $f_{\textrm{mean}}$ & 13.56& 0.17 & & \NN \NN
-Kalla & 88.25 & $f_{\textrm{diff}}$ & 7.51& 0.62 & & \NN
-      &  & $f_{\textrm{mean}}$ & 8.26& 0.53  & & \NN \NN
-Dai & 32.71 & $f_{\textrm{diff}}$ & 2.60& 0.44 & & \NN
-     &   & $f_{\textrm{mean}}$ & 2.67& 0.41  & & 
+%%
+Costanzo & 288.03& $f_{\textrm{diff}}$ & 11.99& 0.24 & 17.97& 0.10 \NN
+       & & $f_{\textrm{mean}}$         & 12.88& 0.12 & 16.96& 0.20\NN \NN
+%%
+Kalla & 88.25 & $f_{\textrm{diff}}$ & 8.12& 0.59 & 6.49& 0.57\NN
+      &  & $f_{\textrm{mean}}$      & 8.79& 0.52 & 8.00& 0.47\NN \NN
+%%
+Dai & 32.71 & $f_{\textrm{diff}}$ & 2.49& 0.48 & 3.21& 0.22\NN
+     &   & $f_{\textrm{mean}}$    & 2.64& 0.41 & 2.86& 0.37
 \LL

### 2012-07-26 rebuttal.tex
Added more edits to tset/pset split

@@
 
-XXX
+We have updated the text to note the training/test split and have include results for both sets in
+the text. However, due to the similar peformance of models built using the whole dataset (and the
+fact that the RF approach trains and tests on out-of-bag data implicitly), we used models built
+using the entire dataset (save for some hold out molecules) in the latter half of Section 4.
 
@@
 quality might be due to the use of small compound sets. Appropriate validation will be essential
-here.
-}
+here.}
 
-XXX
+We have removed this statement as it is not necessarily a function of the random forest approach,
+and more a function of the nature of the dataset. Small datasets, if appropriately constructed can
+lead to low RMSE. In our case, we simulated by sampling from the Cavalli pairwise dataset, using
+different sample sizes. As shown in the attached figure, model performance is clearly a function of
+sample size. Importantly, for certain small samples, the model performance can be quite good. But of
+course, this is due to luck.
 
@@
 
-(7) It appears that there are systematic prediction errors for pairs. At low value ranges (that are
+\textbf{(7) It appears that there are systematic prediction errors for pairs. At low value ranges (that are
 not relevant for activity cliff assessment), SALI values are consistently over-predicted. By
@@
 part significantly. This is not investigated, although the apparent errors at high value ranges are
-the perhaps most critical features of the reported RF models.
+the perhaps most critical features of the reported RF models.}
 
-(8) In the section “Extending a landscape”, predictions are finally reported for only three (!)
+XXX
+
+\textbf{(8) In the section “Extending a landscape”, predictions are finally reported for only three (!)
 molecules taken out of each data set. Carefully put, the predictions are heterogeneous. On the basis
 of the results reported for this little bit of evaluation, one would be hard pressed to make a case
-for the ability of the approach to predict activity cliffs with any certainty.
+for the ability of the approach to predict activity cliffs with any certainty.}
+
+XXX
 

### 2012-07-26 paper.tex
Added more edits to tset/pset split

@@
 predicted values for the entire dataset are obtained by averaging the predictions for each
-observation from from all 500 trees. We also investigated the traditional approach of a
-training/test set split (80\% of the dataset was assigned to the training set) and the results were
-similar. As a result, the remainder of this work employed models built using the entire dataset.
+observation from from all 500 trees. We initially employed a
+training/test set split (80\% of the dataset was assigned to the training set) and these results are
+reported in Table \ref{tab:data}.
 
@@
 performance metrics for the three models. Overall the different aggregation functions do not differ
-dramatically in terms of the final model performance as root mean square error (RMSE) ranges from
-0.98 to 1.11 log units. However, $f_{\textrm{mean}}$ appears to lead to the best performance (RMSE = 0.98 and
-$R^2$ = 0.67). In contrast, the RMSE and $R^2$ values for Y-scrambled\cite{Rucker:2007aa} models
-using the different aggregation functions ranged from 1.58 to 1.61 log units and 0.003 to 0.01, respectively.
+dramatically in terms of the final model performance as root mean square error (RMSE) on the test
+set ranges from 1.01 to 1.07 log units. However, $f_{\textrm{mean}}$ appears to lead to the best
+performance (RMSE = 1.01 and $R^2$ = 0.58). In contrast, the RMSE and $R^2$ values for
+Y-scrambled\cite{Rucker:2007aa} models using the different aggregation functions ranged from 1.58 to
+1.61 log units and 0.003 to 0.01, respectively.
 
@@
 datasets and so the following discussion omits the results obtained when using
-$f_{\textrm{geom}}$. In fact, a paired $t$-test between the predicted SALI values using models built
-with the different aggregation functions (i.e., $f_{\textrm{mean}}$ vs $f_{\textrm{absdiff}}$, etc.)
-indicated that the differences between the sets of predictions was statistically insignificant (all
-$p$-values $> 0.1$). Fig.~\ref{fig:rfmodel-chembl} summarizes the models built on the ChEMBL
-datasets using $f_{\textrm{diff}}$ and $f_{\textrm{mean}}$ as the aggregation functions. Table
-\ref{tab:rfmodel-results} reports the model statistics.  While the differences in model performance
-do not vary significantly with the aggregation function used, we see that for the Costanzo and Kalla
-datasets, $f_{\textrm{diff}}$ leads to a slightly better model in terms of RMSE and $R^2$ than when
-using $f_{\textrm{mean}}$.
+$f_{\textrm{geom}}$. In fact, a paired $t$-test between the predicted (from the training set) SALI
+values using models built with the different aggregation functions (i.e., $f_{\textrm{mean}}$ vs
+$f_{\textrm{absdiff}}$, etc.)  indicated that the differences between the sets of predictions was
+statistically insignificant (all $p$-values $> 0.1$). Fig.~\ref{fig:rfmodel-chembl} summarizes the
+models built on the ChEMBL datasets using $f_{\textrm{diff}}$ and $f_{\textrm{mean}}$ as the
+aggregation functions. Table \ref{tab:rfmodel-results} reports the model statistics.  While the
+differences in model performance do not vary significantly with the aggregation function used, we
+see that for all the ChEMBL datasets, $f_{\textrm{diff}}$ leads to a slightly better model
+in terms of RMSE and $R^2$ than when using $f_{\textrm{mean}}$.
 
-Given this observation, we arbitrarily chose $f_{\textrm{mean}}$ as the aggregation function
+Given this observation, we arbitrarily chose $f_{\textrm{diff}}$ as the aggregation function
 for subsequent analyses. Figs.~\ref{fig:costanzo-pred}, \ref{fig:kalla-pred} and \ref{fig:dai-pred}

### 2012-07-26 rebuttal.tex
updated fig for rf models on original activity. Updated kalla tset/pset plot

@@
 and more a function of the nature of the dataset. Small datasets, if appropriately constructed can
-lead to low RMSE. In our case, we simulated by sampling from the Cavalli pairwise dataset, using
+lead to good RMSE's. In our case, we investigated this by sampling from the Cavalli pairwise dataset, using
 different sample sizes. As shown in the attached figure, model performance is clearly a function of

### 2012-07-26 paper.tex
updated fig for rf models on original activity. Updated kalla tset/pset plot

@@
 using 63\% of the dataset and tested on the remainder (the so called out-of-bag data). The final
-predicted values for the entire dataset are obtained by averaging the predictions for each
-observation from from all 500 trees. We initially employed a
-training/test set split (80\% of the dataset was assigned to the training set) and these results are
-reported in Table \ref{tab:data}.
+predicted values are obtained by averaging the predictions for each observation from from all 500
+trees. We initially employed a training/test set split (80\% of the dataset was assigned to the
+training set) and these results are reported in Table \ref{tab:rfmodel-results}. We also developed
+models (Section \ref{sec:an-altern-valid} using the entire dataset and observed similar RMSE and
+$R^2$ values. Given the fact that the random forest algorithm builds a model based on out-of-bag
+data, using the entire dataset rather than a single training/test split this is not surprising.
 
@@
 differences in model performance do not vary significantly with the aggregation function used, we
-see that for all the ChEMBL datasets, $f_{\textrm{diff}}$ leads to a slightly better model
-in terms of RMSE and $R^2$ than when using $f_{\textrm{mean}}$.
+see that for all the ChEMBL datasets, $f_{\textrm{diff}}$ leads to a slightly better model in terms
+of RMSE and $R^2$ than when using $f_{\textrm{mean}}$. As with the Cavalli dataset, the SALI models
+exhibit more variance for small SALI values (i.e., less significant cliffs) but appear to perform
+better at the higher end of the SALI values. However in all three cases, the most significant cliffs
+exhibit significant variance. Based on these observations, we arbitrarily chose $f_{\textrm{diff}}$
+as the aggregation function for subsequent analyses. Fig.~\ref{fig:rfmodel-orig} summarizes the
+performance of the models built using the reported activity data for the Costanzo, Kalla and Dai
+datasets. Each of these model employed the entire dataset (rather than splitting into a training and
+testing set). These models are relatively poor. This can be partially ascribed to the small sizes of
+the dataset. It should be note that this methodology is not aimed at modeling the original, usually
+small, SAR dataset. It is well known (and evident from the current analyses) that small datasets do
+not necessarily lead to very robust or reliable machine learning models. Instead, the methodology
+focuses on the \emph{pairwise} dataset -- which are usually large enough for a statistically
+significant model.
 
-Given this observation, we arbitrarily chose $f_{\textrm{diff}}$ as the aggregation function
-for subsequent analyses. Figs.~\ref{fig:costanzo-pred}, \ref{fig:kalla-pred} and \ref{fig:dai-pred}
-summarize the performance of the models built using the reported activity data and the pairwise SALI
-values for the Costanzo, Kalla and Dai datasets, respectively. As with the Cavalli dataset, the SALI
-models exhibit more variance for small SALI values (i.e., less significant cliffs) but appear to
-perform better at the higher end of the SALI values. However in all three cases, the most
-significant cliffs exhibit significant variance. In comparison, the models developed on the reported
-activity values are relatively poor. This can be partially ascribed to the small sizes of the
-dataset. It should be note that this methodology is not aimed at modeling the original, usually
-small, SAR dataset. It is well known (and evident from the current analyses) that small datasets do
-not lead to very robust or reliable machine learning models. Instead, the methodology focuses on the
-\emph{pairwise} dataset -- which are usually large enough for a statistically significant model.
+\subsection{An alternative validation approach}
+\label{sec:an-altern-valid}
 
@@
 approach by checking whether it can correctly identify cliffs between a new molecule and the member
-of the training set. To test this, we consider the original set of $n$ individual molecules and hold out
-$m$ molecules (as opposed to holding out a set of observations from the pairwise dataset). We then
-evaluate the SALI values $(n-m)(n-m-1)/2$ pairs of molecules and develop a random forest model. The
-model is then used to predict the SALI values between the $m$ holdout molecules and the $(n-m)$
-remaining molecules.
+of the training set. To test this, we consider the original set of $n$ individual molecules and hold
+out $m$ molecules (as opposed to holding out a set of observations from the pairwise dataset). We
+then evaluate the SALI values $(n-m)(n-m-1)/2$ pairs of molecules and develop a random forest
+model. In contrast to the models previously described, we rebuilt these models using the entire set
+of $n$ molecules. The model is then used to predict the SALI values between the $m$ holdout
+molecules and the $(n-m)$ remaining molecules.
 
@@
 Ref.~\citen{Costanzo:2005uq} \NN
-Kalla & \raggedright{A2B adenosine receptor antagonists} & $K_i$ &
+Kalla & \raggedright{A2B adenosine receptor antagonists} & $429056K_i$ &
 364155 & 38 & 10 & Ref.~\citen{Kalla:2006kx} \NN
@@
 \ctable[cap={}, caption={Performance metrics for the random forest
-  models built for the datasets used in this study. Training RMSE and $R^2$ are
-based on out-of-bag data\cite{Breiman:1984aa} and test metrics are based on a separate
-hold out set. },
+  models built for the datasets used in this study.},
 label={tab:rfmodel-results}]
@@
 \ctable[cap={}, caption={Results of random forest models developed
-  using the Costanzo dataset. The left plot shows the predicted versus
-  observed plot for the model using the original $\mathrm{IC_{50}}$ values, and
-  the right hand plot shows the predicted versus observed log(SALI)
-  values.}, label={fig:costanzo-pred}, figure, botcap]
+  using the three ChEMBL dataset. In each case the model was built using the $\log_{10}$ of the
+  observed activity.}, label={fig:rfmodel-orig}, figure, botcap]
 {c}
 {}
-{\includegraphics[width=\linewidth]{pred305283}}
-
-\ctable[cap={}, caption={Results of random forest models developed
-  using the Kalla dataset. The left plot shows the predicted versus
-  observed plot for the model using the original $K_i$ values, and
-  the right hand plot shows the predicted versus observed log(SALI)
-  values.}, label={fig:kalla-pred}, figure, botcap]
-{c}
-{}
-{\includegraphics[width=\linewidth]{pred364155}}
-
-\ctable[cap={}, caption={Results of random forest models developed
-  using the Dai dataset. The left plot shows the predicted versus
-  observed plot for the model using the original $K_i$ values, and
-  the right hand plot shows the predicted versus observed log(SALI)
-  values.}, label={fig:dai-pred}, figure, botcap]
-{c}
-{}
-{\includegraphics[width=\linewidth]{pred429056}}
+{\includegraphics[width=\linewidth]{rfmodel-orig}}
 

### 2012-07-26 paper.tex
updated log sali histogram

@@
   log(SALI) values for the training set and the hold out
-  set. \textbf{B} - a summary of the training set, where we plot the
+  set. \textbf{B} - a summary of the training set, which plots the
   structural difference versus the
@@
   log(SALI) values for the training set and the hold out
-  set. \textbf{B} - a summary of the training set, where we plot the
+  set. \textbf{B} - a summary of the training set, which plots the
   structural difference versus the logarithm of the ratio of the

### 2012-07-26 paper.tex
updated some more figs

@@
 Ref.~\citen{Costanzo:2005uq} \NN
-Kalla & \raggedright{A2B adenosine receptor antagonists} & $429056K_i$ &
+Kalla & \raggedright{A2B adenosine receptor antagonists} & $K_i$ &
 364155 & 38 & 10 & Ref.~\citen{Kalla:2006kx} \NN
@@
 
-\ctable[cap={}, caption={The hold out molecules for the Dai dataset
+\ctable[cap={}, caption={Three of the  hold out molecules for the Dai dataset
   and training set members with which the hold outs exhibit predicted

### 2012-07-26 paper.tex
updated more figures

@@
 predicted values are obtained by averaging the predictions for each observation from from all 500
-trees. We initially employed a training/test set split (80\% of the dataset was assigned to the
-training set) and these results are reported in Table \ref{tab:rfmodel-results}. We also developed
-models (Section \ref{sec:an-altern-valid} using the entire dataset and observed similar RMSE and
-$R^2$ values. Given the fact that the random forest algorithm builds a model based on out-of-bag
-data, using the entire dataset rather than a single training/test split this is not surprising.
+trees. We initially employed a training/test set split (80\% of the dataset was randomly assigned to
+the training set) and these results are reported in Table \ref{tab:rfmodel-results}. We also
+developed models (Section \ref{sec:an-altern-valid}) using the entire dataset and observed similar
+RMSE and $R^2$ values. Given the fact that the random forest algorithm builds a model based on
+out-of-bag data, the similarity in performance of models built using the entire dataset and models
+built using a training/test split is not surprising.
 
-In this work, we employed the CDK to evaluate 109 2D and constitutional descriptors for each
-individual molecule. For each dataset, we performed objective feature selection by removing
-descriptors with constant or near-constant values followed by removal of descriptors that are highly
-correlated with others (using an $R^2$ cutoff of 0.8). The sizes ($N_{desc}$) of the final
-descriptor pool for each dataset are summarized in Table \ref{tab:data}.
+We employed the CDK to evaluate 109 2D and constitutional descriptors for each  molecule. For each
+dataset, we performed objective feature selection by removing descriptors with constant or
+near-constant values followed by removal of descriptors that are highly correlated with others
+(using an $R^2$ cutoff of 0.8). The sizes ($N_{desc}$) of the final descriptor pool for each dataset
+are summarized in Table \ref{tab:data}.
 
@@
 exhibit significant variance. Based on these observations, we arbitrarily chose $f_{\textrm{diff}}$
-as the aggregation function for subsequent analyses. Fig.~\ref{fig:rfmodel-orig} summarizes the
-performance of the models built using the reported activity data for the Costanzo, Kalla and Dai
-datasets. Each of these model employed the entire dataset (rather than splitting into a training and
-testing set). These models are relatively poor. This can be partially ascribed to the small sizes of
-the dataset. It should be note that this methodology is not aimed at modeling the original, usually
-small, SAR dataset. It is well known (and evident from the current analyses) that small datasets do
-not necessarily lead to very robust or reliable machine learning models. Instead, the methodology
+as the aggregation function for subsequent analyses. Table \ref{tab:rfmodel-orig} and
+Figure \ref{fig:rfmodel-orig} summarizes the performance of the models built using the reported
+activity data (original, single molecule activity) for the Costanzo, Kalla and Dai datasets. Each of
+these models employed the entire dataset (rather than splitting into a training and testing set) and
+exhibit relatively poor performance as indicated by the RMSE (ranging from 0.76 to 1.00 log units)
+and $R^2$ (0.18 to 0.34).. This can be partially ascribed to the small sizes of the dataset. It
+should be note that this methodology is not aimed at modeling the original, usually small, SAR
+dataset. It is well known (and evident from the current analyses) that small datasets do not
+necessarily lead to very robust or reliable machine learning models. Instead, the methodology
 focuses on the \emph{pairwise} dataset -- which are usually large enough for a statistically
@@
 Since the model is built on pairs of observations, performance measures characterizing the model are
-based on the accuracy of pairwise SALI values. The model can also be tested in a slightly different
-approach by checking whether it can correctly identify cliffs between a new molecule and the member
-of the training set. To test this, we consider the original set of $n$ individual molecules and hold
-out $m$ molecules (as opposed to holding out a set of observations from the pairwise dataset). We
-then evaluate the SALI values $(n-m)(n-m-1)/2$ pairs of molecules and develop a random forest
-model. In contrast to the models previously described, we rebuilt these models using the entire set
-of $n$ molecules. The model is then used to predict the SALI values between the $m$ holdout
-molecules and the $(n-m)$ remaining molecules.
+based on the accuracy of pairwise SALI values. Thus the preceding section validated models based on
+holding out pairs of observations and evaluating the RMSE and $R^2$ for the predicted values of
+these held out pairs.
+
+The model can also be tested in a slightly different approach by checking whether it can correctly
+identify cliffs between a new molecule and the member of the training set. In fact, such a scenario
+would be the anticipated, prospective use case for such a modeling methodology. To test this, we
+consider the original set of $n$ individual molecules in a dataset and hold out $m$ molecules (as
+opposed to holding out a set of observations from the pairwise dataset). We then evaluate the SALI
+values $(n-m)(n-m-1)/2$ pairs of molecules and develop a random forest model. In contrast to the
+models previously described, we rebuilt these models using the entire set of $n$ molecules. The
+model is then used to predict the SALI values between the $m$ holdout molecules and the $(n-m)$
+remaining molecules.
 
@@
 values for these three with the training set molecules and obtained predictions of the log(SALI)
-values. Fig.~\ref{fig:tset-pset-kalla}A summarizes the performance of the model on the training set
+values. Figure \ref{fig:tset-pset-kalla}A summarizes the performance of the model on the training set
 as well as for the prediction set. From a numerical point of view, the performance of the model on
@@
 we are interested in the predictions for the actual activity cliffs, where both the difference in
-activity and the structural similarity is very high. Fig.~\ref{fig:tset-pset-kalla}B visualizes this
+activity and the structural similarity is very high. Figure \ref{fig:tset-pset-kalla}B visualizes this
 information. Each point corresponds to a pair of molecules, which are shaded by the absolute
@@
 
-Fig.~\ref{fig:kalla-comparison} displays the structures of the three hold out molecules (top row)
+Figure \ref{fig:kalla-comparison} displays the structures of the three hold out molecules (top row)
 and a selection of structures, which which these hold out molecules were predicted to exhibit an
@@
 of the residuals were were relatively high. While we have highlighted a number of cliffs,
-Fig.~\ref{fig:tset-pset-kalla}C makes it clear that the predictions for this molecule were, on
+Figure \ref{fig:tset-pset-kalla}C makes it clear that the predictions for this molecule were, on
 average, worse than for the other two molecules.
 
-Fig.~\ref{fig:tset-pset-dai} summarizes the results for the same set of analyzes described above,
+Figure \ref{fig:tset-pset-dai} summarizes the results for the same set of analyzes described above,
 applied to the Dai dataset. As before we removed three molecules (top row of
-Fig.~\ref{fig:dai-comparison}) and rebuilt the random forest on the paired cases with the remaining
+Figure \ref{fig:dai-comparison}) and rebuilt the random forest on the paired cases with the remaining
 41 molecules. We then predicted the pairwise SALI values for the three hold-out molecules with the
 remaining 41 molecules (i.e., the prediction set). A plot of the predicted versus observed log(SALI)
-values for the training and prediction sets are displayed in Fig.~\ref{fig:tset-pset-dai}A. As
+values for the training and prediction sets are displayed in Figure \ref{fig:tset-pset-dai}A. As
 before the model performs poorly on the lesser cliffs, but improves for the more significant cliffs
@@
 distribution of absolute residuals aggregated by the hold-out molecule
-(Fig.~\ref{fig:tset-pset-dai}C) are similar, with a median absolute residual of less than
+(Figure \ref{fig:tset-pset-dai}C) are similar, with a median absolute residual of less than
 0.3 log units.  For this dataset, the number of ``real'' cliffs is relatively low as shown in
-Fig.~\ref{fig:tset-pset-dai}B. Interestingly, the compound pairs that display moderate to
+Figure \ref{fig:tset-pset-dai}B. Interestingly, the compound pairs that display moderate to
 significant cliffs are relatively well predicted - in fact, the maximum residuals are observed for
@@
 units. However, the activity differences are not always significant ranging from 1.1-fold to
-13-fold. Two such examples are shown in Fig.~\ref{fig:dai-comparison}, where the only difference is
+13-fold. Two such examples are shown in Figure \ref{fig:dai-comparison}, where the only difference is
 in the position of the methyl substituent. Similar behavior is observed with the other hold-out
@@
 fact that the distribution of the log(SALI) values tends to be skewed to the right
-(Fig.~\ref{fig:logsalidist}). In general we see that the smaller SALI values constitute a relatively
+(Figure \ref{fig:logsalidist}). In general we see that the smaller SALI values constitute a relatively
 small portion of the dataset.
@@
 
+\ctable[cap={}, caption={Performance metrics for the random forest
+  models built on the single molecule activity data, for each ChEMBL dataset. In each case the model
+was built using the logarithm of the reported activities.},
+label={tab:rfmodel-orig}]
+{lrrr}
+{
+
+}
+{\FL
+Dataset & Y-range & RMSE & $R^2$
+\ML
+Costanzo & 4.67 & 0.80 & 0.18 \NN
+Kall          & 4.44 & 1.00 & 0.32 \NN
+Dai          &  3.60 & 0.76 & 0.34
+\LL
+}
+
 \clearpage
@@
 
-\ctable[cap={}, caption={The hold out molecules for the Kalla dataset
+\ctable[cap={}, caption={Three of the hold out molecules for the Kalla dataset
   and training set members with which the hold outs exhibit predicted

### 2012-07-26 paper.tex
more edits ot the paper

@@
 The model can also be tested in a slightly different approach by checking whether it can correctly
-identify cliffs between a new molecule and the member of the training set. In fact, such a scenario
+identify cliffs between a new molecule and the members of the training set. In fact, such a scenario
 would be the anticipated, prospective use case for such a modeling methodology. To test this, we
@@
 model is then used to predict the SALI values between the $m$ holdout molecules and the $(n-m)$
-remaining molecules.
+remaining molecules. As a result, instead of a just $m$ predicted values, we obtain $m \times (n-m)$
+predicted values for the test set.
 
 We first performed this analysis with the Kalla dataset. Since predictions of new molecules involve
-$n$ predictions per molecule, we removed three molecules from the dataset and rebuilt the random
-forest model. The hold out molecules, were specifically selected since they displayed significant
-activity cliffs with various members of the training set. We then evaluated the pairwise descriptor
-values for these three with the training set molecules and obtained predictions of the log(SALI)
-values. Figure \ref{fig:tset-pset-kalla}A summarizes the performance of the model on the training set
-as well as for the prediction set. From a numerical point of view, the performance of the model on
-he prediction set degrades somewhat (RMSE of 0.51 versus 0.41 for the training set). As noted above,
-the variance of the prediction is higher at the lower scale of log(SALI) values. However, at the
-higher end of the scale there a number of pairs that are well predicted. Yet, it is clear that a
-number of relatively significant activity cliffs have been underestimated.
+$n$ predictions per molecule, we removed ten molecules from the original dataset and rebuilt the
+random forest model. The hold out molecules included cases that displayed significant activity
+cliffs with various members of the training set, as well as other cases that did not really exhibit
+cliffs with any member of the training set. We then evaluated the pairwise descriptor values for
+these ten molecules with the training set molecules and obtained predictions of the log(SALI)
+values. Figure \ref{fig:tset-pset-kalla}A summarizes the performance of the model on the training
+set as well as for the prediction set. From a numerical point of view, the performance of the model
+on the prediction set degrades somewhat (RMSE of 0.41 versus 0.63 for the training set). As noted
+above, the variance of the prediction is higher at the lower scale of log(SALI) values. This is
+highlighted in Figure \ref{fig:sali-pred-variance}, where we divided the observed log SALI values
+into three groups, \emph{low}, \emph{medium} and \emph{high} (based on the quartiles of the values)
+and aggregated the prediction residuals for the observations belonging to each group. Clearly, for
+lower SALI values (i.e., less significant cliffs), the prediction errors are larger and more widely
+distributed. While \ref{fig:tset-pset-kalla}A indicates that a number of the more significant
+activity cliffs are quite well predicted, Figure \ref{fig:sali-pred-variance} indicates that as a
+group the observations with higher log SALI values tend to exhibit a small range of non-zero
+residual errors. However it is clear from the scatter plot that number of relatively significant
+activity cliffs have been underestimated.
 
@@
 
-Figure \ref{fig:kalla-comparison} displays the structures of the three hold out molecules (top row)
+Figure \ref{fig:kalla-comparison} displays the structures of three hold out molecules (top row)
 and a selection of structures, which which these hold out molecules were predicted to exhibit an
@@
 Figure \ref{fig:tset-pset-dai} summarizes the results for the same set of analyzes described above,
-applied to the Dai dataset. As before we removed three molecules (top row of
-Figure \ref{fig:dai-comparison}) and rebuilt the random forest on the paired cases with the remaining
-41 molecules. We then predicted the pairwise SALI values for the three hold-out molecules with the
-remaining 41 molecules (i.e., the prediction set). A plot of the predicted versus observed log(SALI)
-values for the training and prediction sets are displayed in Figure \ref{fig:tset-pset-dai}A. As
-before the model performs poorly on the lesser cliffs, but improves for the more significant cliffs
-(RMSE for the training set and prediction sets were 0.37 and 0.43 log units respectively). However,
-the 2 most significant cliffs are under-estimated. However, in comparison to the Kalla dataset, the
-distribution of absolute residuals aggregated by the hold-out molecule
-(Figure \ref{fig:tset-pset-dai}C) are similar, with a median absolute residual of less than
-0.3 log units.  For this dataset, the number of ``real'' cliffs is relatively low as shown in
-Figure \ref{fig:tset-pset-dai}B. Interestingly, the compound pairs that display moderate to
-significant cliffs are relatively well predicted - in fact, the maximum residuals are observed for
-the least significant cliffs.
+applied to the Dai dataset. As before we removed ten molecules (top row of Figure
+\ref{fig:dai-comparison}) and rebuilt the random forest on the paired cases with the remaining 34
+molecules. We then predicted the pairwise SALI values for the ten hold-out molecules with the
+remaining 34 molecules. A plot of the predicted versus observed log(SALI) values for the training
+and prediction sets are displayed in Figure \ref{fig:tset-pset-dai}A. As before the model performs
+poorly on the lesser cliffs, but improves for the more significant cliffs (RMSE for the training set
+and prediction sets were 0.37 and 0.43 log units respectively). Figure \ref{fig:sali-pred-variance}
+also highlights the fact that the model performs poorly at the lower end of the log(SALI)
+spectrum. However, the 2 most significant cliffs are under-estimated. However, in comparison to the
+Kalla dataset, the distribution of absolute residuals aggregated by the hold-out molecule (Figure
+\ref{fig:tset-pset-dai}C) are similar, with a median absolute residual of less than 0.3 log units.
+For this dataset, the number of ``real'' cliffs is relatively low as shown in Figure
+\ref{fig:tset-pset-dai}B. Interestingly, the compound pairs that display moderate to significant
+cliffs are relatively well predicted - in fact, the maximum residuals are observed for the least
+significant cliffs.
 
-We then considered the three hold-out molecules and some of the members of the training set with
+We then considered three of hold-out molecules and some of the members of the training set with
 which they are predicted to show activity cliffs. For this dataset, there are relatively few
@@
 subset of predictions where the $T_c \geq 0.8$ we note that the median absolute residual is 0.24 log
-units. However, the activity differences are not always significant ranging from 1.1-fold to
-13-fold. Two such examples are shown in Figure \ref{fig:dai-comparison}, where the only difference is
-in the position of the methyl substituent. Similar behavior is observed with the other hold-out
-molecules.
+units. However, the activity differences are not always very large ranging from 1.1-fold to
+13-fold. Two such examples are shown in Figure \ref{fig:dai-comparison}, where the only difference
+is in the position of the methyl substituent (\textbf{371259} and \textbf{371224}). Similar behavior
+is observed with the other hold-out molecules.
 
@@
 
+\ctable[cap={}, caption={A summary of the prediction residuals for log(SALI) values, grouped by
+  whether the actual log(SALI) value for that observation was \emph{low}, \emph{medium} or
+  \emph{high}. The grouping is based on  the quartiles of the observed values.}, figure, botcap,
+label={fig:sali-pred-variance}]    
+{cc}
+{}
+{\textbf{Kalla} & \textbf{Dai}  \NN
+\includegraphics[width=0.5\linewidth]{sali-pred-variance-kalla} & 
+\includegraphics[width=0.5\linewidth]{sali-pred-variance-dai}
+}
+
 \ctable[cap={}, caption={Three of the hold out molecules for the Kalla dataset

### 2012-07-27 rebuttal.tex
final set of edits for resubmission

@@
 
-XXX
+Yes, there do appear to be such systematic errors - at this stage it's not clear what is the
+cause. However we have noted that while it is clearly evident for the Cavalli dataset, it is not
+as obvious for the ChEMBL datasets and in some cases is the reverse of what we see in the Cavalli
+case. For example, for the Dai dataset, we observed that lower SALI values are overestimated
+(whereas the Cavalli predictions are underestimaetd at low SALI values).
+
+We hypothesize that the descriptors employed do not differentiate between observations with low and
+high SALI values very well, leading to overestimation of small values and underestimation of high
+SALI values. We have justified this by comparing descriptor distance distributions between the
+group of observations with low SALI values and those with high SALI values. The two distributions
+nearly completely overlap, supporting this hypothesis.
 
@@
 
-XXX
+We have updated this section to include several more hold out molecules. It should be noted that this section
+examines the model from a different viewpoint compared to Section 4. Specifically, Section 4.1
+develops models in which we remove molecules from the original dataset. Given the size of the
+original datasets, removing more than 10 or so, leads smaller pairwise datasets. Simultaneously, the
+predictions for this validation stratgey are not just for the 10 hold out molecules - it will be 10
+$\times m$ where $m$ is the number of the molecules in the original datasets. Thus the number of
+predictiosn range from 250 to 340 values. Summarizing more than a few of these hold out molecules
+becomes cumbersome.
 
@@
 
+In addition, Figures 8 and 10 do suggest that the approach does capture actual activity cliffs - it
+so happens that the datasets considered do not contain too many of them, to allow one to provide a
+large scale validation.
+
+As noted in Section 5, predictive performance can likely be increased by a larger, more diverse
+descriptor pools as well as alternative modeling methods. The goal of this paper is to suggest that
+the proposed pairwise methodology does let as address the identifcation of activity cliffs in a
+prospective manner.
+
 \textbf{(10) It also remains largely obscure how an activity landscape should be extended following
@@
 Section 4.1 in the original manuscript specifically addresses how one might use the proposed
-methodology
+methodology. This has been made more explicit in the updated manuscript.
 

### 2012-07-27 paper.tex
final set of edits for resubmission

@@
 for the three pairs are just 0.2, 0.30 and 0.29, though the activity differences were 4.93, 5.0 and
-5.11 log units respectively.
+5.11 log units respectively.  But more importantly, it is clear from all three panels that the there
+appears to be two different systematic prediction errors - smaller SALI values are overestimated
+whereas larger values are underestimated.  A possible explanation for this behavior is that the
+descriptors do not sufficiently differentiate observations with low and high SALI values. In other
+words, the descriptor profile for observations with low SALI values is sufficiently similar to those
+with high SALI values. To test this out, we split the Cavalli dataset into two groups - low and high
+- using the median SALI value as the cutoff. We then evaluated the pairwise descriptor distance
+distributions for each each of the groups, using the Euclidean distance between all descriptors in
+the pool. Figure \ref{fig:cavalli-desc-dist} indicates that the distance distributions are
+essentially identical. Though random forest models employ a subset of the descriptor pool for each
+tree, the same reasoning can be applied to these subsets. These trends appear in the other datasets
+(Figure \ref{fig:rfmodel-chembl}) but to varying degrees and not always consistently. For example,
+in the Costanzo dataset, we see that the higher SALI values are somewhat underestimated. The concept
+of descriptor similarity could also be applied to these cases. But a more fundamental explanation of
+this behavior is not available at this time.
 
@@
 datasets and so the following discussion omits the results obtained when using
-$f_{\textrm{geom}}$. In fact, a paired $t$-test between the predicted (from the training set) SALI
-values using models built with the different aggregation functions (i.e., $f_{\textrm{mean}}$ vs
-$f_{\textrm{absdiff}}$, etc.)  indicated that the differences between the sets of predictions was
+$f_{\textrm{geom}}$. In fact, a paired $t$-test between the predicted SALI values (for the training
+set) using models built with the different aggregation functions (i.e., $f_{\textrm{mean}}$ vs
+$f_{\textrm{diff}}$, etc.)  indicated that the differences between the sets of predictions was
 statistically insignificant (all $p$-values $> 0.1$). Fig.~\ref{fig:rfmodel-chembl} summarizes the
@@
 exhibit more variance for small SALI values (i.e., less significant cliffs) but appear to perform
-better at the higher end of the SALI values. However in all three cases, the most significant cliffs
-exhibit significant variance. Based on these observations, we arbitrarily chose $f_{\textrm{diff}}$
-as the aggregation function for subsequent analyses. Table \ref{tab:rfmodel-orig} and
-Figure \ref{fig:rfmodel-orig} summarizes the performance of the models built using the reported
-activity data (original, single molecule activity) for the Costanzo, Kalla and Dai datasets. Each of
-these models employed the entire dataset (rather than splitting into a training and testing set) and
+better at the higher end of the SALI values. However in all three cases, systematic prediction
+errors are apparent. Based on these observations, we arbitrarily chose $f_{\textrm{diff}}$ as the
+aggregation function for subsequent analyses. Table \ref{tab:rfmodel-orig} and Figure
+\ref{fig:rfmodel-orig} summarizes the performance of the models built using the reported activity
+data (original, single molecule activity) for the Costanzo, Kalla and Dai datasets. Each of these
+models employed the entire dataset (rather than splitting into a training and testing set) and
 exhibit relatively poor performance as indicated by the RMSE (ranging from 0.76 to 1.00 log units)
-and $R^2$ (0.18 to 0.34).. This can be partially ascribed to the small sizes of the dataset. It
+and $R^2$ (0.18 to 0.34). This can be partially ascribed to the small sizes of the dataset. It
 should be note that this methodology is not aimed at modeling the original, usually small, SAR
@@
 necessarily lead to very robust or reliable machine learning models. Instead, the methodology
-focuses on the \emph{pairwise} dataset -- which are usually large enough for a statistically
+focuses on \emph{pairwise} datasets -- which are usually large enough for a statistically
 significant model.
@@
 with all QSAR models, one is limited by the nature of the underlying data, both in terms of accuracy
-as well as applicability. We have selected a few ChEMBL assays for the purposes of
-highlighting this approach, we have found that for some assays, the model built on pairwise data
-performs worse than the model built on the original data. Given that most of the datasets are small
-in size (less than 40 molecules), the poor performance of the initial model is not always
-surprising. However, in cases where the pairwise model does not exhibit improved performance, we
-believe that predictive modeling of activity cliffs in the original dataset will not be fruitful.
+as well as applicability. We have selected a few ChEMBL assays for the purposes of highlighting this
+approach. Given that most of the datasets are small in size (30 to 50 molecules), models built on
+the actual activity data do not perform very well. In contrast the models built on the pairwise data
+usually exhibit improved performance - which could certainly be ascribed to the larger dataset
+involved. However, in cases where the pairwise model does
+not exhibit improved performance, we believe that predictive modeling of activity cliffs in the
+original dataset will not be fruitful.
 
@@
 
+\ctable[cap={}, caption={A comparison of the pairwise descriptor distribution for low and high SALI
+  values in the Cavalli dataset. This figure employed the descriptor values generated using the
+  $f_{diff}$ aggregation function and Euclidean distances were evaluataed using all the descriptors
+  in the pool. The \emph{low} group is defined as those observations with SALI
+  less than 2.03 and the remainder is assigned to the \emph{high} group.},
+label={fig:cavalli-desc-dist}, botcap, figure]
+{c}
+{}
+{
+\includegraphics[width=\linewidth]{cavalli-desc-dist}
+}
+
 \ctable[cap={}, caption={Predicted versus observed SALI values,
@@
 
-\ctable[cap={}, caption={Distribution of log(SALI) values for the
+\ctable[cap={}, caption={Distribution of observed log(SALI) values for the
   three ChEMBL datasets.}, figure, botcap,

### 2012-07-27 paper.tex
some minor typos

@@
 words, the descriptor profile for observations with low SALI values is sufficiently similar to those
-with high SALI values. To test this out, we split the Cavalli dataset into two groups - low and high
+with high SALI values. To test this, we split the Cavalli dataset into two groups - low and high
 - using the median SALI value as the cutoff. We then evaluated the pairwise descriptor distance
@@
   values in the Cavalli dataset. This figure employed the descriptor values generated using the
-  $f_{diff}$ aggregation function and Euclidean distances were evaluataed using all the descriptors
+  $f_{\textrm{diff}}$ aggregation function and Euclidean distances were evaluated using all the descriptors
   in the pool. The \emph{low} group is defined as those observations with SALI
-  less than 2.03 and the remainder is assigned to the \emph{high} group.},
+  less than 2.03 and the remainder are assigned to the \emph{high} group.},
 label={fig:cavalli-desc-dist}, botcap, figure]

### 2012-07-27 paper.tex
final edits

@@
   descriptors. The models exhibited reasonable RMSE's though, surprisingly, performance on the more
-  significant cliffs tended to be better than on the lesser ones. Our results indicate that the
-  predictive models are able to prioritize molecules in terms of their ability to activity cliffs
-  and can be used as a way to extend an observed activity landscape to identify molecules that could
-  lead to significant improvements in activity. Conversely, the method can be used to deprioritize
-  molecules that are predicted to show an activity cliff with a potent molecule already in the
-  dataset.
+  significant cliffs tended to be better than on the lesser ones. While the models do not exhibit
+  very high levels of accuracy, our results indicate that they are able to prioritize molecules in
+  terms of their ability to activity cliffs, thus serving as a tool to prospectively identify
+  activity cliffs..
 \end{abstract}
@@
 cliffs, which an then be examined in detail. Apart from identifying individual activity cliffs, a
-variety of other SAR constructs, such as ``activity ridges'' \cite{Vogt:2011bs} and multitarget
+variety of other SAR constructs, such as ``activity ridges'' \cite{Vogt:2011bs} and multi-target
 landscapes\cite{Dimova:2011fk} can also be identified and characterized.
@@
 dramatically in terms of the final model performance as root mean square error (RMSE) on the test
-set ranges from 1.01 to 1.07 log units. However, $f_{\textrm{mean}}$ appears to lead to the best
-performance (RMSE = 1.01 and $R^2$ = 0.58). In contrast, the RMSE and $R^2$ values for
-Y-scrambled\cite{Rucker:2007aa} models using the different aggregation functions ranged from 1.58 to
-1.61 log units and 0.003 to 0.01, respectively.
+set ranges from 1.01 to 1.07 log units, which is lower compared to the standard deviation of the
+dependent variable. However, $f_{\textrm{mean}}$ appears to lead to the best performance (RMSE =
+1.01 and $R^2$ = 0.58). In contrast, the RMSE and $R^2$ values for Y-scrambled\cite{Rucker:2007aa}
+models using the different aggregation functions ranged from 1.58 to 1.61 log units and 0.003 to
+0.01, respectively.
 
@@
 models built on the ChEMBL datasets using $f_{\textrm{diff}}$ and $f_{\textrm{mean}}$ as the
-aggregation functions. Table \ref{tab:rfmodel-results} reports the model statistics.  While the
-differences in model performance do not vary significantly with the aggregation function used, we
-see that for all the ChEMBL datasets, $f_{\textrm{diff}}$ leads to a slightly better model in terms
-of RMSE and $R^2$ than when using $f_{\textrm{mean}}$. As with the Cavalli dataset, the SALI models
-exhibit more variance for small SALI values (i.e., less significant cliffs) but appear to perform
-better at the higher end of the SALI values. However in all three cases, systematic prediction
-errors are apparent. Based on these observations, we arbitrarily chose $f_{\textrm{diff}}$ as the
-aggregation function for subsequent analyses. Table \ref{tab:rfmodel-orig} and Figure
-\ref{fig:rfmodel-orig} summarizes the performance of the models built using the reported activity
-data (original, single molecule activity) for the Costanzo, Kalla and Dai datasets. Each of these
-models employed the entire dataset (rather than splitting into a training and testing set) and
-exhibit relatively poor performance as indicated by the RMSE (ranging from 0.76 to 1.00 log units)
-and $R^2$ (0.18 to 0.34). This can be partially ascribed to the small sizes of the dataset. It
-should be note that this methodology is not aimed at modeling the original, usually small, SAR
-dataset. It is well known (and evident from the current analyses) that small datasets do not
-necessarily lead to very robust or reliable machine learning models. Instead, the methodology
-focuses on \emph{pairwise} datasets -- which are usually large enough for a statistically
-significant model.
+aggregation functions. Table \ref{tab:rfmodel-results} reports the model statistics Of the three,
+the Costanzo stands out as being rather poor given the training and test set RMSE values being quite
+close to (or greater than) the standard deviation of the dependent variable.  While the differences
+in model performance do not vary significantly with the aggregation function used, we see that for
+all the ChEMBL datasets, $f_{\textrm{diff}}$ leads to a slightly better model in terms of RMSE and
+$R^2$ than when using $f_{\textrm{mean}}$. As with the Cavalli dataset, the SALI models exhibit more
+variance for small SALI values (i.e., less significant cliffs) but appear to perform better at the
+higher end of the SALI values. However in all three cases, systematic prediction errors are
+apparent. Based on these observations, we arbitrarily chose $f_{\textrm{diff}}$ as the aggregation
+function for subsequent analyses. Table \ref{tab:rfmodel-orig} and Figure \ref{fig:rfmodel-orig}
+summarizes the performance of the models built using the reported activity data (original, single
+molecule activity) for the Costanzo, Kalla and Dai datasets. Each of these models employed the
+entire dataset (rather than splitting into a training and testing set) and exhibit relatively poor
+performance as indicated by the RMSE (ranging from 0.76 to 1.00 log units) and $R^2$ (0.18 to
+0.34). The model for the Kalla dataset is especially poor, given the RMSE is greater than the
+standard deviation of the dependent variable. The poor performance of these models can be partially
+ascribed to the small sizes of the dataset. It should be note that this methodology is not aimed at
+modeling the original, usually small, SAR dataset. It is well known (and evident from the current
+analyses) that small datasets do not necessarily lead to very robust or reliable machine learning
+models. Instead, the methodology focuses on \emph{pairwise} datasets -- which are usually large
+enough for a statistically significant model. Based on the poor performance observed for the
+Costanzo dataset, we excluded it from our subsequent discussions.
 
@@
 \ctable[cap={}, caption={Performance metrics for the random forest
-  models built for the datasets used in this study.},
+  models built for the datasets used in this study. SD is the standard
+deviation of the dependent variable.},
 label={tab:rfmodel-results}]
-{lrcrrrr}
+{lrrcrrrr}
 {
@@
 {\FL
-        &         &                      & \multicolumn{2}{c}{Training} & \multicolumn{2}{c}{Test}
+        &    &     &                      & \multicolumn{2}{c}{Training} & \multicolumn{2}{c}{Test}
 \NN
-\cmidrule{4-5}  \cmidrule{6-7}
-Dataset & Y-range\tmark & Aggregation Function & RMSE & $R^2$ & RMSE & $R^2$   
+\cmidrule{5-6}  \cmidrule{7-8}
+Dataset & Y-range\tmark & SD\tmark & Aggregation Function & RMSE & $R^2$ & RMSE & $R^2$   
 \ML
-Cavalli & 7.21 & $f_{\textrm{diff}}$ & 1.16& 0.54 & 1.07& 0.56\NN
-        & & $f_{\textrm{mean}}$      & 1.03& 0.64 & 1.01& 0.58\NN
-        & & $f_{\textrm{geom}}$      & 1.07& 0.61 & 1.01& 0.59\NN \NN
+Cavalli & 7.21 & 1.57 & $f_{\textrm{diff}}$ & 1.16& 0.54 & 1.07& 0.56\NN
+        & & & $f_{\textrm{mean}}$      & 1.03& 0.64 & 1.01& 0.58\NN
+        & & & $f_{\textrm{geom}}$      & 1.07& 0.61 & 1.01& 0.59\NN \NN
 %%
-Costanzo & 288.03& $f_{\textrm{diff}}$ & 11.99& 0.24 & 17.97& 0.10 \NN
-       & & $f_{\textrm{mean}}$         & 12.88& 0.12 & 16.96& 0.20\NN \NN
+Costanzo & 288.02& 14.84 & $f_{\textrm{diff}}$ & 11.99& 0.24 & 17.97& 0.10 \NN
+       & & &  $f_{\textrm{mean}}$         & 12.88& 0.12 & 16.96& 0.20\NN \NN
 %%
-Kalla & 88.25 & $f_{\textrm{diff}}$ & 8.12& 0.59 & 6.49& 0.57\NN
-      &  & $f_{\textrm{mean}}$      & 8.79& 0.52 & 8.00& 0.47\NN \NN
+Kalla & 88.25 & 12.16 & $f_{\textrm{diff}}$ & 8.12& 0.59 & 6.49& 0.57\NN
+      &  & &  $f_{\textrm{mean}}$      & 8.79& 0.52 & 8.00& 0.47\NN \NN
 %%
-Dai & 32.71 & $f_{\textrm{diff}}$ & 2.49& 0.48 & 3.21& 0.22\NN
-     &   & $f_{\textrm{mean}}$    & 2.64& 0.41 & 2.86& 0.37
+Dai & 32.71 & 3.48 & $f_{\textrm{diff}}$ & 2.49& 0.48 & 3.21& 0.22\NN
+     &   & & $f_{\textrm{mean}}$    & 2.64& 0.41 & 2.86& 0.37
 \LL
@@
   models built on the single molecule activity data, for each ChEMBL dataset. In each case the model
-was built using the logarithm of the reported activities.},
+was built using the $\log_{10}$ of the reported activities.},
 label={tab:rfmodel-orig}]
-{lrrr}
+{lrrrr}
 {
@@
 {\FL
-Dataset & Y-range & RMSE & $R^2$
+Dataset & Y-range & SD & RMSE & $R^2$
 \ML
-Costanzo & 4.67 & 0.80 & 0.18 \NN
-Kall          & 4.44 & 1.00 & 0.32 \NN
-Dai          &  3.60 & 0.76 & 0.34
+Costanzo & 4.67 & 1.22 & 0.80 & 0.18 \NN
+Kalla          & 4.44 & 0.78 & 1.00 & 0.32 \NN
+Dai          &  3.60 & 0.94 & 0.76 & 0.34
 \LL

### 2012-07-27 letter-resubmit.tex
added new submission letter. fixed some minor typos

--- 2012-07-10 letter.tex
@@
     \begin{newlfm}
-      I would like to submit an article titled ``\emph{Exploring
+      I would like to resubmit the article titled ``\emph{Exploring
         Uncharted Territories - Predicting Activty Cliffs in
-        Structure-Activity Landscapes}'' for publication in the
+        Structure-Activity Landscapes}'' (ci-2012-00047k) for consideration in the
       Journal of Chemical Information and Modeling.
 
-      The article builds on previous work that defined a numerical
-      quantification of structure activity landscapes, termed the
-      Structure Activity Landscape Index. This (and other methods such
-      as SARI) have been used to characterize SAR data using the
-      landscape paradigm, allowing users to identify activity cliffs
-      in their datasets. However, these approaches have primarily been
-      retrospective. In this paper I highlight how SALI can be used in
-      a prospective manner, such that one can predict SALI values for
-      pairs of molecules, one of which is a new, untested molecule. In
-      other words, the method discussed in this paper allows one to
-      prospectively prioritize new molecules as likely to exhibit
-      activity cliffs with one or more members of the original
-      dataset. The predictive approach is based on random forest
-      models and considers the data set in a pairwise fashion - thus
-      making it amenable for small SAR datasets. The models exhibit
-      good predictive performance, suggesting the utility of this
-      approach.
+A separate document has been included that addresses each of the
+reviewers comments.
+
 %  \vskip 0.5em

### 2012-07-27 submit-2/paper.tex
added resubmission folder

--- 2012-07-27 paper.tex
@@
   terms of their ability to activity cliffs, thus serving as a tool to prospectively identify
-  activity cliffs..
+  activity cliffs.
 \end{abstract}
@@
 
-\bibliography{paper}
+%%\bibliography{paper}
+
+\providecommand{\refin}[1]{\\ \textbf{Referenced in:} #1}
+\begin{thebibliography}{10}
+
+\bibitem{Johnson:1990ys}
+Johnson,~M.;\ \ Maggiora,~G. \textit{Concepts and Applications of Molecular
+  Similarity;} John Wiley \& Sons: New York, 1990.
+
+\bibitem{Maggiora:2006aa}
+Maggiora,~G.~M.  On Outliers and Activity Cliffs--Why {QSAR} Often
+  Disappoints..  \textit{J.~Chem.~Inf.~Model.} \textbf{2006,} \textsl{46,}
+  1535--1535.
+
+\bibitem{Leach:2006aa}
+Leach,~A.;\ \ Jones,~H.;\ \ Cosgrove,~D.;\ \ Kenny,~P.;\ \ Ruston,~L.;\ \
+  MacFaul,~P.;\ \ Wood,~J.;\ \ Colclough,~N.;\ \ Law,~B.  Matched Molecular
+  Pairs as a Guide in the Optimization of Pharmaceutical Properties; a Study of
+  Aqueous Solubility, Plasma Protein Binding and Oral Exposure.
+  \textit{J.~Med.~Chem.} \textbf{2006,} \textsl{49,} 6672--6682.
+
+\bibitem{Maggiora:2011kx}
+Shanmugasundaram,~V.;\ \ Maggiora,~G.  Characterizing Property and Activity
+  Landscapes Using an Information-Theoretic Approach.   In  \textit{CINF-032.
+  222nd ACS National Meeting, Chicago, IL, United States}; American Chemical
+  Society: Washington, D.C., 2001.
+
+\bibitem{Guha:2008aa}
+Guha,~R.;\ \ Van~Drie,~J.  The Structure-Activity Landscape Index:
+  {I}dentifying and Quantifying Activity-Cliffs.  \textit{J.~Chem.~Inf.~Model.}
+  \textbf{2008,} \textsl{48,} 646--658.
+
+\bibitem{Peltason:2007aa}
+Peltason,~L.;\ \ Bajorath,~J.  SAR Index: Quantifying the Nature of
+  Structure-Activity Relationships.  \textit{J.~Med.~Chem.} \textbf{2007,}
+  \textsl{50,} 5571--5578.
+
+\bibitem{Wawer:2008aa}
+Wawer,~M.;\ \ Peltason,~L.;\ \ Weskamp,~N.;\ \ Teckentrup,~A.;\ \ Bajorath,~J.
+  Structure-Activity Relationship Anatomy by Network-like Similarity Graphs and
+  Local Structure-Activity Relationship Indices.  \textit{J.~Med.~Chem.}
+  \textbf{2008,} \textsl{51,} 6075--6084.
+
+\bibitem{Vogt:2011bs}
+Vogt,~M.;\ \ Huang,~Y.;\ \ Bajorath,~J.  From Activity Cliffs to Activity
+  Ridges: {I}nformative Data Structures for SAR Analysis.
+  \textit{J.~Chem.~Inf.~Model.} \textbf{2011,} \textsl{51,} 1848--1856.
+
+\bibitem{Dimova:2011fk}
+Dimova,~D.;\ \ Wawer,~M.;\ \ Wassermann,~A.~M.;\ \ Bajorath,~J.  Design of
+  Multitarget Activity Landscapes That Capture Hierarchical Activity Cliff
+  Distributions.  \textit{J.~Chem.~Inf.~Model.} \textbf{2011,} \textsl{51,}
+  258--266.
+
+\bibitem{Wawer:2010fv}
+Wawer,~M.;\ \ Bajorath,~J.  Similarity-potency trees: {A} method to search for
+  {SAR} information in compound data sets and derive {SAR} rules.
+  \textit{J.~Chem.~Inf.~Model.} \textbf{2010,} \textsl{50,} 1395--1409.
+
+\bibitem{Hu:2010kl}
+Hu,~Y.;\ \ Bajorath,~J.  Molecular Scaffolds with High Propensity to Form
+  Multi-Target Activity Cliffs.  \textit{J.~Chem.~Inf.~Model.} \textbf{2010,}
+  \textsl{50,} 500--510.
+
+\bibitem{Wassermann:2010oq}
+Wassermann,~A.~M.;\ \ Bajorath,~J.  Chemical Substitutions that Introduce
+  Activity Cliffs Across Different Compound Classes and Biological Targets.
+  \textit{J.~Chem.~Inf.~Model.} \textbf{2010,} \textsl{50,} 1248-1256.
+
+\bibitem{Peltason:2009vn}
+Peltason,~L.;\ \ Weskamp,~N.;\ \ Teckentrup,~A.;\ \ Bajorath,~J.  Exploration
+  of Structure-Activity Relationship Determinants in Analogue Series.
+  \textit{J.~Med.~Chem.} \textbf{2009,} \textsl{52,} 3212--3224.
+
+\bibitem{Scheiber:2009fk}
+Scheiber,~J.;\ \ Jenkins,~J.~L.;\ \ Bender,~A.;\ \ Milik,~M.;\ \ Mikhalov,~D.;\
+  \ Sukuru,~S.;\ \ Cornett,~B.;\ \ Whitebread,~S.;\ \ Laszlo,~U.;\ \
+  Davies,~J.;\ \ Glick,~M.  {SPREAD} - exploiting chemical features that cause
+  differential activity behavior.  \textit{Stat.~Anal.~Data Mining}
+  \textbf{2009,} \textsl{2,} 115-122.
+
+\bibitem{Cavalli:2002aa}
+Cavalli,~A.;\ \ Poluzzi,~E.;\ \ De~Ponti,~F.;\ \ Recanatini,~M.  Toward a
+  Pharmacophore for Drugs Inducing the Long {QT} Syndrome: Insights from a
+  {CoMFA} Study of {HERG K+} Channel Blockers.  \textit{J.~Med.~Chem.}
+  \textbf{2002,} \textsl{45,} 3844--3853.
+
+\bibitem{Costanzo:2005uq}
+Costanzo,~M.~J.;\ \ Almond,~Jr,~H.~R.;\ \ Hecker,~L.~R.;\ \ Schott,~M.~R.;\ \
+  Yabut,~S.~C.;\ \ Zhang,~H.-C.;\ \ Andrade-Gordon,~P.;\ \ Corcoran,~T.~W.;\ \
+  Giardino,~E.~C.;\ \ Kauffman,~J.~A.;\ \ Lewis,~J.~M.;\ \ de~Garavilla,~L.;\ \
+  Haertlein,~B.~J.;\ \ Maryanoff,~B.~E.  In-depth study of tripeptide-based
+  alpha-ketoheterocycles as inhibitors of thrombin. {E}ffective utilization of
+  the {S1'} subsite and its implications to structure-based drug design.
+  \textit{J.~Med.~Chem.} \textbf{2005,} \textsl{48,} 1984--2008.
+
+\bibitem{Kalla:2006kx}
+Kalla,~R.~V.;\ \ Elzein,~E.;\ \ Perry,~T.;\ \ Li,~X.;\ \ Palle,~V.;\ \
+  Varkhedkar,~V.;\ \ Gimbel,~A.;\ \ Maa,~T.;\ \ Zeng,~D.;\ \ Zablocki,~J.
+  Novel 1,3-disubstituted 8-(1-benzyl-1H-pyrazol-4-yl) xanthines: high affinity
+  and selective A2B adenosine receptor antagonists.  \textit{J.~Med.~Chem.}
+  \textbf{2006,} \textsl{49,} 3682--3692.
+
+\bibitem{Dai:2007kl}
+Dai,~Y. \textit{et al.}\   Discovery of
+  N-(4-(3-amino-1H-indazol-4-yl)phenyl)-N'-(2-fluoro-5-methylphenyl)urea
+  (ABT-869), a 3-aminoindazole-based orally active multitargeted receptor
+  tyrosine kinase inhibitor.  \textit{J.~Med.~Chem.} \textbf{2007,}
+  \textsl{50,} 1584--1597.
+
+\bibitem{Steinbeck:2006aa}
+Steinbeck,~C.;\ \ Hoppe,~C.;\ \ Kuhn,~S.;\ \ Floris,~M.;\ \ Guha,~R.;\ \
+  Willighagen,~E.  Recent Developments of the Chemistry Development Kit ({CDK})
+  - An Open-Source {Java} Library for Chemo- and Bioinformatics.
+  \textit{Curr.~Pharm.~Des.} \textbf{2006,} \textsl{12,} 2110--2120.
+
+\bibitem{Steinbeck:2003bh}
+Steinbeck,~C.;\ \ Han,~Y.~Q.;\ \ Kuhn,~S.;\ \ Horlacher,~O.;\ \ Luttmann,~E.;\
+  \ Willighagen,~E.  The Chemistry Development Kit ({CDK}): An Open-Source
+  {J}ava Library for Chemo- and Bioinformatics.
+  \textit{J.~Chem.~Inf.~Comput.~Sci.} \textbf{2003,} \textsl{43,} 493--500.
+
+\bibitem{Breiman:1984aa}
+Breiman,~L.;\ \ Friedman,~J.;\ \ Olshen,~R.;\ \ Stone,~C.
+  \textit{Classification and Regression Trees;} Chapman \& Hall/CRC: Boca
+  Raton, FL, 1984.
+
+\bibitem{Breiman:2001nx}
+Breiman,~L.  Statistical Modeling: Two Cultures.  \textit{Stat.~Sci.}
+  \textbf{2001,} \textsl{16,} 199--231.
+
+\bibitem{r}
+{R Development Core Team}, ``{R}: {A} Language and Environment for Statistical
+  Computing'',  R Foundation for Statistical Computing, Vienna, Austria, 2008
+  {ISBN} 3-900051-07-0.
+
+\bibitem{Rucker:2007aa}
+Rucker,~C.;\ \ Rucker,~G.;\ \ Meringer,~M.  y-Randomization and Its Variants in
+  {QSPR/QSAR}.  \textit{J.~Chem.~Inf.~Model.} \textbf{2007,} \textsl{47,}
+  2345--2357.
+
+\end{thebibliography}
 
